### 2021-08-26 README.Rmd
initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# cetaceanbcg

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2021-08-26 analysis/paper/paper.Rmd
initial commit

---
title: "Something about cetacean heart rates and ballistocardiograms"
author:
  - Max Czapanskiy:
      email: maxczap@stanford.edu
      institute: [SU]
      correspondence: true
  - More Authors:
      email: more@authors.edu
      institute: [UofA]
      correspondence: false
institute:
  - SU: Stanford University
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
library(cetaceanbcg)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r process-acc}
acc_400hz <- readRDS(here::here("analysis/data/derived_data/acc_400hz.rds"))
fs <- 400
fny <- fs / 2
acc_filter <- signal::butter(4, c(1.0, 25.0) / fny, type = "pass") 
apply_filter <- function(x) signal::filtfilt(acc_filter, x)
acc_400hz <- acc_400hz %>% 
  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))

# This window (in seconds) will be used for smoothing multiple signals and in
# heartbeat detection
window <- 0.25
prh_10hz <- readRDS(here::here("analysis/data/derived_data/prh_10hz.rds")) %>% 
  mutate(jerk_smooth = tma(prh_10hz$jerk, window * 10))
```

```{r align-ecg}
ecg_beats <- readRDS(here::here("analysis/data/derived_data/ecg_beats.rds"))

acc_400hz$ecg_beat <- FALSE
beat_idx <- approx(acc_400hz$dt, 
                   seq_along(acc_400hz$dt), 
                   xout = ecg_beats$dt,
                   method = "constant")$y %>% 
  na.omit()
acc_400hz$ecg_beat[beat_idx] = TRUE

prh_10hz$ecg_beat <- FALSE
beat_idx <- approx(prh_10hz$dt, 
                   seq_along(prh_10hz$dt), 
                   xout = ecg_beats$dt,
                   method = "constant")$y %>% 
  na.omit()
prh_10hz$ecg_beat[beat_idx] = TRUE
```

```{r bcg-beats}
acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
acc_400hz$bcg_beat <- find_beats(acc_400hz$surge_filt, window, fs)
acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
  filter(acc_400hz, ecg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)
acc_400hz$bcg_bpm[acc_400hz$bcg_beat] <- with(
  filter(acc_400hz, bcg_beat), 
  1 / as.numeric(lead(dt) - dt, unit = "mins")
)

prh_10hz$bcg_beat <- FALSE
bcg_idx <- approx(prh_10hz$dt, 
                  seq_along(prh_10hz$bcg_beat), 
                  acc_400hz$dt[acc_400hz$bcg_beat],
                  "constant")$y
prh_10hz$bcg_beat[bcg_idx] <- TRUE
```

```{r bcg-results, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration). Solid line: surge bandpass filtered to [1.0 25.0] Hz. Dashed line: local range of surge (LRS) in 0.25 s sliding window. Blue dots indicate peak of the BCG I-wave. B: Heart rates estimated from BCG (blue) match those from ECG (red). Temporal offset is due to both physiology and sensor clock differences. C: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge. The sway signal is likely due to movement of the chest cavity (see video S1). D: Some, but not all, heart beats are visible in the norm of the jerk vector."}
ca_tz <- "Etc/GMT+7"
bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
bcg_sample <- filter(acc_400hz, between(dt, bcg_begin, bcg_end))
bcg_sample_10hz <- filter(prh_10hz, between(dt, bcg_begin, bcg_end))

t <- theme_minimal() +
  theme(axis.title.x = element_blank(),
        panel.grid.minor = element_blank())

# Fig 1a: surge and 
bcg_surge_lsr <- bcg_sample %>% 
  rename(Surge = surge_filt, LSR = surge_range) %>% 
  pivot_longer(c(Surge, LSR), 
               names_to = "var",
               values_to = "val")

p1 <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
  labs(y = bquote(Surge~(m~s^{-2}))) +
  t +
  theme(legend.position = "none")

bcg_sway_heave <- bcg_sample %>% 
  rename(Sway = sway_filt, Heave = heave_filt) %>% 
  pivot_longer(c(Sway, Heave), 
               names_to = "axis",
               values_to = "acceleration")

p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
  geom_line() +
  geom_point(color = "blue") +
  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat)) +
  geom_point(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), color = "red") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = "Heart rate (bpm)") +
  expand_limits(x = range(bcg_sample$dt), y = c(30, 70)) +
  t

p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  facet_grid(axis ~ .) +
  labs(y = bquote(Acc~(m~s^{-2}))) +
  t

p4 <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
  geom_line(size = 0.2) +
  geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  labs(y = bquote(Jerk~(m~s^{-3}))) +
  expand_limits(y = 0) +
  t

cowplot::plot_grid(p1, p2, p3, p4, 
                   align = "v", 
                   axis = "lr",
                   nrow = 4, 
                   rel_heights = c(1.5, 0.75, 1, 1),
                   labels = "AUTO")
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2021-08-26 R/find_beats.R

@@ -1,0 +1,58 @@
+#' Find beats in BCG
+#'
+#' @param surge Filtered surge signal
+#' @param window_s Width of search window in seconds (0.25 s by default)
+#' @param fs_hz Sampling frequency of surge in Hz (400 Hz by default)
+#'
+#' @return A logical vector the same length as surge indicating a heart beat
+#' @export
+find_beats <- function(surge, window_s = 0.25, fs_hz = 400) {
+  window <- as.integer(window_s * fs)
+
+  # Local range of surge
+  surge_range <- local_range(surge, window)
+
+  # Find local peaks in smoothed local range
+  local_range_max <- RcppRoll::roll_max(surge_range, window, fill = NA)
+  range_peaks <- surge_range == local_range_max
+
+  # Align range peaks with surge peaks
+  beats <- logical(length(range_peaks))
+  i <- 1
+  for (peak in which(range_peaks)) {
+    i1 <- max(peak - window / 2, 1)
+    i2 <- min(peak + window / 2, length(surge))
+    b <- which.max(surge[i1:i2]) + i1 - 1
+    beats[b] <- TRUE
+    i <- i + 1
+  }
+
+  beats
+}
+
+#' Local range of signal
+#'
+#' @param x Original signal
+#' @param k Window width
+#'
+#' @return Smoothed local range of x
+#' @export
+local_range <- function(x, k) {
+  local_range_max <- RcppRoll::roll_max(x, k, fill = NA)
+  local_range_min <- RcppRoll::roll_min(x, k, fill = NA)
+  result <- local_range_max - local_range_min
+  tma(result, k)
+}
+
+#' Triangular moving average
+#'
+#' @param x Original signal
+#' @param k Window width
+#'
+#' @return Smoothed x
+#' @export
+tma <- function(x, k) {
+  x %>%
+    RcppRoll::roll_mean(k, fill = NA) %>%
+    RcppRoll::roll_mean(k, fill = NA)
+}

### 2021-08-26 R/zzz.R

@@ -1,0 +1,2 @@
+#' @import tidyverse
+NULL

### 2021-08-26 analysis/data/import_data.R

@@ -1,0 +1,65 @@
+library(tidyverse)
+
+# 400 Hz acceleration data
+acc_400hz_mat <- R.matlab::readMat(
+  "analysis/data/raw_data/20210816-134844-CATS 97.mat"
+)
+tz <- "Etc/GMT+7"
+acc_400hz <- tibble(
+  dn = as.vector(acc_400hz_mat$Atime),
+  surge = as.vector(acc_400hz_mat$Adata[, 1]),
+  sway = as.vector(acc_400hz_mat$Adata[, 2]),
+  heave = as.vector(acc_400hz_mat$Adata[, 3])
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, surge, sway, heave)
+
+saveRDS(acc_400hz, "analysis/data/derived_data/acc_400hz.rds")
+
+# 10 Hz PRH data
+njerk <- function(A, fs) {
+  jerk <- (A[-1, ] - A[-nrow(A), ]) / fs
+  jerk <- rbind(rep(NA, 3), jerk)
+  apply(jerk, 1, function(row) sqrt(sum(row^2)))
+}
+prh_10hz_mat <- R.matlab::readMat(
+  "analysis/data/raw_data/corky210816-97 10Hzprh.mat"
+)
+prh_10hz <- tibble(
+  dn = as.vector(prh_10hz_mat$DN),
+  aw = as.matrix(prh_10hz_mat$Aw),
+  pitch = as.vector(prh_10hz_mat$pitch),
+  roll = as.vector(prh_10hz_mat$roll),
+  head = as.vector(prh_10hz_mat$head)
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz),
+         jerk = njerk(aw * 9.8, 10)) %>%
+  select(dt, pitch, roll, head, jerk)
+
+saveRDS(prh_10hz, "analysis/data/derived_data/prh_10hz.rds")
+
+# ECG data
+set_date <- function(datetime, new_date) {
+  result <- rep(new_date, length(datetime))
+  lubridate::hour(result) <- lubridate::hour(datetime)
+  lubridate::minute(result) <- lubridate::minute(datetime)
[…]

### 2021-08-27 analysis/paper/paper.Rmd
In progress: finding beats in manually identified motionless sections
Former-commit-id: ae905f37d61d7725792abc0f663843fd9b1eb82f

@@
 
-```{r setup, echo = FALSE}
+```{r setup, echo=FALSE, message=FALSE}
 knitr::opts_chunk$set(
@@
 library(cetaceanbcg)
+library(tidyverse)
 ```
@@
 prh_10hz <- readRDS(here::here("analysis/data/derived_data/prh_10hz.rds")) %>% 
-  mutate(jerk_smooth = tma(prh_10hz$jerk, window * 10))
+  mutate(jerk_smooth = tma(jerk, window * 10))
 ```
@@
 acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
-acc_400hz$bcg_beat <- find_beats(acc_400hz$surge_filt, window, fs)
+acc_400hz$bcg_beat <- find_beats_lsr(acc_400hz$surge_filt, window, fs)
 acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
@@
 
-```{r bcg-results, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
+```{r corky-bcg, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
 Solid line: surge bandpass filtered to [1.0 25.0] Hz.
-Dashed line: local range of surge (LRS) in 0.25 s sliding window.
+Dashed line: local range of surge in 0.25 s sliding window (smoothed).
 Blue dots indicate peak of the BCG I-wave.
@@
 The sway signal is likely due to movement of the chest cavity (see video S1).
-D: Some, but not all, heart beats are visible in the norm of the jerk vector."}
+D: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
 ca_tz <- "Etc/GMT+7"
@@
 
-# Fig 1a: surge and 
+# A: Surge + LSR
 bcg_surge_lsr <- bcg_sample %>% 
@@
 
-bcg_sway_heave <- bcg_sample %>% 
-  rename(Sway = sway_filt, Heave = heave_filt) %>% 
-  pivot_longer(c(Sway, Heave), 
-               names_to = "axis",
-               values_to = "acceleration")
+# B: BPM
 
@@
 
+# C: Sway, heave
+
+bcg_sway_heave <- bcg_sample %>% 
+  rename(Sway = sway_filt, Heave = heave_filt) %>% 
+  pivot_longer(c(Sway, Heave), 
+               names_to = "axis",
+               values_to = "acceleration")
+
 p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
@@
   t
+
+# D: Jerk
 
@@
 
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
+```{r bw-data}
+fs <- 400
+fny <- fs / 2
+acc_filter <- signal::butter(4, c(1.5, 7.5) / fny, type = "pass") 
+apply_filter <- function(x) signal::filtfilt(acc_filter, x)
+window <- 1.8 #s
+bw_400hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_400hz.rds")) %>% 
+  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))
+
+bw_10hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_10hz.rds")) %>% 
+  mutate(jerk_smooth = tma(jerk, window * 10))
+
+bw_elg <- readRDS(here::here("analysis/data/derived_data/bw180905_53_elg.rds"))
 ```
 
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
+```{r bw-beats}
+bw_400hz <- mutate(bw_400hz, surge_range = NA, bcg_beat = NA, bcg_bpm = NA)
 
-```{r demo-inline-code}
-x <- round(pi, 2)
+for (row in seq(nrow(bw_eligible))) {
+  i <- approx(bw_400hz$dt, 
+              seq_along(bw_400hz$dt), 
+              bw_elg[row, c("start", "stop")])$y
+  i <- floor(i[1]):floor(i[2])
+  region <- bw_400hz[i, ]
+  surge <- region$surge_filt
+  surge_range <- local_range(surge, window *fs)
+  bcg_beat <- find_beats_lsr(surge, window, fs)
+  bcg_bpm <- rep(NA, length(i))
+  if (sum(bcg_beat) > 1) {
+    bcg_dt <- region$dt[bcg_beat]
+    bcg_bpm[bcg_beat] <- 1 / as.numeric(lead(bcg_dt) - bcg_dt, unit = "mins")
+  }
+  bw_400hz$surge_range[i] <- surge_range
+  bw_400hz$bcg_beat[i] <- bcg_beat
+  bw_400hz$bcg_bpm[i] <- bcg_bpm
+}
+
+bw_10hz$bcg_beat <- FALSE
+bcg_idx <- approx(bw_10hz$dt, 
+                  seq_along(bw_10hz$bcg_beat), 
+                  bw_400hz$dt[bw_400hz$bcg_beat],
+                  "constant")$y
+bw_10hz$bcg_beat[bcg_idx] <- TRUE
 ```
+
+```{r bw-bcg}
+ca_tz <- "Etc/GMT+7"
+bcg_begin <- lubridate::ymd_hms("2018-09-05 21:29:05", tz = ca_tz)
+bcg_end <- lubridate::ymd_hms("2018-09-05 21:30:30", tz = ca_tz)
+bcg_sample <- filter(bw_400hz, between(dt, bcg_begin, bcg_end))
+bcg_sample_10hz <- filter(bw_10hz, between(dt, bcg_begin, bcg_end))
+
+# A: Surge + LSR
+bcg_surge_lsr <- bcg_sample %>% 
+  rename(Surge = surge_filt, LSR = surge_range) %>% 
+  pivot_longer(c(Surge, LSR), 
+               names_to = "var",
+               values_to = "val")
+
+p1 <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
+  labs(y = bquote(Surge~(m~s^{-2}))) +
+  t +
+  theme(legend.position = "none")
+
+# B: BPM
+
+p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
+  geom_line() +
+  geom_point(color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = "Heart rate (bpm)") +
+  expand_limits(x = range(bcg_sample$dt), y = 0) +
+  t
+
+# C: Sway, heave
+
+bcg_sway_heave <- bcg_sample %>% 
+  rename(Sway = sway_filt, Heave = heave_filt) %>% 
+  pivot_longer(c(Sway, Heave), 
+               names_to = "axis",
+               values_to = "acceleration")
+
+p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  facet_grid(axis ~ .) +
+  labs(y = bquote(Acc~(m~s^{-2}))) +
+  t
+
+# D: Jerk
+
+p4 <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = bquote(Jerk~(m~s^{-3}))) +
+  expand_limits(y = 0) +
+  t
+
+cowplot::plot_grid(p1, p2, p3, p4, 
+                   align = "v", 
+                   axis = "lr",
+                   nrow = 4, 
+                   rel_heights = c(1.5, 0.75, 1, 1),
+                   labels = "AUTO")
+```
+
+Figure \@ref(fig:corky-bcg) shows how we can have a caption and cross-reference for a plot.
 

### 2021-08-27 R/find_beats.R

@@ -1,4 +1,4 @@
-#' Find beats in BCG
+#' Find beats in BCG with local surge range cue
 #'
 #' @param surge Filtered surge signal
 #' @param window_s Width of search window in seconds (0.25 s by default)
@@ -6,7 +6,7 @@
 #'
 #' @return A logical vector the same length as surge indicating a heart beat
 #' @export
-find_beats <- function(surge, window_s = 0.25, fs_hz = 400) {
+find_beats_lsr <- function(surge, window_s = 0.25, fs_hz = 400) {
   window <- as.integer(window_s * fs)
 
   # Local range of surge
@@ -28,6 +28,11 @@ find_beats <- function(surge, window_s = 0.25, fs_hz = 400) {
   }
 
   beats
+}
+
+find_beats_jerk <- function(jerk, window_s = 1.8, fs_hz = 10) {
+  window <- as.integer(window_s * fs)
+  jerk_smooth <- tma(jerk, window * 10)
 }
 
 #' Local range of signal

### 2021-08-27 R/util.R

@@ -1,0 +1,12 @@
+#' Norm of the jerk vector
+#'
+#' @param A Tri-axial acceleration (n x 3 matrix)
+#' @param fs Sampling rate (Hz)
+#'
+#' @return Norm of the jerk vector of A
+#' @export
+njerk <- function(A, fs) {
+  jerk <- (A[-1, ] - A[-nrow(A), ]) / fs
+  jerk <- rbind(rep(NA, 3), jerk)
+  apply(jerk, 1, function(row) sqrt(sum(row^2)))
+}

### 2021-08-27 analysis/data/import_bw180905-53.R

@@ -1,0 +1,39 @@
+library(tidyverse)
+
+bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
+tz <- "Etc/GMT+7"
+bw180905_53_400hz <- tibble(
+  dn = as.vector(bw180905_53_mat$DN400Hz),
+  surge = as.vector(bw180905_53_mat$Aw400Hz[, 1]),
+  sway = as.vector(bw180905_53_mat$Aw400Hz[, 2]),
+  heave = as.vector(bw180905_53_mat$Aw400Hz[, 3])
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, surge, sway, heave)
+
+saveRDS(bw180905_53_400hz, "analysis/data/derived_data/bw180905_53_400hz.rds")
+
+# 10 Hz PRH data
+bw180905_53_10hz <- tibble(
+  dn = as.vector(bw180905_53_mat$DN10Hz),
+  depth = as.vector(bw180905_53_mat$depth10Hz),
+  jerk = njerk(as.matrix(bw180905_53_mat$Aw10Hz) * 9.8, 10)
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, depth, jerk)
+
+saveRDS(bw180905_53_10hz, "analysis/data/derived_data/bw180905_53_10hz.rds")
+
+# Regions eligible for BCG
+# Manually identified as motionless periods (bottom phase of dive, no fluking)
+b180905_53_elg <- read_csv("analysis/data/raw_data/bcg_eligible.csv") %>%
+  mutate(across(c(start, stop), ~ lubridate::force_tz(.x, tz))) %>%
+  select(dive, start, stop)
+
+saveRDS(b180905_53_elg, "analysis/data/derived_data/bw180905_53_elg.rds")

### 2021-08-27 analysis/paper/paper.Rmd
Blue whale heart beat detection with jerk cue
Former-commit-id: fd01e15d2ec49284807899fcc372591105ff5c00

@@
 window <- 1.8 #s
+
+bw_elg <- readRDS(here::here("analysis/data/derived_data/bw180905_53_elg.rds"))
+
 bw_400hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_400hz.rds")) %>% 
-  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))
+  mutate(
+    across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"),
+    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
+    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
+    region_id = ifelse(rid_left == rid_right, rid_left, NA)
+  ) %>% 
+  select(-c(rid_left, rid_right))
 
@@
   mutate(jerk_smooth = tma(jerk, window * 10))
-
-bw_elg <- readRDS(here::here("analysis/data/derived_data/bw180905_53_elg.rds"))
 ```
@@
 ```{r bw-beats}
-bw_400hz <- mutate(bw_400hz, surge_range = NA, bcg_beat = NA, bcg_bpm = NA)
+bw_400hz_beats <- bw_400hz %>% 
+  mutate(jerk = approx(bw_10hz$dt, bw_10hz$jerk_smooth, bw_400hz$dt, rule = 2)$y) %>% 
+  filter(!is.na(region_id)) %>%
+  group_by(region_id) %>% 
+  mutate(bcg_beat = find_beats_jerk(jerk, surge_filt, window, fs),
+         bcg_bpm = bpm(bcg_beat, dt)) %>% 
+  ungroup() %>% 
+  select(dt, bcg_beat, bcg_bpm)
 
-for (row in seq(nrow(bw_eligible))) {
-  i <- approx(bw_400hz$dt, 
-              seq_along(bw_400hz$dt), 
-              bw_elg[row, c("start", "stop")])$y
-  i <- floor(i[1]):floor(i[2])
-  region <- bw_400hz[i, ]
-  surge <- region$surge_filt
-  surge_range <- local_range(surge, window *fs)
-  bcg_beat <- find_beats_lsr(surge, window, fs)
-  bcg_bpm <- rep(NA, length(i))
-  if (sum(bcg_beat) > 1) {
-    bcg_dt <- region$dt[bcg_beat]
-    bcg_bpm[bcg_beat] <- 1 / as.numeric(lead(bcg_dt) - bcg_dt, unit = "mins")
-  }
-  bw_400hz$surge_range[i] <- surge_range
-  bw_400hz$bcg_beat[i] <- bcg_beat
-  bw_400hz$bcg_bpm[i] <- bcg_bpm
-}
+bw_400hz <- bw_400hz %>% 
+  mutate(surge_range = local_range(surge_filt, window * fs)) %>% 
+  left_join(bw_400hz_beats, by = "dt")
 
@@
 ```{r bw-bcg}
+bw_bcg_plot <- function(rid) {
   ca_tz <- "Etc/GMT+7"
-bcg_begin <- lubridate::ymd_hms("2018-09-05 21:29:05", tz = ca_tz)
-bcg_end <- lubridate::ymd_hms("2018-09-05 21:30:30", tz = ca_tz)
-bcg_sample <- filter(bw_400hz, between(dt, bcg_begin, bcg_end))
-bcg_sample_10hz <- filter(bw_10hz, between(dt, bcg_begin, bcg_end))
+  bcg_sample <- filter(bw_400hz, region_id == rid)
+  bcg_sample_10hz <- bw_10hz %>% 
+    filter(between(dt, bcg_sample$dt[1], bcg_sample$dt[nrow(bcg_sample)]))
 
@@
 
-cowplot::plot_grid(p1, p2, p3, p4, 
+  p <- cowplot::plot_grid(p1, p2, p3, p4, 
                           align = "v", 
@@
                           labels = "AUTO")
+
+  ggsave(here::here(sprintf("analysis/figures/bw_bcg/bw_bcg_%02i.png", rid)), p)
+}
+
+foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
 ```

### 2021-08-27 analysis/data/import_bw180905-53.R

@@ -1,10 +1,13 @@
 library(tidyverse)
 
-bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
 tz <- "Etc/GMT+7"
+
+bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
+
+# 400 Hz acceleration
 bw180905_53_400hz <- tibble(
-  dn = as.vector(bw180905_53_mat$DN400Hz),
   surge = as.vector(bw180905_53_mat$Aw400Hz[, 1]),
+  dn = as.vector(bw180905_53_mat$DN400Hz),
   sway = as.vector(bw180905_53_mat$Aw400Hz[, 2]),
   heave = as.vector(bw180905_53_mat$Aw400Hz[, 3])
 ) %>%
@@ -33,7 +36,8 @@ saveRDS(bw180905_53_10hz, "analysis/data/derived_data/bw180905_53_10hz.rds")
 # Regions eligible for BCG
 # Manually identified as motionless periods (bottom phase of dive, no fluking)
 b180905_53_elg <- read_csv("analysis/data/raw_data/bcg_eligible.csv") %>%
-  mutate(across(c(start, stop), ~ lubridate::force_tz(.x, tz))) %>%
-  select(dive, start, stop)
+  mutate(across(c(start, stop), ~ lubridate::force_tz(.x, tz)),
+         region_id = row_number()) %>%
+  select(dive, region_id, start, stop)
 
 saveRDS(b180905_53_elg, "analysis/data/derived_data/bw180905_53_elg.rds")

### 2021-08-27 R/find_beats.R
Blue whale heart beat detection with jerk cue

@@ -7,7 +7,7 @@
 #' @return A logical vector the same length as surge indicating a heart beat
 #' @export
 find_beats_lsr <- function(surge, window_s = 0.25, fs_hz = 400) {
-  window <- as.integer(window_s * fs)
+  window <- floor(window_s * fs)
 
   # Local range of surge
   surge_range <- local_range(surge, window)
@@ -28,11 +28,6 @@ find_beats_lsr <- function(surge, window_s = 0.25, fs_hz = 400) {
   }
 
   beats
-}
-
-find_beats_jerk <- function(jerk, window_s = 1.8, fs_hz = 10) {
-  window <- as.integer(window_s * fs)
-  jerk_smooth <- tma(jerk, window * 10)
 }
 
 #' Local range of signal
@@ -60,4 +55,51 @@ tma <- function(x, k) {
   x %>%
     RcppRoll::roll_mean(k, fill = NA) %>%
     RcppRoll::roll_mean(k, fill = NA)
+}
+
+#' Find beats in BCG with jerk cue
+#'
+#' @param jerk Smoothed jerk signal
+#' @param surge Filtered surge signal
+#' @param window_s Width of search window in seconds (1.8 s by default)
+#' @param fs_hz Sampling frequency of jerk and surge in Hz (400 Hz by default)
+#'
+#' @return
+#' @export
+find_beats_jerk <- function(jerk, surge, window_s = 1.8, fs_hz = 400){
+  window <- floor(window_s * fs)
+
+  # Find local peaks in jerk
+  jerkpeaks <- pracma::findpeaks(jerk, minpeakdistance = window)
+  jerk_hts <- jerkpeaks[, 1]
+  jerk_pks <- jerkpeaks[, 2]
+  jerk_prm <- peak_prominences(jerk, jerk_pks)
+
+  # Find major peaks
+  if (length(jerk_hts) == 0) {
+    return(logical(length(jerk)))
+  } else if (length(jerk_hts) == 1) {
+    major_peaks <- jerk_pks
+  } else {
+    kclusts <- kmeans(cbind(c(0, jerk_hts), c(0, jerk_prm)), centers = 2)
+    clust_order <- apply(kclusts$centers, 2, diff)
+    if (sign(clust_order[1]) != sign(clust_order[2])) {
+      stop("Height and prominence thresholds don't agree.")
+    }
+    major_cluster <- which.max(kclusts$centers[, 1])
[…]

### 2021-08-27 R/util.R

@@ -10,3 +10,55 @@ njerk <- function(A, fs) {
   jerk <- rbind(rep(NA, 3), jerk)
   apply(jerk, 1, function(row) sqrt(sum(row^2)))
 }
+
+#' Calculate beats-per-minute
+#'
+#' @param beats Logical vector indicating heart beats
+#' @param dt POSIXct vector of timestamps
+#'
+#' @return Where `beats` is TRUE, the BPM. Otherwise NA.
+#' @export
+bpm <- function(beats, dt) {
+  result <- rep(NA, length(beats))
+  beats_i <- which(beats)
+  beats_dt <- dt[beats_i]
+  next_beat <- c(dt[beats_i[-1]], NA)
+  result[beats] <- 1 / as.numeric(next_beat - beats_dt, unit = "mins")
+  result
+}
+
+#' Find prominences of peaks
+#'
+#' @param x Vector of numeric values
+#' @param peaks Indices of peak locations
+#'
+#' @return Prominences of peaks in `x` at indices `peaks`
+#' @export
+peak_prominences <- function(x, peaks) {
+  result <- numeric(length(peaks))
+  for (i in seq_along(peaks)) {
+    # Highest peak case
+    if (x[peaks[i]] == max(x[peaks])) {
+      result[i] <- x[peaks[i]] - min(x, na.rm = TRUE)
+      next
+    }
+
+    # Find closest, higher peaks and the valleys between
+    higher <- which(x[peaks] >= x[peaks[i]])
+    higher_left <- suppressWarnings(max(higher[higher < i]))
+    if (is.infinite(higher_left)) {
+      valley_left <- min(x[1:peaks[i]], na.rm = TRUE)
+    } else {
+      valley_left <- min(x[peaks[higher_left]:peaks[i]], na.rm = TRUE)
+    }
+    higher_right <- suppressWarnings(min(higher[higher > i]))
+    if (is.infinite(higher_right)) {
+      valley_right <- min(x[peaks[i]:length(x)], na.rm = TRUE)
+    } else {
+      valley_right <- min(x[peaks[i]:peaks[higher_right]], na.rm = TRUE)
+    }
+
+    result[i] <- x[peaks[i]] - max(valley_left, valley_right)
+  }
+  result
+}

### 2021-08-30 R/data.R
move data from "analysis/data/" level to package level
Former-commit-id: 877589b8d80f80084c939e861fd90dff6dd9e622

@@ -1,0 +1,21 @@
+#' Blue whale 10 Hz depth and acceleration
+#'
+#' Note to self: add PRH
+"bw180905_53_10hz"
+
+#' Blue whale 400 Hz acceleration
+"bw180905_53_400hz"
+
+#' Blue whale BCG eligible regions
+"bw180905_53_elg"
+
+#' Killer whale 10 Hz body orientation and jerk
+#'
+#' Note to self: add depth, Aw. Remove jerk.
+"corky_10hz"
+
+#' Killer whale 400 Hz acceleration
+"corky_400hz"
+
+#' Killer whale heart beats recorded by ECG
+"ecg_beats"

### 2021-08-30 data-raw/bw180905_53_elg.R

@@ -1,0 +1,13 @@
+## code to prepare `bw180905_53_elg` dataset goes here
+library(tidyverse)
+
+tz <- "Etc/GMT+7"
+
+# Regions eligible for BCG
+# Manually identified as motionless periods (bottom phase of dive, no fluking)
+bw180905_53_elg <- read_csv("analysis/data/raw_data/bcg_eligible.csv") %>%
+  mutate(across(c(start, stop), ~ lubridate::force_tz(.x, tz)),
+         region_id = row_number()) %>%
+  select(dive, region_id, start, stop)
+
+usethis::use_data(bw180905_53_elg, overwrite = TRUE)

### 2021-08-30 data-raw/corky_10hz.R

@@ -1,0 +1,21 @@
+## code to prepare `corky_10hz` dataset goes here
+library(tidyverse)
+
+corky_10hz_mat <- R.matlab::readMat(
+  "analysis/data/raw_data/corky210816-97 10Hzprh.mat"
+)
+corky_10hz <- tibble(
+  dn = as.vector(corky_10hz_mat$DN),
+  aw = as.matrix(corky_10hz_mat$Aw),
+  pitch = as.vector(corky_10hz_mat$pitch),
+  roll = as.vector(corky_10hz_mat$roll),
+  head = as.vector(corky_10hz_mat$head)
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz),
+         jerk = njerk(aw * 9.8, 10)) %>%
+  select(dt, pitch, roll, head, jerk)
+
+usethis::use_data(corky_10hz, overwrite = TRUE)

### 2021-08-30 data-raw/corky_400hz.R

@@ -1,0 +1,20 @@
+## code to prepare `corky_400hz` dataset goes here
+library(tidyverse)
+
+corky_400hz_mat <- R.matlab::readMat(
+  "analysis/data/raw_data/20210816-134844-CATS 97.mat"
+)
+tz <- "Etc/GMT+7"
+corky_400hz <- tibble(
+  dn = as.vector(corky_400hz_mat$Atime),
+  surge = as.vector(corky_400hz_mat$Adata[, 1]),
+  sway = as.vector(corky_400hz_mat$Adata[, 2]),
+  heave = as.vector(corky_400hz_mat$Adata[, 3])
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, surge, sway, heave)
+
+usethis::use_data(corky_400hz, overwrite = TRUE)

### 2021-08-30 data-raw/ecg_beats.R

@@ -1,0 +1,24 @@
+## code to prepare `ecg_beats` dataset goes here
+library(tidyverse)
+
+set_date <- function(datetime, new_date) {
+  result <- rep(new_date, length(datetime))
+  lubridate::hour(result) <- lubridate::hour(datetime)
+  lubridate::minute(result) <- lubridate::minute(datetime)
+  lubridate::second(result) <- lubridate::second(datetime)
+  result
+}
+
+tz <- "Etc/GMT+7"
+corky_date <- lubridate::ymd_hm("2021-08-16 00:00", tz = tz)
+
+ecg_beats <- readxl::read_excel(
+  "analysis/data/raw_data/Corky 8-16-21 Heart rate data for BCG test.xlsx",
+  col_names = FALSE,
+  na = "--"
+) %>%
+  select(1:3) %>%
+  set_names(c("dt", "ecg", "bpm")) %>%
+  mutate(dt = set_date(dt, corky_date))
+
+usethis::use_data(ecg_beats, overwrite = TRUE)

### 2021-08-30 analysis/data/import_bw180905-53.R
move data from "analysis/data/" level to package level

@@ -1,43 +1,0 @@
-library(tidyverse)
-
-tz <- "Etc/GMT+7"
-
-bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
-
-# 400 Hz acceleration
-bw180905_53_400hz <- tibble(
-  surge = as.vector(bw180905_53_mat$Aw400Hz[, 1]),
-  dn = as.vector(bw180905_53_mat$DN400Hz),
-  sway = as.vector(bw180905_53_mat$Aw400Hz[, 2]),
-  heave = as.vector(bw180905_53_mat$Aw400Hz[, 3])
-) %>%
-  mutate(dt = as.POSIXct((dn - 719529) * 86400,
-                         origin = "1970-01-01",
-                         tz = "UTC") %>%
-           lubridate::force_tz(tz)) %>%
-  select(dt, surge, sway, heave)
-
-saveRDS(bw180905_53_400hz, "analysis/data/derived_data/bw180905_53_400hz.rds")
-
-# 10 Hz PRH data
-bw180905_53_10hz <- tibble(
-  dn = as.vector(bw180905_53_mat$DN10Hz),
-  depth = as.vector(bw180905_53_mat$depth10Hz),
-  jerk = njerk(as.matrix(bw180905_53_mat$Aw10Hz) * 9.8, 10)
-) %>%
-  mutate(dt = as.POSIXct((dn - 719529) * 86400,
-                         origin = "1970-01-01",
-                         tz = "UTC") %>%
-           lubridate::force_tz(tz)) %>%
-  select(dt, depth, jerk)
-
-saveRDS(bw180905_53_10hz, "analysis/data/derived_data/bw180905_53_10hz.rds")
-
-# Regions eligible for BCG
-# Manually identified as motionless periods (bottom phase of dive, no fluking)
-b180905_53_elg <- read_csv("analysis/data/raw_data/bcg_eligible.csv") %>%
-  mutate(across(c(start, stop), ~ lubridate::force_tz(.x, tz)),
-         region_id = row_number()) %>%
-  select(dive, region_id, start, stop)
-
-saveRDS(b180905_53_elg, "analysis/data/derived_data/bw180905_53_elg.rds")

### 2021-08-30 analysis/data/import_corky.R

@@ -1,65 +1,0 @@
-library(tidyverse)
-
-# 400 Hz acceleration data
-acc_400hz_mat <- R.matlab::readMat(
-  "analysis/data/raw_data/20210816-134844-CATS 97.mat"
-)
-tz <- "Etc/GMT+7"
-acc_400hz <- tibble(
-  dn = as.vector(acc_400hz_mat$Atime),
-  surge = as.vector(acc_400hz_mat$Adata[, 1]),
-  sway = as.vector(acc_400hz_mat$Adata[, 2]),
-  heave = as.vector(acc_400hz_mat$Adata[, 3])
-) %>%
-  mutate(dt = as.POSIXct((dn - 719529) * 86400,
-                         origin = "1970-01-01",
-                         tz = "UTC") %>%
-           lubridate::force_tz(tz)) %>%
-  select(dt, surge, sway, heave)
-
-saveRDS(acc_400hz, "analysis/data/derived_data/acc_400hz.rds")
-
-# 10 Hz PRH data
-njerk <- function(A, fs) {
-  jerk <- (A[-1, ] - A[-nrow(A), ]) / fs
-  jerk <- rbind(rep(NA, 3), jerk)
-  apply(jerk, 1, function(row) sqrt(sum(row^2)))
-}
-prh_10hz_mat <- R.matlab::readMat(
-  "analysis/data/raw_data/corky210816-97 10Hzprh.mat"
-)
-prh_10hz <- tibble(
-  dn = as.vector(prh_10hz_mat$DN),
-  aw = as.matrix(prh_10hz_mat$Aw),
-  pitch = as.vector(prh_10hz_mat$pitch),
-  roll = as.vector(prh_10hz_mat$roll),
-  head = as.vector(prh_10hz_mat$head)
-) %>%
-  mutate(dt = as.POSIXct((dn - 719529) * 86400,
-                         origin = "1970-01-01",
-                         tz = "UTC") %>%
-           lubridate::force_tz(tz),
-         jerk = njerk(aw * 9.8, 10)) %>%
-  select(dt, pitch, roll, head, jerk)
-
-saveRDS(prh_10hz, "analysis/data/derived_data/prh_10hz.rds")
-
-# ECG data
-set_date <- function(datetime, new_date) {
-  result <- rep(new_date, length(datetime))
-  lubridate::hour(result) <- lubridate::hour(datetime)
-  lubridate::minute(result) <- lubridate::minute(datetime)
[…]

### 2021-08-30 data-raw/bw180905_53_10hz.R

@@ -1,0 +1,20 @@
+## code to prepare `bw180905_53_10hz` dataset goes here
+library(tidyverse)
+
+tz <- "Etc/GMT+7"
+
+bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
+
+# 10 Hz PRH data
+bw180905_53_10hz <- tibble(
+  dn = as.vector(bw180905_53_mat$DN10Hz),
+  depth = as.vector(bw180905_53_mat$depth10Hz),
+  aw = as.matrix(bw180905_53_mat$Aw10Hz)
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, depth, aw)
+
+usethis::use_data(bw180905_53_10hz, overwrite = TRUE)

### 2021-08-30 data-raw/bw180905_53_400hz.R

@@ -1,0 +1,21 @@
+## code to prepare `bw180905_53_400hz` dataset goes here
+library(tidyverse)
+
+tz <- "Etc/GMT+7"
+
+bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
+
+# 400 Hz acceleration
+bw180905_53_400hz <- tibble(
+  surge = as.vector(bw180905_53_mat$Aw400Hz[, 1]),
+  dn = as.vector(bw180905_53_mat$DN400Hz),
+  sway = as.vector(bw180905_53_mat$Aw400Hz[, 2]),
+  heave = as.vector(bw180905_53_mat$Aw400Hz[, 3])
+) %>%
+  mutate(dt = as.POSIXct((dn - 719529) * 86400,
+                         origin = "1970-01-01",
+                         tz = "UTC") %>%
+           lubridate::force_tz(tz)) %>%
+  select(dt, surge, sway, heave)
+
+usethis::use_data(bw180905_53_400hz, overwrite = TRUE)

### 2021-08-30 analysis/interact/ui.R
Add shiny app for tuning hyperparameters
Former-commit-id: c44be668d72c1c5391ed7f5393fe65a2b3def9db

@@ -1,0 +1,44 @@
+ui <- fillPage(
+
+  fillRow(
+
+    fillCol(
+      h1("Tune BCG Hyperparameters"),
+      numericInput("filt_thr_lower",
+                   "Filter lower threshold",
+                   value = filter_thr0[1],
+                   min = 0, max = 50),
+      numericInput("filt_thr_upper",
+                   "Filter upper threshold",
+                   value = filter_thr0[2],
+                   min = 0, max = 50),
+      numericInput("filt_order",
+                   "Filter order",
+                   value = filter_order0,
+                   min = 1, max = 40,
+                   step = 1),
+      numericInput("window_size",
+                   "Window size (seconds)",
+                   value = window0,
+                   min = 0.25, max = 10),
+      numericInput("yaw_offset",
+                   "Yaw offset (degrees)",
+                   value = yaw_offset0,
+                   min = -180, max = 180),
+      actionButton("update",
+                   "Update")
+    ),
+
+    fillCol(
+      plotOutput("depth_plot", click = "depth_click", height = "100%"),
+      plotOutput("dive_plot", click = "dive_click", height = "100%"),
+      plotOutput("bcg_plot", height = "100%"),
+      flex = c(1, 1, 2)
+    ),
+
+    flex = c(1, 3)
+  ),
+
+  padding = 10
+
+)

### 2021-08-30 analysis/interact/global.R
Add shiny app for tuning hyperparameters

@@ -1,0 +1,206 @@
+library(shinybusy)
+
+filter_thr0 <- c(1.5, 7.5)
+filter_order0 <- 4
+window0 <- 1.8
+yaw_offset0 <- 0
+
+n_pts <- 1e3
+bw180905_53_10hz_sparse <- bw180905_53_10hz[floor(seq(1, nrow(bw180905_53_10hz), length.out = n_pts)), ]
+
+update_results <- function(hyperparams) {
+  show_modal_progress_circle(.10, "Filtering acceleration")
+
+  window <- hyperparams$window
+  fs <- 400
+  fny <- fs / 2
+  acc_filter <- signal::butter(hyperparams$filter_order,
+                               hyperparams$filter_thr / fny,
+                               type = "pass")
+  apply_filter <- function(x) signal::filtfilt(acc_filter, x)
+
+  rid_fun <- function(x, xout) {
+    approx(x, bw180905_53_elg$region_id, xout, "constant", yleft = 0)$y
+  }
+  bw_400hz <- bw180905_53_400hz %>%
+    mutate(
+      across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"),
+      rid_left = rid_fun(bw180905_53_elg$start, dt),
+      rid_right = rid_fun(bw180905_53_elg$stop, dt) + 1,
+      region_id = ifelse(rid_left == rid_right, rid_left, NA),
+      jerk = njerk(cbind(surge_filt, sway_filt, heave_filt), fs),
+      jerk_smooth = tma(jerk, window * fs)
+    )  %>%
+    select(-c(rid_left, rid_right))
+
+  update_modal_progress(.40, "Finding heartbeats")
+
+  bw_400hz_beats <- bw_400hz %>%
+    filter(!is.na(region_id)) %>%
+    group_by(region_id) %>%
+    mutate(surge_range = local_range(surge_filt, window * fs),
+           bcg_beat = find_beats_jerk(jerk_smooth, surge_filt, window, fs),
+           bcg_bpm = bpm(bcg_beat, dt)) %>%
+    ungroup() %>%
+    select(dt, surge_range, bcg_beat, bcg_bpm)
+
+  update_modal_progress(.80, "Wrapping up")
+
+  bw_400hz <- bw_400hz %>%
+    left_join(bw_400hz_beats, by = "dt")
+
+  bw_10hz <- bw180905_53_10hz
[…]

### 2021-08-30 analysis/interact/server.R

@@ -1,0 +1,55 @@
+library(cetaceanbcg)
+library(shiny)
+library(tidyverse)
+
+server <- function(input, output, session) {
+  # Hyperparameters
+  hyperparams <- reactive(list(
+    filter_thr = c(input$filt_thr_lower, input$filt_thr_upper),
+    filter_order = as.integer(input$filt_order),
+    window = input$window,
+    yaw_offset = input$yaw_offset * pi / 180
+  ))
+
+  # Data
+  bw_data <- reactiveVal(update_results(list(
+    filter_thr = filter_thr0,
+    filter_order = filter_order0,
+    window = window0,
+    yaw_offset = yaw_offset0
+  )))
+
+  # I/O
+  depth_click <- reactiveValues(x = NULL, y = NULL)
+  dive_click <- reactiveValues(x = NULL, y = NULL)
+  observeEvent(input$update,
+               bw_data(update_results(hyperparams())))
+  observeEvent(input$depth_click, {
+    depth_click$x <- convert_clickx(input$depth_click$x)
+    depth_click$y <- input$depth_click$y
+    dive_click$x <- NULL
+    dive_click$y <- NULL
+  })
+  observeEvent(input$dive_click, {
+    dive_click$x <- convert_clickx(input$dive_click$x)
+    dive_click$y <- input$dive_click$y
+  })
+
+  # Plots
+  output$depth_plot <- renderPlot(
+    ggplot(bw180905_53_10hz_sparse, aes(dt, depth)) +
+      geom_line(size = 0.2) +
+      annotate_click(depth_click$x, depth_click$y) +
+      scale_x_datetime(date_labels = "%H:%M:%S") +
+      scale_y_reverse() +
+      theme_minimal()
+  )
+  # Why do I have to do bw_data()[2]$bw_10hz? No clue why bw_data()$bw_10hz
+  # doesn't work
+  output$dive_plot <- renderPlot(
+    bw_dive_plot(bw_data()[2]$bw_10hz, depth_click, dive_click)
+  )
+  output$bcg_plot <- renderPlot(
+    bw_bcg_plot(bw_data()[1]$bw_400hz, dive_click)
+  )
+}

### 2021-08-30 R/find_beats.R
Fix incorrect variable name in find_beats_jerk()
Former-commit-id: 0f76774ef759860b0516db0010b87d6e5e9ea37f

@@ -67,7 +67,7 @@ tma <- function(x, k) {
 #' @return
 #' @export
 find_beats_jerk <- function(jerk, surge, window_s = 1.8, fs_hz = 400){
-  window <- floor(window_s * fs)
+  window <- floor(window_s * fs_hz)
 
   # Find local peaks in jerk
   jerkpeaks <- pracma::findpeaks(jerk, minpeakdistance = window)

### 2021-08-31 R/data.R
Add ECG data

@@ -19,3 +19,6 @@
 
 #' Killer whale heart beats recorded by ECG
 "ecg_beats"
+
+#' Killer whale ECG
+"corky_ecg"

### 2021-08-31 data-raw/corky_ecg.R

@@ -1,0 +1,20 @@
+## code to prepare `corky_ecg` dataset goes here
+
+# .UBC is a somewhat funky text file produced by the ECG recorder's software
+corky_ecg <- suppressWarnings(readLines("analysis/data/raw_data/Corky 08-16-21 old float Plus K unit  for BCG test .ubc"))
+# Drop the header lines
+corky_ecg <- corky_ecg[-c(1, 2)]
+# Replace weird control characters with "|", which will be the column delimiter
+corky_ecg <- gsub("[^0-9:/.]", "|", corky_ecg)
+# Convert to data.frame. Columns are date, time, (uncalibrated) ecg
+corky_ecg <- strsplit(corky_ecg, "[|]+")
+corky_ecg <- data.frame(matrix(unlist(corky_ecg), ncol = 3, byrow = TRUE))
+# Parse datetime
+tz <- "Etc/GMT+7"
+corky_ecg$dt <- lubridate::mdy_hms(paste(corky_ecg[, 1], corky_ecg[, 2]), tz = tz)
+# Calibrate ECG data using scaling factor of 0.15
+corky_ecg$ecg_uV <- as.numeric(corky_ecg[, 3]) * 0.15
+# Keep only datetime and ecg
+corky_ecg <- corky_ecg[, c("dt", "ecg_uV")]
+
+usethis::use_data(corky_ecg, overwrite = TRUE)

### 2021-08-31 data-raw/corky_ecg.R
Add heart rates to `corky_ecg`

@@ -1,20 +1,38 @@
+library(tidyverse)
+
 ## code to prepare `corky_ecg` dataset goes here
 
 # .UBC is a somewhat funky text file produced by the ECG recorder's software
-corky_ecg <- suppressWarnings(readLines("analysis/data/raw_data/Corky 08-16-21 old float Plus K unit  for BCG test .ubc"))
-# Drop the header lines
-corky_ecg <- corky_ecg[-c(1, 2)]
-# Replace weird control characters with "|", which will be the column delimiter
-corky_ecg <- gsub("[^0-9:/.]", "|", corky_ecg)
-# Convert to data.frame. Columns are date, time, (uncalibrated) ecg
-corky_ecg <- strsplit(corky_ecg, "[|]+")
-corky_ecg <- data.frame(matrix(unlist(corky_ecg), ncol = 3, byrow = TRUE))
-# Parse datetime
+ubc_path <- "analysis/data/raw_data/Corky 08-16-21 old float Plus K unit  for BCG test .ubc"
 tz <- "Etc/GMT+7"
-corky_ecg$dt <- lubridate::mdy_hms(paste(corky_ecg[, 1], corky_ecg[, 2]), tz = tz)
-# Calibrate ECG data using scaling factor of 0.15
-corky_ecg$ecg_uV <- as.numeric(corky_ecg[, 3]) * 0.15
-# Keep only datetime and ecg
-corky_ecg <- corky_ecg[, c("dt", "ecg_uV")]
+corky_ecg <- suppressWarnings(readLines(ubc_path)) %>%
+  # Drop the header lines
+  `[`(-c(1, 2)) %>%
+  # Replace weird control characters with "|", which will be the column delimiter
+  gsub("[^0-9:/.]", "|", .) %>%
+  # Convert to data.frame. Columns are date, time, (uncalibrated) ecg
+  strsplit("[|]+") %>%
+  { data.frame(matrix(unlist(.), ncol = 3, byrow = TRUE)) } %>%
+  # Clean up data.frame
+  set_names(c("date", "time", "ecg")) %>%
+  mutate(dt = lubridate::mdy_hms(paste(date, time), tz = tz),
+         ecg_uV = as.numeric(ecg) * 0.15) %>%
+  select(dt, ecg_uV)
+
+# Annotate with heartbeats. Adjust time up to 0.5 s to match R-wave peak.
[…]

### 2021-08-31 analysis/paper/paper.Rmd
Split Corky BCG figure into two

@@
   fig.path = "../figures/",
+  fig.height = 6,
+  fig.width = 6,
   dpi = 300
@@
 ```{r process-acc}
-acc_400hz <- readRDS(here::here("analysis/data/derived_data/acc_400hz.rds"))
+acc_400hz <- corky_400hz
 fs <- 400
@@
 window <- 0.25
-prh_10hz <- readRDS(here::here("analysis/data/derived_data/prh_10hz.rds")) %>% 
+prh_10hz <- corky_10hz %>% 
   mutate(jerk_smooth = tma(jerk, window * 10))
@@
 ```{r align-ecg}
-ecg_beats <- readRDS(here::here("analysis/data/derived_data/ecg_beats.rds"))
-
 acc_400hz$ecg_beat <- FALSE
@@
 
-```{r corky-bcg, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
-Solid line: surge bandpass filtered to [1.0 25.0] Hz.
-Dashed line: local range of surge in 0.25 s sliding window (smoothed).
-Blue dots indicate peak of the BCG I-wave.
-B: Heart rates estimated from BCG (blue) match those from ECG (red).
-Temporal offset is due to both physiology and sensor clock differences.
-C: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge.
-The sway signal is likely due to movement of the chest cavity (see video S1).
-D: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
+```{r bcg-figs}
 ca_tz <- "Etc/GMT+7"
@@
         panel.grid.minor = element_blank())
+```
 
+```{r corky-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
 # A: Surge + LSR
@@
 
-p1 <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
+pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
   geom_line(size = 0.2) +
@@
   labs(y = bquote(Surge~(m~s^{-2}))) +
+  expand_limits(x = max(bcg_sample$dt) + 1) +
   t +
@@
 
-# B: BPM
+# B: ECG
+ecg_sample <- corky_ecg %>% 
+  filter(between(dt, first(bcg_sample$dt), last(bcg_sample$dt)))
+pB <- ggplot(ecg_sample, aes(dt, ecg_uV)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(ecg_sample, ecg_beat), color = "red") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = bquote(ECG~(mu * V))) +
+  expand_limits(x = max(bcg_sample$dt) + 1) +
+  t
 
-p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
-  geom_line() +
+# C: BCG + ECG
+pC <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
+  geom_line(size = 0.2) +
+  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat)) +
   geom_point(color = "blue") +
-  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat)) +
   geom_point(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), color = "red") +
@@
   labs(y = "Heart rate (bpm)") +
-  expand_limits(x = range(bcg_sample$dt), y = c(30, 70)) +
+  expand_limits(x = range(bcg_sample$dt) + c(0, 1), y = c(30, 70)) +
   t
 
-# C: Sway, heave
+cowplot::plot_grid(pA, pB, pC,
+                   align = "v", 
+                   axis = "lr",
+                   nrow = 3, 
+                   rel_heights = c(1.5, 1.0, 0.5),
+                   labels = "AUTO")
+```
 
+```{r corky-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
+Solid line: surge bandpass filtered to [1.0 25.0] Hz.
+Dashed line: local range of surge in 0.25 s sliding window (smoothed).
+Blue dots indicate peak of the BCG I-wave.
+B: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge.
+The sway signal is likely due to movement of the chest cavity (see video S1).
+C: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
+# A: Surge + LSR
+# pA as above
+
+# B: Sway, heave
 bcg_sway_heave <- bcg_sample %>% 
@@
                values_to = "acceleration")
-
-p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
+pB <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
   geom_line(size = 0.2) +
@@
 
-# D: Jerk
-
-p4 <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
+# C: Jerk
+pC <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
   geom_line(size = 0.2) +
@@
 
-cowplot::plot_grid(p1, p2, p3, p4, 
+cowplot::plot_grid(pA, pB, pC,
                    align = "v", 
                    axis = "lr",
-                   nrow = 4, 
-                   rel_heights = c(1.5, 0.75, 1, 1),
+                   nrow = 3, 
+                   # rel_heights = c(1, 1, ),
                    labels = "AUTO")
@@
 
-bw_elg <- readRDS(here::here("analysis/data/derived_data/bw180905_53_elg.rds"))
+bw_elg <- bw180905_53_elg
 
-bw_400hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_400hz.rds")) %>% 
+bw_400hz <- bw180905_53_400hz %>% 
   mutate(
@@
 
-bw_10hz <- readRDS(here::here("analysis/data/derived_data/bw180905_53_10hz.rds")) %>% 
-  mutate(jerk_smooth = tma(jerk, window * 10))
+bw_10hz <- bw180905_53_10hz %>% 
+  mutate(jerk = njerk(aw, 10),
+         jerk_smooth = tma(jerk, window * 10))
 ```
@@
 
-foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
+# foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
 ```
 
-Figure \@ref(fig:corky-bcg) shows how we can have a caption and cross-reference for a plot.
-
-Here is an example of inline code `r x` in the middle of a sentence.
+Figure \@ref(fig:corky-bcg-ecg) shows how we can have a caption and cross-reference for a plot.
 

### 2021-09-03 R/find_beats.R
Switch from clustering to thresholding approach for identifying "major" jerk peaks

@@ -66,7 +66,8 @@ tma <- function(x, k) {
 #'
 #' @return
 #' @export
-find_beats_jerk <- function(jerk, surge, window_s = 1.8, fs_hz = 400){
+find_beats_jerk <- function(jerk, mask, window_s = 1.8, fs_hz = 400) {
+  jerk[!mask] <- 0
   window <- floor(window_s * fs_hz)
 
   # Find local peaks in jerk
@@ -76,30 +77,16 @@ find_beats_jerk <- function(jerk, surge, window_s = 1.8, fs_hz = 400){
   jerk_prm <- peak_prominences(jerk, jerk_pks)
 
   # Find major peaks
-  if (length(jerk_hts) == 0) {
-    return(logical(length(jerk)))
-  } else if (length(jerk_hts) == 1) {
-    major_peaks <- jerk_pks
-  } else {
-    kclusts <- kmeans(cbind(c(0, jerk_hts), c(0, jerk_prm)), centers = 2)
-    clust_order <- apply(kclusts$centers, 2, diff)
-    if (sign(clust_order[1]) != sign(clust_order[2])) {
-      stop("Height and prominence thresholds don't agree.")
-    }
-    major_cluster <- which.max(kclusts$centers[, 1])
-    major_peaks <- jerk_pks[kclusts$cluster[-1] == major_cluster]
-  }
 
-  # Align jerk peaks with surge peaks
+  # Distance from largest peak in height/prominence space
+  dist <- sqrt((max(jerk_hts) - jerk_hts)^2 + (max(jerk_prm) - jerk_prm)^2)
+  # Keep all peaks up to knee in dist
+  dist2 <- signal::sgolayfilt(sort(dist), p = 2, n = 25, m = 2)
+  knee <- which.max(dist2)
+  dist_thr <- sort(dist)[knee]
+  major_peaks <- jerk_pks[dist < dist_thr]
+
   beats <- logical(length(jerk))
-  i <- 1
-  for (peak in major_peaks) {
-    i1 <- max(peak - window, 1)
-    i2 <- min(peak + window, length(surge))
-    b <- which.max(surge[i1:i2]) + i1 - 1
-    beats[b] <- TRUE
-    i <- i + 1
-  }
-
+  beats[major_peaks] <- TRUE
   beats
 }

### 2021-09-03 R/util.R
Fix bug in peak_prominences due to unordered peaks and use Savitzky-Golay filtering for calculating jerk to reduce noise effects

@@ -1,13 +1,17 @@
 #' Norm of the jerk vector
 #'
+#' Uses Savitzky-Golay filtering to de-noise before differentiating acceleration
+#'
 #' @param A Tri-axial acceleration (n x 3 matrix)
 #' @param fs Sampling rate (Hz)
+#' @param p Order of Savitzky-Golay filter (3 by default)
+#' @param n Window size of Savitzky-Golay filter (11 by default)
 #'
 #' @return Norm of the jerk vector of A
 #' @export
-njerk <- function(A, fs) {
-  jerk <- (A[-1, ] - A[-nrow(A), ]) / fs
-  jerk <- rbind(rep(NA, 3), jerk)
+njerk <- function(A, fs, p = 3, n = 11) {
+  n <- n + 1 - n %% 2
+  jerk <- apply(A, 2, function(axis) signal::sgolayfilt(axis, p, n, m = 1))
   apply(jerk, 1, function(row) sqrt(sum(row^2)))
 }
 
@@ -35,30 +39,33 @@ bpm <- function(beats, dt) {
 #' @return Prominences of peaks in `x` at indices `peaks`
 #' @export
 peak_prominences <- function(x, peaks) {
-  result <- numeric(length(peaks))
-  for (i in seq_along(peaks)) {
+  # Peaks have to be in ascending order for finding adjacent valleys
+  peaks2 <- sort(peaks)
+  result <- numeric(length(peaks2))
+  for (i in seq_along(peaks2)) {
     # Highest peak case
-    if (x[peaks[i]] == max(x[peaks])) {
-      result[i] <- x[peaks[i]] - min(x, na.rm = TRUE)
+    if (x[peaks2[i]] == max(x[peaks2])) {
+      result[i] <- x[peaks2[i]] - min(x, na.rm = TRUE)
       next
     }
 
     # Find closest, higher peaks and the valleys between
-    higher <- which(x[peaks] >= x[peaks[i]])
+    higher <- which(x[peaks2] >= x[peaks2[i]])
     higher_left <- suppressWarnings(max(higher[higher < i]))
     if (is.infinite(higher_left)) {
-      valley_left <- min(x[1:peaks[i]], na.rm = TRUE)
+      valley_left <- min(x[1:peaks2[i]], na.rm = TRUE)
     } else {
-      valley_left <- min(x[peaks[higher_left]:peaks[i]], na.rm = TRUE)
[…]

### 2021-09-03 analysis/paper/paper.Rmd
Refactor paper to incorporate changes in filtering steps

@@
 ```{r process-acc}
-acc_400hz <- corky_400hz
 fs <- 400
 fny <- fs / 2
-acc_filter <- signal::butter(4, c(1.0, 25.0) / fny, type = "pass") 
-apply_filter <- function(x) signal::filtfilt(acc_filter, x)
-acc_400hz <- acc_400hz %>% 
-  mutate(across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"))
-
-# This window (in seconds) will be used for smoothing multiple signals and in
-# heartbeat detection
-window <- 0.25
-prh_10hz <- corky_10hz %>% 
-  mutate(jerk_smooth = tma(jerk, window * 10))
+window <- 0.25 #s
+bandpass_filter <- function(x, pass) {
+  signal::butter(4, pass / fny, type = "pass") %>% 
+    signal::filtfilt(x)
+}
+jerk_filter <- function(x, size) {
+  signal::sgolayfilt(x, p = 4, n = size + (size %% 2) + 1, m = 1)
+}
+oo_400hz <- corky_400hz %>% 
+  mutate(
+    # Pass-band filter acceleration
+    across(surge:heave, 
+           bandpass_filter, pass = c(1.0, 25.0), 
+           .names = "{.col}_filt"),
+    # Detect heart beats using BCG's local surge range
+    surge_range = local_range(surge_filt, window * fs),
+    bcg_beat = find_beats_lsr(surge_filt, window, fs),
+    bcg_bpm = bpm(bcg_beat, dt),
+    # Calculate jerk norm and apply smoother
+    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
+                  p = 4, n = window * fs),
+    njerk_smooth = tma(njerk, window * fs)
+  )
 ```
@@
 ```{r align-ecg}
-acc_400hz$ecg_beat <- FALSE
-beat_idx <- approx(acc_400hz$dt, 
-                   seq_along(acc_400hz$dt), 
+# Align ECG-detected heart beats
+oo_400hz$ecg_beat <- FALSE
+beat_idx <- approx(oo_400hz$dt, 
+                   seq_along(oo_400hz$dt), 
                    xout = ecg_beats$dt,
@@
   na.omit()
-acc_400hz$ecg_beat[beat_idx] = TRUE
-
-prh_10hz$ecg_beat <- FALSE
-beat_idx <- approx(prh_10hz$dt, 
-                   seq_along(prh_10hz$dt), 
-                   xout = ecg_beats$dt,
-                   method = "constant")$y %>% 
-  na.omit()
-prh_10hz$ecg_beat[beat_idx] = TRUE
+oo_400hz$ecg_beat[beat_idx] = TRUE
+oo_400hz$ecg_bpm <- with(oo_400hz, bpm(ecg_beat, dt))
 ```
 
-```{r bcg-beats}
-acc_400hz$surge_range <- local_range(acc_400hz$surge_filt, window * fs)
-acc_400hz$bcg_beat <- find_beats_lsr(acc_400hz$surge_filt, window, fs)
-acc_400hz$ecg_bpm[acc_400hz$ecg_beat] <- with(
-  filter(acc_400hz, ecg_beat), 
-  1 / as.numeric(lead(dt) - dt, unit = "mins")
-)
-acc_400hz$bcg_bpm[acc_400hz$bcg_beat] <- with(
-  filter(acc_400hz, bcg_beat), 
-  1 / as.numeric(lead(dt) - dt, unit = "mins")
-)
-
-prh_10hz$bcg_beat <- FALSE
-bcg_idx <- approx(prh_10hz$dt, 
-                  seq_along(prh_10hz$bcg_beat), 
-                  acc_400hz$dt[acc_400hz$bcg_beat],
-                  "constant")$y
-prh_10hz$bcg_beat[bcg_idx] <- TRUE
-```
-
-```{r bcg-figs}
+```{r oo-sample}
 ca_tz <- "Etc/GMT+7"
@@
 bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
-bcg_sample <- filter(acc_400hz, between(dt, bcg_begin, bcg_end))
-bcg_sample_10hz <- filter(prh_10hz, between(dt, bcg_begin, bcg_end))
+bcg_sample <- filter(oo_400hz, between(dt, bcg_begin, bcg_end))
 
@@
 
-```{r corky-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
+```{r oo-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
 # A: Surge + LSR
@@
   geom_line(size = 0.2) +
-  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat)) +
+  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), size = 0.2) +
   geom_point(color = "blue") +
@@
 
-```{r corky-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
+```{r ecg_bcg_comparison}
+ecg_bpm <- filter(bcg_sample, ecg_beat)$ecg_bpm %>% tail(-1)
+bcg_bpm <- filter(bcg_sample, bcg_beat)$bcg_bpm %>% head(-1)
+bpm_diff <- abs((bcg_bpm - ecg_bpm) / ecg_bpm * 100)
+sprintf("%0.1f%% \u00B1 %0.1f%% (mean \u00B1 sd))", mean(bpm_diff), sd(bpm_diff))
+```
+
+```{r oo-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
 Solid line: surge bandpass filtered to [1.0 25.0] Hz.
@@
 C: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
-# A: Surge + LSR
-# pA as above
+# A: Surge
+pA <- ggplot(bcg_sample, aes(dt, surge_filt)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = bquote(Surge~(m~s^{-2}))) +
+  expand_limits(x = max(bcg_sample$dt) + 1) +
+  t
 
@@
 pB <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
+  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
   geom_line(size = 0.2) +
-  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
   scale_x_datetime(date_labels = "%H:%M:%S") +
@@
 # C: Jerk
-pC <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
+pC <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
   geom_line(size = 0.2) +
-  geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
+  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
   scale_x_datetime(date_labels = "%H:%M:%S") +
   labs(y = bquote(Jerk~(m~s^{-3}))) +
-  expand_limits(y = 0) +
   t
@@
                    nrow = 3, 
-                   # rel_heights = c(1, 1, ),
                    labels = "AUTO")
@@
 
-```{r bw-data}
-fs <- 400
-fny <- fs / 2
-acc_filter <- signal::butter(4, c(1.5, 7.5) / fny, type = "pass") 
-apply_filter <- function(x) signal::filtfilt(acc_filter, x)
-window <- 1.8 #s
-
+```{r bw-beats}
+window = 1.8 # s
 bw_elg <- bw180905_53_elg
 
+# Apply filters as with killer whale data, but narrower band
 bw_400hz <- bw180905_53_400hz %>% 
   mutate(
-    across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"),
+    # Pass-band filter acceleration
+    across(surge:heave, 
+           bandpass_filter, pass = c(1.0, 10.0), 
+           .names = "{.col}_filt"),
+    # Calculate jerk norm and apply smoother
+    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
+                  p = 4, n = window * fs),
+    njerk_smooth = tma(njerk, window * fs),
+    # Annotate regions
     rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
     rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
-    region_id = ifelse(rid_left == rid_right, rid_left, NA)
+    region_id = ifelse(rid_left == rid_right, rid_left, NA),
+    # Find heart beats using jerk cue
+    surge_range = local_range(surge_filt, window * fs),
+    bcg_beat = find_beats_jerk(njerk_smooth, !is.na(region_id), window, fs)
   ) %>% 
+  # Calculate heart rate within each region
+  group_by(region_id) %>% 
+  mutate(bcg_bpm = bpm(bcg_beat, dt)) %>% 
+  ungroup() %>% 
   select(-c(rid_left, rid_right))
-
-bw_10hz <- bw180905_53_10hz %>% 
-  mutate(jerk = njerk(aw, 10),
-         jerk_smooth = tma(jerk, window * 10))
-```
-
-```{r bw-beats}
-bw_400hz_beats <- bw_400hz %>% 
-  mutate(jerk = approx(bw_10hz$dt, bw_10hz$jerk_smooth, bw_400hz$dt, rule = 2)$y) %>% 
-  filter(!is.na(region_id)) %>%
-  group_by(region_id) %>% 
-  mutate(bcg_beat = find_beats_jerk(jerk, surge_filt, window, fs),
-         bcg_bpm = bpm(bcg_beat, dt)) %>% 
-  ungroup() %>% 
-  select(dt, bcg_beat, bcg_bpm)
-
-bw_400hz <- bw_400hz %>% 
-  mutate(surge_range = local_range(surge_filt, window * fs)) %>% 
-  left_join(bw_400hz_beats, by = "dt")
-
-bw_10hz$bcg_beat <- FALSE
-bcg_idx <- approx(bw_10hz$dt, 
-                  seq_along(bw_10hz$bcg_beat), 
-                  bw_400hz$dt[bw_400hz$bcg_beat],
-                  "constant")$y
-bw_10hz$bcg_beat[bcg_idx] <- TRUE
 ```
@@
   bcg_sample <- filter(bw_400hz, region_id == rid)
-  bcg_sample_10hz <- bw_10hz %>% 
-    filter(between(dt, bcg_sample$dt[1], bcg_sample$dt[nrow(bcg_sample)]))
 
@@
 
-  p1 <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
+  pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
+    # geom_vline(aes(xintercept = dt),
+    #            data = filter(bcg_surge_lsr, bcg_beat), 
+    #            color = "blue", alpha = 0.5) +
     geom_line(size = 0.2) +
@@
 
-  p2 <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
+  pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
     geom_line() +
@@
 
-  p3 <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
+  pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
+    # geom_vline(aes(xintercept = dt),
+    #            filter(bcg_sway_heave, bcg_beat), 
+    #            color = "blue", alpha = 0.5) +
     geom_line(size = 0.2) +
@@
 
-  p4 <- ggplot(bcg_sample_10hz, aes(dt, jerk_smooth)) +
+  pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
     geom_line(size = 0.2) +
-    geom_point(data = filter(bcg_sample_10hz, bcg_beat), color = "blue") +
+    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
     scale_x_datetime(date_labels = "%H:%M:%S") +
@@
 
-  p <- cowplot::plot_grid(p1, p2, p3, p4, 
+  p <- cowplot::plot_grid(pA, pB, pC, pD,
                           align = "v", 
@@
 
-Figure \@ref(fig:corky-bcg-ecg) shows how we can have a caption and cross-reference for a plot.
+Figure \@ref(fig:oo-bcg-ecg) shows how we can have a caption and cross-reference for a plot.
 

### 2021-09-05 analysis/paper/paper.Rmd
Add bradycardia analysis and figure for classifying jerk peaks

@@
   pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
-    # geom_vline(aes(xintercept = dt),
-    #            data = filter(bcg_surge_lsr, bcg_beat), 
-    #            color = "blue", alpha = 0.5) +
     geom_line(size = 0.2) +
@@
   # B: BPM
-
   pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
@@
   # C: Sway, heave
-
   bcg_sway_heave <- bcg_sample %>% 
@@
   pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
-    # geom_vline(aes(xintercept = dt),
-    #            filter(bcg_sway_heave, bcg_beat), 
-    #            color = "blue", alpha = 0.5) +
     geom_line(size = 0.2) +
@@
   # D: Jerk
-
   pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
@@
 # foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
+```
+
+```{r bradycardia-relax}
+bw10hz <- bw180905_53_10hz %>%
+  mutate(
+    dive_id = split_dives(dt, depth,
+                          surface = 2, min_depth = 10, min_dur = 5 * 60),
+    dive_norm = normalize_dives(dive_id)
+  )
+bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y
+
+bpm_data <- bw_400hz %>%
+  drop_na(bcg_bpm)
+
+bpm_ts <- RobustLinearReg::theil_sen_regression(bcg_bpm ~ dive_norm, bpm_data)
+
+bpm_pred <- tibble(dive_norm = seq(0, 1, length.out = 100)) %>%
+  mutate(bcg_bpm = predict(bpm_ts, newdata = .))
+
+ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
+  geom_point() +
+  geom_line(data = bpm_pred, color = "blue", size = 1.5) +
+  scale_x_continuous(labels = scales::percent) +
+  labs(x = "Dive time (normalized)",
+       y = "Heartrate (bpm)") +
+  expand_limits(y = 0) +
+  theme_classic()
+```
+
+The magnitude of bradycardia was strongest at the start of dives and relaxed towards the end.
+On average, heartrate increased from `round(predict(bpm_ts, data.frame(dive_norm = 0)), 1)` bpm at the start of dives to `round(predict(bpm_ts, data.frame(dive_norm = 1)), 1)` bpm by the end (Theil-Sen regression, $p < 10^{-10}$).
+
+```{r jerk-peaks, fig.cap="Major peak selection procedure for identifying heartbeats.
+A Even after smoothing, the norm of the jerk vector has peaks that do not correspond to heartbeats (e.g.
+2, 4).
+B In height-prominence space, there are two clear clusters corresponding to valid heartbeats (upper right) and artifacts (lower left).
+C "}
+# Visualize peak thresholding algorithm
+jerk <- bw_400hz$njerk_smooth
+jerk[is.na(bw_400hz$region_id)] <- 0
+window_s <- 1.8
+fs_hz <- 400
+
+# Find peaks
+peaks <- pracma::findpeaks(jerk, minpeakdistance = window * fs_hz)
+jerk_peaks <- slice(bw_400hz, peaks[, 2]) %>% 
+  mutate(
+    height = peaks[, 1],
+    prominence = peak_prominences(jerk, peaks[, 2]),
+    distance = sqrt((max(height) - height)^2 + (max(prominence) - prominence)^2)
+  ) %>% 
+  arrange(dt)
+
+# Using region 14 as an example
+jerk_example <- filter(bw_400hz, region_id == 14)
+peak_example <- jerk_peaks %>% 
+  filter(between(dt, first(jerk_example$dt), last(jerk_example$dt))) %>% 
+  mutate(i = row_number())
+
+# Labeled peaks in jerk
+pA <- ggplot(jerk_example, aes(dt, njerk_smooth)) +
+  geom_line() +
+  geom_point(data = peak_example) +
+  ggrepel::geom_text_repel(aes(label = i), peak_example) +
+  expand_limits(y = 0) +
+  labs(y = bquote(Jerk~(m~s^{-3}))) +
+  theme_classic()
+# Peak height, prominence
+pB <- ggplot(peak_example, aes(height, prominence)) +
+  geom_point() +
+  ggrepel::geom_text_repel(aes(label = i), max.overlaps = Inf) +
+  coord_fixed() +
+  theme_classic()
+
+# All heights, prominences (with threshold)
+
+height_prom <- jerk_peaks %>% select(height, prominence)
+peak_pca <- prcomp(height_prom, scale = FALSE)
+pc1 <- peak_pca$x[, "PC1"]
+pc1_density <- density(pc1)
+pc1_peaks <- pracma::findpeaks(pc1_density$y)
+pc1_valley <- pracma::findpeaks(-pc1_density$y[pc1_peaks[1, 2]:pc1_peaks[2, 2]])
+pc1_thr <- pc1_density$x[pc1_valley[1, 2] + pc1_peaks[1, 2] - 1]
+pc1_axis <- c(peak_pca$sdev[1], 0) %*% t(peak_pca$rotation) %>% 
+  scale(center = -colMeans(height_prom), scale = FALSE)
+pc2_axis <- c(0, peak_pca$sdev[2]) %*% t(peak_pca$rotation) %>% 
+  scale(center = -colMeans(height_prom), scale = FALSE)
+pc1_density_rotate <- cbind(pc1_density$x, pc1_density$y) %*% 
+  t(peak_pca$rotation) %>% 
+  scale(center = -colMeans(height_prom), scale = FALSE)
+pc1_thr_rot <- c(pc1_thr, 0) %*% t(peak_pca$rotation) %>% 
+  scale(center = -colMeans(height_prom), scale = FALSE)
+slope <- (pc2_axis - peak_pca$center)[2] / (pc2_axis - peak_pca$center)[1]
+intercept <- pc1_thr_rot[2] - slope * pc1_thr_rot[1]
+pc1_angle <- atan2(peak_pca$rotation[2, 1], peak_pca$rotation[1, 1]) * 180 / pi
+pc2_angle <- pc1_angle - 90
+
+pC <- ggplot(jerk_peaks, aes(height, prominence)) +
+  geom_point() +
+  annotate("segment", 
+           x = peak_pca$center[1], y = peak_pca$center[2], 
+           xend = pc1_axis[1], yend = pc1_axis[2],
+           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
+  annotate("text", label = "PC1",
+           x = pc1_axis[1], y = pc1_axis[2],
+           hjust = -0.1, angle = pc1_angle,
+           color = "red") + 
+  annotate("segment", 
+           x = peak_pca$center[1], y = peak_pca$center[2], 
+           xend = pc2_axis[1], yend = pc2_axis[2],
+           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
+  annotate("text", label = "PC2",
+           x = pc2_axis[1], y = pc2_axis[2],
+           hjust = 1.1, angle = pc2_angle,
+           color = "red") + 
+  geom_abline(slope = slope, intercept = intercept, 
+              color = "red", linetype = 2) +
+  coord_fixed() +
+  theme_classic()
+
+# density along pc1
+pD <- as_tibble(peak_pca$x) %>% 
+  ggplot(aes(PC1)) +
+  geom_histogram(aes(y = ..density..), bins = 50) +
+  geom_density(color = "blue") +
+  geom_vline(xintercept = pc1_thr, color = "red", linetype = 2) +
+  theme_classic()
+
+cowplot::plot_grid(pA, pB, pC, pD,
+                   nrow = 2,
+                   labels = "AUTO")
 ```

### 2021-09-05 R/dives.R

@@ -1,0 +1,44 @@
+#' Split dive profile into dives
+#'
+#' @param dt Datetime (POSIXct)
+#' @param depth Depth (numeric)
+#' @param surface Threshold for surface activity (numeric(1))
+#' @param min_depth Minimum depth for a dive (numeric(1))
+#' @param min_dur Minimum duration of a dive (numeric(1), in seconds)
+#'
+#' @return A factor, the same length as \code{depth}, with levels corresponding
+#'   to individual dives. Inter-dive periods marked as NA.
+#' @export
+split_dives <- function(dt, depth, surface = 0, min_depth = 0, min_dur = 0) {
+  dive_start <- depth >= surface & lag(depth) < surface
+  dive_start[1] <- FALSE
+  dive_id <- cumsum(dive_start)
+  dive_id[depth < surface | dive_id == 0] <- NA
+  dive_id <- factor(dive_id)
+  dive_depth <- tapply(depth, dive_id, max)
+  dive_dur <- tapply(dt,
+                     dive_id,
+                     function(x) as.numeric(max(x) - min(x), unit = "secs"))
+  not_dives <- which(dive_depth < min_depth | dive_dur < min_dur)
+  dive_id[dive_id %in% not_dives] <- NA
+  dive_id <- droplevels(dive_id)
+  levels(dive_id) <- seq_along(levels(dive_id))
+  dive_id
+}
+
+#' Normalize dive time
+#'
+#' Normalize the time in each dive to [0, 1] where 0 is start of dive and 1 is
+#' the end.
+#'
+#' @param dive_id A factor of dive id's, as from \code{split_dives()}.
+#'
+#' @return A numeric vector, the same length as \code{dive_id}.
+#' @export
+normalize_dives <- function(dive_id) {
+  dive_rle <- rle(as.character(dive_id))
+  mapply(function(l, v) if (!is.na(v)) seq(0, 1, length.out = l) else NA,
+         dive_rle$lengths,
+         dive_rle$values) %>%
+    unlist()
+}

### 2021-09-05 analysis/scratchpad/divesegmentation.R

@@ -1,0 +1,81 @@
+split_dives <- function(dt, depth, surface = 0, min_depth = 0, min_dur = 0) {
+  dive_start <- depth >= surface & lag(depth) < surface
+  dive_start[1] <- FALSE
+  dive_id <- cumsum(dive_start)
+  dive_id[depth < surface | dive_id == 0] <- NA
+  dive_id <- factor(dive_id)
+  dive_depth <- tapply(depth, dive_id, max)
+  dive_dur <- tapply(dt,
+                     dive_id,
+                     function(x) as.numeric(max(x) - min(x), unit = "secs"))
+  not_dives <- which(dive_depth < min_depth | dive_dur < min_dur)
+  dive_id[dive_id %in% not_dives] <- NA
+  dive_id <- droplevels(dive_id)
+  levels(dive_id) <- seq_along(levels(dive_id))
+  dive_id
+}
+
+segment_phases <- function(depth, dive_id) {
+  by_dive <- split(depth, dive_id) %>%
+    lapply(function(x) {
+      i <- seq_along(x)
+      depth_lm <- lm(x ~ i)
+      depth_seg <- segmented::segmented(depth_lm, seg.Z = ~i, npsi = 2)
+      psi <- floor(depth_seg$psi[, "Est."])
+      result <- character(length(x))
+      result[1:psi[1]] <- "descent"
+      result[psi[1]:psi[2]] <- "bottom"
+      result[psi[2]:length(x)] <- "ascent"
+      result
+    })
+  phases_all <- rep(NA, length(dive_id))
+  for (dive in names(by_dive)) {
+    phases_all[!is.na(dive_id) & dive_id == dive] <- by_dive[[dive]]
+  }
+  factor(phases_all, levels = c("descent", "bottom", "ascent"))
+}
+
+normalize_dive <- function(dive_phase, dive_id) {
+  by_dive <-split(dive_phase, dive_id) %>%
+    lapply(function(x) {
+      result <- rep(NA, length(x))
+      b <- which(x == "bottom")
+      result[x == "bottom"] <- (b - min(b)) / (max(b) - min(b))
+      result
+    })
+  all_bottom <- rep(NA, length(dive_id))
+  for (dive in names(by_dive)) {
+    all_bottom[!is.na(dive_id) & dive_id == dive] <- by_dive[[dive]]
+  }
[…]

### 2021-09-09 analysis/paper/paper.Rmd
Use distance threshold for jerk peak classification

--- 2021-09-03 analysis/paper/paper.Rmd
@@
 # foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
+```
+
+```{r bradycardia-relax, fig.cap="Heartrates observed in the ballistocardiogram exhibit characteristic diving physiology patterns.
+Bradycardia is greatest at the start of the dive (~4-5 bpm), increasing towards the end (~8-9 bpm)."}
+bw10hz <- bw180905_53_10hz %>%
+  mutate(
+    dive_id = split_dives(dt, depth,
+                          surface = 2, min_depth = 10, min_dur = 5 * 60),
+    dive_norm = normalize_dives(dive_id)
+  )
+bw_400hz$dive_norm <- approx(bw10hz$dt, bw10hz$dive_norm, bw_400hz$dt)$y
+
+bpm_data <- bw_400hz %>%
+  drop_na(bcg_bpm)
+
+bpm_ts <- RobustLinearReg::theil_sen_regression(bcg_bpm ~ dive_norm, bpm_data)
+
+bpm_pred <- tibble(dive_norm = seq(0, 1, length.out = 100)) %>%
+  mutate(bcg_bpm = predict(bpm_ts, newdata = .))
+
+ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
+  geom_point() +
+  geom_line(data = bpm_pred, color = "blue", size = 1.5) +
+  scale_x_continuous(labels = scales::percent) +
+  labs(x = "Dive time (normalized)",
+       y = "Heartrate (bpm)") +
+  expand_limits(y = 0) +
+  theme_classic()
+```
+
+The magnitude of bradycardia was strongest at the start of dives and relaxed towards the end.
+On average, heartrate increased from `r round(predict(bpm_ts, data.frame(dive_norm = 0)), 1)` bpm at the start of dives to `r round(predict(bpm_ts, data.frame(dive_norm = 1)), 1)` bpm by the end (Theil-Sen regression, $p < 10^{-10}$).
+
+```{r jerk-peaks, fig.cap="Major peak selection procedure for identifying heartbeats.
+A Even after smoothing, the norm of the jerk vector (line) has peaks that correspond to heartbeats (solid points) and artifacts (hollow points).
+We used a data-driven approach to retaining only major peaks.
+75 s data excerpt shown as example.
+B In height-prominence space, peaks form two clusters corresponding to valid heartbeats (solid) and artifacts (hollow).
+C To separate the clusters, we calculated each peak's distance from the greatest peak (triangle); retaining all peaks closer to the greatest peak (solid circles) than the valley in the bimodal distance distribution (red dashed line) and rejecting the farther peaks (hollow circles).
+D The bi-modal distribution of peak distances.
+We selected the the valley in the distance density (blue line) for the distance threshold (red dashed line).
+Solid and hollow histogram bars correspond to retained and rejected peaks, respectively."}
+# Visualize peak thresholding algorithm
+jerk <- bw_400hz$njerk_smooth
+jerk[is.na(bw_400hz$region_id)] <- 0
+window_s <- 1.8
+fs_hz <- 400
+
+# Find peaks
+peaks <- pracma::findpeaks(jerk, minpeakdistance = window * fs_hz)
+jerk_peaks <- slice(bw_400hz, peaks[, 2]) %>% 
+  mutate(
+    height = peaks[, 1],
+    prominence = peak_prominences(jerk, peaks[, 2]),
+    distance = sqrt((max(height) - height)^2 + (max(prominence) - prominence)^2)
+  ) %>% 
+  arrange(dt)
+dist_density <- density(jerk_peaks$distance)
+dist_peaks <- pracma::findpeaks(dist_density$y)
+interpeaks <- dist_peaks[1, 2]:dist_peaks[2, 2]
+dist_valley <- pracma::findpeaks(-dist_density$y[interpeaks])
+dist_thr <- dist_density$x[dist_valley[1, 2] + dist_peaks[1, 2] - 1]
+jerk_peaks <- jerk_peaks %>%  
+  mutate(distance_cat = cut(distance, 
+                            c(-Inf, 
+                              min(distance[distance > 0]) / 2, 
+                              dist_thr, 
+                              Inf),
+                            labels = c("max",
+                                       "keep",
+                                       "reject")))
+shape_scale <- scale_shape_manual(values = c(max = 6, keep = 16, reject = 1))
+hp_labs <- labs(x = "Peak height", y = "Peak prominence")
+
+# Using region 14 as an example
+jerk_example <- filter(bw_400hz, region_id == 14)
+peak_example <- jerk_peaks %>% 
+  filter(between(dt, first(jerk_example$dt), last(jerk_example$dt))) %>% 
+  mutate(i = row_number())
+
+# Labeled peaks in jerk
+pA <- ggplot(jerk_example, aes(dt, njerk_smooth)) +
+  geom_line() +
+  geom_point(aes(shape = distance_cat), peak_example) +
+  ggrepel::geom_text_repel(aes(label = i), peak_example) +
+  shape_scale +
+  expand_limits(y = 0) +
+  labs(y = bquote(Jerk~(m~s^{-3}))) +
+  theme_classic() +
+  theme(axis.title.x = element_blank(),
+        legend.position = "none")
+# Peak height, prominence
+pB <- ggplot(peak_example, aes(height, prominence, shape = distance_cat)) +
+  geom_point() +
+  ggrepel::geom_text_repel(aes(label = i), max.overlaps = Inf) +
+  shape_scale +
+  hp_labs +
+  coord_fixed() +
+  theme_classic() +
+  theme(legend.position = "none")
+
+# All heights, prominences (with threshold)
+x0 <- max(jerk_peaks$height)
+y0 <- max(jerk_peaks$prominence)
+r <- dist_thr
+theta <- seq(0, 2 * pi, length.out = 256)
+threshold <- tibble(
+  height = x0 + r * cos(theta),
+  prominence = y0 + r * sin(theta)
+) %>% 
+  filter(height <= x0,
+         prominence <= y0)
+
+pC <- jerk_peaks %>% 
+    mutate() %>% 
+  ggplot(aes(height, prominence)) +
+  geom_point(aes(shape = distance_cat, size = distance_cat)) +
+  annotate("path", 
+           x = threshold$height, y = threshold$prominence,
+           color = "red", linetype = "dashed") +
+  shape_scale +
+  scale_size_manual(values = c(max = 3, keep = 1, reject = 1)) +
+  hp_labs +
+  coord_fixed() +
+  theme_classic() +
+  theme(legend.position = "none")
+
+pD <- ggplot(jerk_peaks, aes(distance)) +
+  geom_histogram(aes(y = ..density.., fill = distance < dist_thr), 
+                 bins = 25, color = "black") +
+  geom_density(color = "blue") +
+  geom_vline(xintercept = dist_thr, color = "red", linetype = 2) +
+  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white")) +
+  labs(x = "Distance from max") +
+  theme_classic() +
+  theme(legend.position = "none")
+
+cowplot::plot_grid(pA, pB, pC, pD,
+                   align = "h",
+                   axis = "bt",
+                   nrow = 2,
+                   labels = "AUTO")
 ```

### 2021-09-09 R/find_beats.R

@@ -77,15 +77,18 @@ find_beats_jerk <- function(jerk, mask, window_s = 1.8, fs_hz = 400) {
   jerk_prm <- peak_prominences(jerk, jerk_pks)
 
   # Find major peaks
+  # Assuming a bimodal distance-to-max distribution, set threshold to valley
+  # in density
+  height_max <- max(jerk_hts)
+  prom_max <- max(jerk_prm)
+  distance <- sqrt((jerk_hts - height_max)^2 + (jerk_prm - prom_max)^2)
+  dist_density <- density(distance)
+  dist_peaks <- pracma::findpeaks(dist_density$y)
+  interpeaks <- dist_peaks[1, 2]:dist_peaks[2, 2]
+  dist_valley <- pracma::findpeaks(-dist_density$y[interpeaks])
+  dist_thr <- dist_density$x[dist_valley[1, 2] + dist_peaks[1, 2] - 1]
 
-  # Distance from largest peak in height/prominence space
-  dist <- sqrt((max(jerk_hts) - jerk_hts)^2 + (max(jerk_prm) - jerk_prm)^2)
-  # Keep all peaks up to knee in dist
-  dist2 <- signal::sgolayfilt(sort(dist), p = 2, n = 25, m = 2)
-  knee <- which.max(dist2)
-  dist_thr <- sort(dist)[knee]
-  major_peaks <- jerk_pks[dist < dist_thr]
-
+  major_peaks <- jerk_pks[distance <= dist_thr]
   beats <- logical(length(jerk))
   beats[major_peaks] <- TRUE
   beats

### 2021-09-09 analysis/scratchpad/psd.R
Calculate BCG signal-to-noise ratio using power spectral density

@@ -1,0 +1,121 @@
+# Signal band within these bounds
+bw_bpm <- c(4, 8) / 60
+
+bcg_psd <- function(low, width, bw400hz,
+                    fs = 400, window = 1.8, bpm = c(4, 8)) {
+  pass <- c(low, low + width)
+  fny <- fs / 2
+  bandpass_filter <- function(x) {
+    signal::butter(4, pass / fny, type = "pass") %>%
+      signal::filtfilt(x)
+  }
+  acc <- as.matrix(bw400hz[, c("surge", "heave", "sway")])
+  acc_filt <- apply(acc, 2, bandpass_filter)
+
+  if (sum(is.finite(acc_filt)) == 0) {
+    return(tibble())
+  }
+
+  surge_filt <- acc[, 1]
+  lsr <- local_range(surge_filt, window * fs)
+  j = njerk(acc_filt, p = 4, n = window * fs)
+  njerk_smooth = tma(j, window * fs)
+
+  # PSD of surge, LSR, and jerk
+  psds <- list(surge_filt, lsr, njerk_smooth) %>%
+    map(~ ts(.x[!is.na(bw400hz$region_id)], frequency = 400)) %>%
+    map(~ suppressMessages(psd::pspectrum(.x)))
+
+  freq <- psds[[1]]$freq
+
+  tibble(freq,
+         surge = psds[[1]]$spec,
+         lsr = psds[[2]]$spec,
+         jerk = psds[[3]]$spec) %>%
+    pivot_longer(surge:jerk,
+                 names_to = "norm_fun",
+                 values_to = "spec")
+}
+
+# Latin hypercube sampling
+low <- 2^seq(-2, 2, length.out = 256)
+width <- 2^seq(1, 5, length.out = 256)
+bcg_lhs <- tibble(
+  low = sample(low, length(low), replace = FALSE),
+  width = sample(width, length(width), replace = FALSE)
+)
+pb <- progress::progress_bar$new(total = nrow(bcg_lhs))
+bcg_psd_results <- bcg_lhs %>%
+  group_by_all() %>%
+  group_modify(function(x, y) {
+    pb$tick()
+    bcg_psd(y$low, y$width, bw_400hz)
+  }) %>%
+  ungroup()
+
+calc_s2n <- function(freq, spec) {
+  is_signal <- between(freq, bw_bpm[1], bw_bpm[2])
+  is_noise <- !is_signal
+
[…]

### 2021-10-20 analysis/paper/paper.Rmd
Use power spectral density to quantify signal:noise
Former-commit-id: c7c773281971e09c5e41b25c3027a993ef596d4e

@@
 ---
-title: "Something about cetacean heart rates and ballistocardiograms"
+title: "Accelerometer-derived ballistocardiogram confirms blue whale (*Balaenoptera musculus*) heart rate during diving bradycardia"
 author:
@@
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
 ---
@@
 
-# Background
+# Materials and methods
 
-# Methods
-
-# Results
+# Results and discussion
 
@@
 ```{r bw-beats}
-window = 1.8 # s
+window <- 1.8 # s
 bw_elg <- bw180905_53_elg
@@
 
-# foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
+foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
 ```
@@
 
-# Discussion
-
-# Conclusion
-
 # Acknowledgements

### 2021-10-20 analysis/scratchpad/bcg_s2n.R

@@ -1,0 +1,110 @@
+library(cetaceanbcg)
+library(tidyverse)
+
+bcg_psd_results <- readRDS("analysis/data/derived_data/bcg_psd_results.rds")
+n <- as.integer(readLines("analysis/data/derived_data/psd_par_n.txt"))
+
+# "Signal" band (typical blue whale diving heartrates)
+bw_bpm <- c(4, 8) / 60
+
+# Calculate signal-to-noise ratio by integrating the PSD curve
+calc_s2n <- function(freq, spec) {
+  is_signal <- between(freq, bw_bpm[1], bw_bpm[2])
+  is_noise <- !is_signal
+
+  psignal = pracma::trapz(freq[is_signal], spec[is_signal])
+  pnoise = pracma::trapz(freq[is_noise], spec[is_noise])
+
+  psignal / pnoise
+}
+
+# Ranked signal-to-noise result by bandpass parameters
+bcg_s2n <- bcg_psd_results %>%
+  drop_na(norm_fun) %>%
+  group_by(low, width, norm_fun) %>%
+  summarize(s2n = calc_s2n(freq, spec), .groups = "drop") %>%
+  arrange(-s2n)
+
+# Parameter space, covered by sampling. Use interpolation to fill in gaps.
+low <- 2^seq(-2, 2, length.out = n)
+width <- 2^seq(1, 5, length.out = n)
+interp_s2n <- akima::interp(
+  x = log2(bcg_s2n$low[bcg_s2n$norm_fun == "jerk"]),
+  y = log2(bcg_s2n$width[bcg_s2n$norm_fun == "jerk"]),
+  z = bcg_s2n$s2n[bcg_s2n$norm_fun == "jerk"],
+  xo = log2(low),
+  yo = log2(width),
+  linear = FALSE,
+  extrap = FALSE
+)
+bcg_s2n_interp <- expand_grid(low, width) %>%
+  mutate(i = match(log2(low), interp_s2n$x),
+         j = match(log2(width), interp_s2n$y),
+         s2n = interp_s2n$z[cbind(i, j)]) %>%
+  select(-i, -j)
+
+interp_s2n <- akima::interp(
+  x = bcg_s2n$low[bcg_s2n$norm_fun == "jerk"],
+  y = bcg_s2n$width[bcg_s2n$norm_fun == "jerk"],
+  z = bcg_s2n$s2n[bcg_s2n$norm_fun == "jerk"],
+  xo = low,
[…]

### 2021-10-20 analysis/scratchpad/create_bw_400hz.R

@@ -1,0 +1,39 @@
+library(cetaceanbcg)
+library(tidyverse)
+
+fs <- 400
+fny <- fs / 2
+window <- 1.8 # s
+bw_elg <- bw180905_53_elg
+
+bandpass_filter <- function(x, pass) {
+  signal::butter(4, pass / fny, type = "pass") %>%
+    signal::filtfilt(x)
+}
+
+# Apply filters as with killer whale data, but narrower band
+bw_400hz <- bw180905_53_400hz %>%
+  mutate(
+    # Pass-band filter acceleration
+    across(surge:heave,
+           bandpass_filter, pass = c(1.0, 10.0),
+           .names = "{.col}_filt"),
+    # Calculate jerk norm and apply smoother
+    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt),
+                  p = 4, n = window * fs),
+    njerk_smooth = tma(njerk, window * fs),
+    # Annotate regions
+    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
+    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
+    region_id = ifelse(rid_left == rid_right, rid_left, NA),
+    # Find heart beats using jerk cue
+    surge_range = local_range(surge_filt, window * fs),
+    bcg_beat = find_beats_jerk(njerk_smooth, !is.na(region_id), window, fs)
+  ) %>%
+  # Calculate heart rate within each region
+  group_by(region_id) %>%
+  mutate(bcg_bpm = bpm(bcg_beat, dt)) %>%
+  ungroup() %>%
+  select(-c(rid_left, rid_right))
+
+saveRDS(bw_400hz, "analysis/data/derived_data/bw_400hz.rds")

### 2021-10-20 analysis/scratchpad/psd.R

@@ -1,3 +1,5 @@
+
+
 # Signal band within these bounds
 bw_bpm <- c(4, 8) / 60
 
@@ -63,7 +65,7 @@ calc_s2n <- function(freq, spec) {
   psignal / pnoise
 }
 
-plot_bcg_psd <- function(dat, key) {
+plot_bcg_psd <- function(dat, key, outdir) {
   is_signal <- between(dat$freq, bw_bpm[1], bw_bpm[2])
   is_noise <- !is_signal
 

### 2021-10-20 analysis/scratchpad/psd_par.R

@@ -1,0 +1,90 @@
+library(cetaceanbcg)
+library(doParallel)
+library(tidyverse)
+
+n <- as.integer(commandArgs(trailingOnly = TRUE))
+
+# use the environment variable SLURM_NTASKS_PER_NODE to set
+# the number of cores to use
+if (Sys.getenv("SLURM_NTASKS_PER_NODE") == "") {
+  ncores <- 1
+} else {
+  ncores <- Sys.getenv("SLURM_NTASKS_PER_NODE")
+}
+registerDoParallel(cores = ncores)
+
+bw_400hz <- readRDS("analysis/data/derived_data/bw_400hz.rds")
+
+# Signal band within these bounds
+bw_bpm <- c(4, 8) / 60
+
+bcg_psd <- function(low, width, bw400hz,
+                    fs = 400, window = 1.8, bpm = c(4, 8), max_bpm = 64) {
+  tryCatch({
+    pass <- c(low, low + width)
+    fny <- fs / 2
+    bandpass_filter <- function(x) {
+      signal::butter(4, pass / fny, type = "pass") %>%
+        signal::filtfilt(x)
+    }
+    acc <- as.matrix(bw400hz[, c("surge", "heave", "sway")])
+    acc_filt <- apply(acc, 2, bandpass_filter)
+
+    if (sum(is.finite(acc_filt)) == 0) {
+      return(tibble())
+    }
+
+    surge_filt <- acc[, 1]
+    lsr <- local_range(surge_filt, window * fs)
+    j = njerk(acc_filt, p = 4, n = window * fs)
+    njerk_smooth = tma(j, window * fs)
+
+    # PSD of surge, LSR, and jerk
+    psds <- list(surge_filt, lsr, njerk_smooth) %>%
+      map(~ ts(.x[!is.na(bw400hz$region_id)], frequency = 400)) %>%
+      map(~ suppressMessages(psd::pspectrum(.x)))
+
+    freq <- psds[[1]]$freq
+
+    tibble(freq,
+           surge = psds[[1]]$spec,
+           lsr = psds[[2]]$spec,
+           jerk = psds[[3]]$spec) %>%
+      pivot_longer(surge:jerk,
+                   names_to = "norm_fun",
+                   values_to = "spec") %>%
+      filter(freq * 60 <= max_bpm)
+  }, error = function(e) {
+    cat(sprintf("low = %0.3f; width = %0.3f; error = \"%s\"",
[…]

### 2021-10-21 R/find_beats.R
Staging find_beats.R before switching to the approach described in Lee et al. 2016 Sensors
Former-commit-id: f595ccc5aced013905354673bf636755598b5ca9

@@ -16,6 +16,14 @@ find_beats_lsr <- function(surge, window_s = 0.25, fs_hz = 400) {
   local_range_max <- RcppRoll::roll_max(surge_range, window, fill = NA)
   range_peaks <- surge_range == local_range_max
 
+  # Keep only most prominent peaks by looking for a break in the distribution
+  range_peak_idx <- which(range_peaks)
+  prominence <- peak_prominences(surge_range, range_peak_idx)
+  prominence_asc <- sort(prominence)
+  prominence_steps <- diff(prominence_asc)
+  prominence_thr <- prominence_asc[which.max(prominence_steps)]
+  range_peaks[range_peak_idx[prominence < prominence_thr]] <- FALSE
+
   # Align range peaks with surge peaks
   beats <- logical(length(range_peaks))
   i <- 1
@@ -57,17 +65,32 @@ tma <- function(x, k) {
     RcppRoll::roll_mean(k, fill = NA)
 }
 
+#' Shannon entropy
+#'
+#' Defined as -|x| * log(|x|)
+#'
+#' @param x `[numeric]`
+#'
+#' @return Shannon entropy of x
+#' @export
+shannon_entropy <- function(x) {
+  # -|x[n]| x log(|x[n]|)
+  # Lee et al. 2016 Sensors
+  -abs(x) * log(abs(x))
+}
+
 #' Find beats in BCG with jerk cue
 #'
-#' @param jerk Smoothed jerk signal
-#' @param surge Filtered surge signal
-#' @param window_s Width of search window in seconds (1.8 s by default)
-#' @param fs_hz Sampling frequency of jerk and surge in Hz (400 Hz by default)
+#' @param jerk `[numeric]` Smoothed jerk signal
+#' @param window_s `[integer(1)]`  Width of search window in seconds
+#' @param fs_hz `[numeric(1)]` Sampling frequency of jerk and surge in Hz
+#' @param mask `[logical]` Search mask (optional)
 #'
 #' @return
 #' @export
-find_beats_jerk <- function(jerk, mask, window_s = 1.8, fs_hz = 400) {
-  jerk[!mask] <- 0
+find_beats_jerk <- function(jerk, window_s, fs_hz, mask = NULL) {
+  if (!is.null(mask))
+    jerk[!mask] <- 0
   window <- floor(window_s * fs_hz)
 
   # Find local peaks in jerk
[…]

### 2021-10-22 analysis/paper/paper.Rmd
Rearrange manuscript and include PSD
Former-commit-id: ba640e1f70d651386b7bdc5134f7ecc2795decff

--- 2021-09-05 analysis/paper/paper.Rmd
@@
 abstract: |
-  Text of abstract
+  Comparative physiologists have long been interested in heart rate, especially in breath-holding divers like marine mammals.
+To date, an electrocardiogram has only been recorded for a single large cetacean individual, but hundreds of accelerometer tags have been deployed on diverse large cetacean species.
+Here we demonstrate a ballistocardiogram method that detects resting apneic heart rate using only high-resolution accelerometer data.
+We validated the method on a killer whale (*Orcinus orca*) simultaneously tagged with accelerometer and ECG.
+We applied the method to accelerometer data from a blue whale (*Balaenoptera musculus*) and confirmed previously observed cardiovascular patterns, including the degree of diving bradycardia in this species (4-8 bpm) and progressive bradycardia relaxation during dives.
+Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and greatly expand our understanding of comparative cardiovascular physiology.
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  ballistocardiogram; diving physiology; cardiovascular physiology; accelerometer; bio-logging; marine mammal
 ---
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
 Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
   fig.path = "../figures/",
-  fig.height = 6,
-  fig.width = 6,
+  fig.height = 8,
+  fig.width = 8,
   dpi = 300
@@
 
-<!-- The actual document text starts here: -->
-
 # Introduction
@@
 
-# Background
+# Materials and methods
 
-# Methods
+## Animal tagging
 
-# Results
+* A captive killer whale (*Orcinus orca*) was double-tagged with a CATS unit and ECG recorder.
+* The CATS unit was placed behind the left pectoral flipper and the ECG recorder on the ventral center line. 
+* Trainers instructed the whale to hold a submerged position.
+* A blue whale (*Balaenoptera musculus*) was tagged with a CATS unit.
+* The tag slid behind the right pectoral flipper, where it remained overnight.
+* While tagged, the blue whale engaged in apparent resting behavior.
+We manually labeled motionless periods.
+Specifically, we looked for regions with low amplitude y-axis gyroscope signal, since fluke-strokes are readily apparent in this signal (Gough et al.
+2019).
+* CATS tags programmed to collect 400 Hz accelerometer, 50 Hz magnetometer, 50 Hz gyroscope, and 10 Hz pressure. Data processed using CATS Toolbox (Cade et al. 2021 Animal Biotelemetry)
+* Relevant permits (IACUC, NMFS, ???)
 
-<!-- Here's some example analysis code: -->
+## Signal processing
+
+The BCG waveform is three dimensional, but strongest in the longitudinal axis (Inan et al.
+2014 IEEE JBHI).
+We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of Lee et al.
+(2016 Sensors).
+
+For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.
+
+**Procedure**
+
+1. Remove noise and de-trend acceleration with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) (cite `signal`).
+2. Enhance peaks by differentiating acceleration using a 4th order Savitzky-Golay filter (cite `signal`).
+3.
+Make all peaks positive and further enhance peaks by calculating the Shannon entropy.
+$H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$ where $k$ is the acceleration axis.
+In the 1d case, $k$ is surge only.
+4. Remove noise by applying a triangular moving average smoother.
+5. Extract peaks and heuristically remove minor peaks (see supplemental).
+
+This procedure can be applied to either 1d (i.e., surge-only) or 3d acceleration. In the case of 3d acceleration, the band-pass and Savitzky-Golay filters are applied to each axis independently.
+
+## ECG validation on killer whale
+
+We calculated the percent error in paired BCG- and ECG-derived instantaneous heart rates (1d BCG only).
+
+## BCG application to blue whale
+
+We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios.
+For both BCGs, we calculated the power spectral density (cite package `psd`).
+Blue whale apneic heart rate (in the absence of high effort activities such as feeding) is 4-8 bpm (Goldbogen et al.
+2019 PNAS), so we quantified "signal" as the integration of the power spectral density curve from 4-8 bpm and "noise" as the integrated remainder, up to 60 bpm.
+
+We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals (Goldbogen et al.
+2019 PNAS, McDonald and Ponganis, 2014).
+We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration.
+Dive times were normalized from 0 (start of dive) to 1 (end of dive).
+We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (cite package `RobustLinearReg`; Sen, 1968; and Theil, 1950) due to heteroscedascity and tested whether the slope was greater than 0.
+
+## Reproducibility
+
+The data and code used in this analysis were packaged as a research compendium using `rrtools` (Marwick et al., 2018; cite `rrtools`).
+
+# Results and discussion
+
+## ECG validation on killer whale
 
 ```{r process-acc}
-fs <- 400
-fny <- fs / 2
-window <- 0.25 #s
-bandpass_filter <- function(x, pass) {
-  signal::butter(4, pass / fny, type = "pass") %>% 
-    signal::filtfilt(x)
-}
-jerk_filter <- function(x, size) {
-  signal::sgolayfilt(x, p = 4, n = size + (size %% 2) + 1, m = 1)
-}
+ca_tz <- "Etc/GMT+7"
+bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
+bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:38", tz = ca_tz)
 oo_400hz <- corky_400hz %>% 
-  mutate(
-    # Pass-band filter acceleration
-    across(surge:heave, 
-           bandpass_filter, pass = c(1.0, 25.0), 
-           .names = "{.col}_filt"),
-    # Detect heart beats using BCG's local surge range
-    surge_range = local_range(surge_filt, window * fs),
-    bcg_beat = find_beats_lsr(surge_filt, window, fs),
-    bcg_bpm = bpm(bcg_beat, dt),
-    # Calculate jerk norm and apply smoother
-    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
-                  p = 4, n = window * fs),
-    njerk_smooth = tma(njerk, window * fs)
-  )
+  filter(between(dt, bcg_begin - 10, bcg_end + 10)) %>% 
+  mutate(surge_filt = filter_acc(surge, 400, 25.0),
+         surge_diff = lead(surge_filt) - surge_filt,
+         surge_se = shannon_entropy(surge_diff),
+         surge_smooth = tma(surge_se, 0.5 * 400)) %>% 
+  filter(between(dt, bcg_begin, bcg_end)) %>% 
+  mutate(bcg_beat = find_beats(surge_smooth, 400, 0.5),
+         bcg_bpm = bpm(bcg_beat, dt))
 ```
@@
 
-```{r oo-sample}
-ca_tz <- "Etc/GMT+7"
-bcg_begin <- lubridate::ymd_hms("2021-08-16 10:09:24.5", tz = ca_tz)
-bcg_end <- lubridate::ymd_hms("2021-08-16 10:09:39", tz = ca_tz)
-bcg_sample <- filter(oo_400hz, between(dt, bcg_begin, bcg_end))
+```{r ecg_bcg_comparison}
+bcg_bpm <- filter(oo_400hz, bcg_beat) %>% 
+  select(dt, bcg_bpm)
+nearest_bcg <- function(t) {
+  bcg_bpm$bcg_bpm[which.min(abs(t - bcg_bpm$dt))]
+}
+ecg_bpm <- filter(oo_400hz, ecg_beat) %>% 
+  mutate(bcg_bpm = map_dbl(dt, nearest_bcg),
+         bpm_diff = abs(bcg_bpm - ecg_bpm) / ecg_bpm) %>% 
+  drop_na(bpm_diff)
 
-t <- theme_minimal() +
-  theme(axis.title.x = element_blank(),
-        panel.grid.minor = element_blank())
+mean_bpm_diff <- mean(ecg_bpm$bpm_diff)
+sd_bpm_diff <- sd(ecg_bpm$bpm_diff)
 ```
 
-```{r oo-bcg-ecg, fig.cap="A: BCG, B: ECG, C: BPM comparison"}
-# A: Surge + LSR
-bcg_surge_lsr <- bcg_sample %>% 
-  rename(Surge = surge_filt, LSR = surge_range) %>% 
-  pivot_longer(c(Surge, LSR), 
-               names_to = "var",
-               values_to = "val")
+The ECG and BCG yielded nearly identical heart rate estimations (Fig.
+\@ref(fig:oo-bcg-ecg)).
+We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold.
+BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean \u00B1 standard deviation).
 
-pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
-  geom_line(size = 0.2) +
-  geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
-  labs(y = bquote(Surge~(m~s^{-2}))) +
-  expand_limits(x = max(bcg_sample$dt) + 1) +
-  t +
-  theme(legend.position = "none")
+## BCG application to blue whale
 
-# B: ECG
-ecg_sample <- corky_ecg %>% 
-  filter(between(dt, first(bcg_sample$dt), last(bcg_sample$dt)))
-pB <- ggplot(ecg_sample, aes(dt, ecg_uV)) +
-  geom_line(size = 0.2) +
-  geom_point(data = filter(ecg_sample, ecg_beat), color = "red") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = bquote(ECG~(mu * V))) +
-  expand_limits(x = max(bcg_sample$dt) + 1) +
-  t
-
-# C: BCG + ECG
-pC <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
-  geom_line(size = 0.2) +
-  geom_line(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), size = 0.2) +
-  geom_point(color = "blue") +
-  geom_point(aes(y = ecg_bpm), filter(bcg_sample, ecg_beat), color = "red") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "Heart rate (bpm)") +
-  expand_limits(x = range(bcg_sample$dt) + c(0, 1), y = c(30, 70)) +
-  t
-
-cowplot::plot_grid(pA, pB, pC,
-                   align = "v", 
-                   axis = "lr",
-                   nrow = 3, 
-                   rel_heights = c(1.5, 1.0, 0.5),
-                   labels = "AUTO")
+```{r motionless}
+bw_elg <- bw180905_53_elg %>% 
+  mutate(depth_min = map2_dbl(start, stop, function(t1, t2) {
+    bw180905_53_10hz %>% 
+      filter(between(dt, t1, t2)) %>% 
+      pull(depth) %>% 
+      min()
+  }),
+  depth_max = map2_dbl(start, stop, function(t1, t2) {
+    bw180905_53_10hz %>% 
+      filter(between(dt, t1, t2)) %>% 
+      pull(depth) %>% 
+      max()
+  })
+  )
+n_motionless <- nrow(bw_elg)
+motionless_minutes <- as.numeric(sum(bw_elg$stop - bw_elg$start), unit = "mins")
 ```
 
-```{r ecg_bcg_comparison}
-ecg_bpm <- filter(bcg_sample, ecg_beat)$ecg_bpm %>% tail(-1)
-bcg_bpm <- filter(bcg_sample, bcg_beat)$bcg_bpm %>% head(-1)
-bpm_diff <- abs((bcg_bpm - ecg_bpm) / ecg_bpm * 100)
-sprintf("%0.1f%% \u00B1 %0.1f%% (mean \u00B1 sd))", mean(bpm_diff), sd(bpm_diff))
-```
-
-```{r oo-bcg-acc, fig.cap="A: The BCG signal is clearest in surge (longitudinal axis acceleration).
-Solid line: surge bandpass filtered to [1.0 25.0] Hz.
-Dashed line: local range of surge in 0.25 s sliding window (smoothed).
-Blue dots indicate peak of the BCG I-wave.
-B: BCG signal is not as strong in heave (dorso-ventral axis) or sway (lateral axis) as in surge.
-The sway signal is likely due to movement of the chest cavity (see video S1).
-C: Some, but not all, heart beats are visible as peaks in the norm of the jerk vector."}
-# A: Surge
-pA <- ggplot(bcg_sample, aes(dt, surge_filt)) +
-  geom_line(size = 0.2) +
-  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = bquote(Surge~(m~s^{-2}))) +
-  expand_limits(x = max(bcg_sample$dt) + 1) +
-  t
-
-# B: Sway, heave
-bcg_sway_heave <- bcg_sample %>% 
-  rename(Sway = sway_filt, Heave = heave_filt) %>% 
-  pivot_longer(c(Sway, Heave), 
-               names_to = "axis",
-               values_to = "acceleration")
-pB <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
-  geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
-  geom_line(size = 0.2) +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  facet_grid(axis ~ .) +
-  labs(y = bquote(Acc~(m~s^{-2}))) +
-  t
-
-# C: Jerk
-pC <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
-  geom_line(size = 0.2) +
-  geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = bquote(Jerk~(m~s^{-3}))) +
-  t
-
-cowplot::plot_grid(pA, pB, pC,
-                   align = "v", 
-                   axis = "lr",
-                   nrow = 3, 
-                   labels = "AUTO")
-```
+We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig.
+\@ref(fig:bw-profile)).
 
 ```{r bw-beats}
-window = 1.8 # s
-bw_elg <- bw180905_53_elg
-
-# Apply filters as with killer whale data, but narrower band
 bw_400hz <- bw180905_53_400hz %>%
   mutate(
-    # Pass-band filter acceleration
     across(surge:heave, 
-           bandpass_filter, pass = c(1.0, 10.0), 
+           filter_acc, fs = 400, upper = 10.0, 
            .names = "{.col}_filt"),
-    # Calculate jerk norm and apply smoother
-    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt), 
-                  p = 4, n = window * fs),
-    njerk_smooth = tma(njerk, window * fs),
+    surge_diff = lead(surge_filt) - surge_filt,
+    surge_se = shannon_entropy(surge_diff),
+    surge_smooth = tma(surge_se, 0.5 * 400),
+    jerk = jerk(cbind(surge_filt, sway_filt, heave_filt), 
+                fs = 400, p = 4, n = 2 * 400 + 1),
+    jerk_se = shannon_entropy(jerk),
+    jerk_smooth = tma(jerk_se, 2 * 400),
     # Annotate regions
-    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
-    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
+    rid_left = approx(bw_elg$start, 
+                      bw_elg$region_id, 
+                      dt, 
+                      "constant")$y,
+    rid_right = approx(bw_elg$stop, 
+                       bw_elg$region_id, 
+                       dt, 
+                       "constant", 
+                       yleft = 0)$y + 1,
     region_id = ifelse(rid_left == rid_right, rid_left, NA),
-    # Find heart beats using jerk cue
-    surge_range = local_range(surge_filt, window * fs),
-    bcg_beat = find_beats_jerk(njerk_smooth, !is.na(region_id), window, fs)
+    # Zero-out signal in non-valid regions (i.e. remove movement artifacts)
+    jerk_smooth = ifelse(is.na(region_id), 0, jerk_smooth),
+    bcg_beat = find_beats(jerk_smooth, 400, 2)
   ) %>% 
@@
 
-```{r bw-bcg}
-bw_bcg_plot <- function(rid) {
-  ca_tz <- "Etc/GMT+7"
-  bcg_sample <- filter(bw_400hz, region_id == rid)
-
-  # A: Surge + LSR
-  bcg_surge_lsr <- bcg_sample %>% 
-    rename(Surge = surge_filt, LSR = surge_range) %>% 
-    pivot_longer(c(Surge, LSR), 
-                 names_to = "var",
-                 values_to = "val")
-
-  pA <- ggplot(bcg_surge_lsr, aes(dt, val, linetype = var)) +
-    geom_line(size = 0.2) +
-    geom_point(data = filter(bcg_surge_lsr, bcg_beat), color = "blue") +
-    scale_x_datetime(date_labels = "%H:%M:%S") +
-    scale_linetype_manual(values = c(Surge = "solid", LSR = "dashed")) +
-    labs(y = bquote(Surge~(m~s^{-2}))) +
-    t +
-    theme(legend.position = "none")
-
-  # B: BPM
-  pB <- ggplot(filter(bcg_sample, bcg_beat), aes(dt, bcg_bpm)) +
-    geom_line() +
-    geom_point(color = "blue") +
-    scale_x_datetime(date_labels = "%H:%M:%S") +
-    labs(y = "Heart rate (bpm)") +
-    expand_limits(x = range(bcg_sample$dt), y = 0) +
-    t
-
-  # C: Sway, heave
-  bcg_sway_heave <- bcg_sample %>% 
-    rename(Sway = sway_filt, Heave = heave_filt) %>% 
-    pivot_longer(c(Sway, Heave), 
-                 names_to = "axis",
-                 values_to = "acceleration")
-
-  pC <- ggplot(bcg_sway_heave, aes(dt, acceleration)) +
-    geom_line(size = 0.2) +
-    geom_point(data = filter(bcg_sway_heave, bcg_beat), color = "blue") +
-    scale_x_datetime(date_labels = "%H:%M:%S") +
-    facet_grid(axis ~ .) +
-    labs(y = bquote(Acc~(m~s^{-2}))) +
-    t
-
-  # D: Jerk
-  pD <- ggplot(bcg_sample, aes(dt, njerk_smooth)) +
-    geom_line(size = 0.2) +
-    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
-    scale_x_datetime(date_labels = "%H:%M:%S") +
-    labs(y = bquote(Jerk~(m~s^{-3}))) +
-    expand_limits(y = 0) +
-    t
-
-  p <- cowplot::plot_grid(pA, pB, pC, pD,
-                          align = "v", 
-                          axis = "lr",
-                          nrow = 4, 
-                          rel_heights = c(1.5, 0.75, 1, 1),
-                          labels = "AUTO")
-
+```{r save-bw-bcg, eval=FALSE}
+walk(unique(na.omit(bw_400hz$region_id)), function(rid) {
+  p <- bw_bcg_plot(rid)
   ggsave(here::here(sprintf("analysis/figures/bw_bcg/bw_bcg_%02i.png", rid)), p)
-}
-
-# foo <- lapply(unique(na.omit(bw_400hz$region_id)), bw_bcg_plot)
+})
 ```
 
-```{r bradycardia-relax}
+The triaxial BCG (Fig.
+\@ref(fig:bw-bcg)) produced clearer signal than the surge-only BCG.
+The signal-to-noise ratio was 2.00 for the triaxial BCG, compared to 0.17 for the surge-only BCG (Fig.
+\@ref(fig:psd-plot)).
+
+```{r bw-psd}
+bcg1d_psd <- with(bw_400hz, pspectrum(surge_smooth[!is.na(region_id)], 400))
+bcg3d_psd <- with(bw_400hz, pspectrum(jerk_smooth[!is.na(region_id)], 400))
+
+bw_psd <- tibble(
+  metric = "Surge only",
+  freq_hz = bcg1d_psd$freq,
+  spec = bcg1d_psd$spec
+) %>% 
+  rbind(
+    tibble(
+      metric = "Triaxial",
+      freq_hz = bcg3d_psd$freq,
+      spec = bcg3d_psd$spec
+    )
+  ) %>% 
+  mutate(freq_bpm = freq_hz * 60) %>% 
+  filter(freq_bpm <= 60)
+
+psd_s2n <- bw_psd %>% 
+  group_by(metric) %>% 
+  summarize(signal = pracma::trapz(freq_bpm[freq_bpm >= 4 & freq_bpm <= 8],
+                                   spec[freq_bpm >= 4 & freq_bpm <= 8]),
+            noise = pracma::trapz(freq_bpm, spec) - signal) %>% 
+  mutate(s2n = signal / noise)
+```
+
+```{r brady-relax}
 bw10hz <- bw180905_53_10hz %>%
@@
   mutate(bcg_bpm = predict(bpm_ts, newdata = .))
+```
 
+3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
+Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig \@ref(bradycardia-relax)).
+
+# Acknowledgements
+
+* Everyone who helped collect and process the blue whale data
+* The Sea World trainers
+
+\newpage
+
+# Figures
+
+```{r oo-bcg-ecg, fig.cap="The ECG (A) and BCG (E) produced nearly identical heart beat predictions (red and blue points, respectively).
+B-D display the intermediate steps in the BCG signal processing procedure.
+B: surge after filtering, C: after differencing, D: Shannon entropy.
+Y-axis labeling follows Lee et al.
+(2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
+t <- theme_classic() +
+  theme(axis.title.x = element_blank(),
+        panel.grid.minor = element_blank(),
+        strip.background = element_blank())
+
+# A: ECG
+ecg_sample <- corky_ecg %>% 
+  filter(between(dt, first(oo_400hz$dt), last(oo_400hz$dt)))
+pA <- ggplot(ecg_sample, aes(dt, ecg_uV)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(ecg_sample, ecg_beat), color = "red") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = bquote(ECG~(mu * V))) +
+  expand_limits(x = max(oo_400hz$dt) + 1) +
+  t
+
+# B: Filtered surge
+pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
+  geom_line(size = 0.2) +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = bquote(f[HR])) +
+  t +
+  theme(axis.text.y = element_blank())
+
+# C: Surge diff
+pC <- ggplot(oo_400hz, aes(dt, surge_diff)) +
+  geom_line(size = 0.2) +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = "dBCG") +
+  t +
+  theme(axis.text.y = element_blank())
+
+# D: Shannon entropy
+pD <- ggplot(oo_400hz, aes(dt, surge_se)) +
+  geom_line(size = 0.2) +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = "seBCG") +
+  t +
+  theme(axis.text.y = element_blank())
+
+# E: Smoothed
+pE <- ggplot(oo_400hz, aes(dt, surge_smooth)) +
+  geom_line(size = 0.2) +
+  geom_point(data = filter(oo_400hz, bcg_beat), color = "blue") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  labs(y = "Y") +
+  t +
+  theme(axis.text.y = element_blank())
+
+cowplot::plot_grid(pA, pB, pC, pD, pE,
+                   align = "v", 
+                   axis = "lr",
+                   ncol = 1, 
+                   labels = "AUTO")
+```
+
+```{r bw-profile, fig.cap="Blue whale dive profile. Motionless periods indicated by pink boxes."}
+ggplot(bw180905_53_10hz, aes(dt, depth)) +
+  geom_line(size = 0.5) +
+  geom_rect(aes(xmin = start, xmax = stop, ymin = depth_min, ymax = depth_max),
+            bw_elg,
+            inherit.aes = FALSE,
+            fill = "orchid1",
+            color = NA,
+            alpha = 0.5) +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  scale_y_reverse("Depth (m)") +
+  theme_classic() +
+  theme(axis.title.x = element_blank())
+```
+
+```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG.
+A: filtered triaxial acceleration, B: after differencing, C: Shannon entropy, D: after smoothing.
+Identified heart beats in blue.
+Y-axis labeling follows Lee et al.
+(2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
+bw_bcg_plot <- function(rid) {
+  bcg_sample <- filter(bw_400hz, region_id == rid)
+
+  t <- theme_classic() +
+  theme(axis.title.x = element_blank(),
+        panel.grid.minor = element_blank(),
+        strip.background = element_blank())
+
+  # A: Filtered surge
+  pA <- bcg_sample %>% 
+    rename(Surge = surge_filt, Sway = sway_filt, Heave = heave_filt) %>% 
+    pivot_longer(Surge:Heave, names_to = "axis", values_to = "acc") %>% 
+    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
+    ggplot(aes(dt, acc)) +
+    geom_line(size = 0.2) +
+    scale_x_datetime(date_labels = "%H:%M:%S") +
+    scale_y_continuous(n.breaks = 3) +
+    facet_grid(rows = vars(axis)) +
+    labs(y = bquote(f[HR])) +
+    t +
+    theme(axis.text.y = element_blank())
+
+  # B: Jerk vectors
+  pB <- bcg_sample %>% 
+    mutate(jerk = set_names(as_tibble(jerk), c("Surge", "Sway", "Heave"))) %>% 
+    unpack(jerk) %>% 
+    pivot_longer(Surge:Heave, names_to = "axis", values_to = "jerk") %>% 
+    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
+    ggplot(aes(dt, jerk)) +
+    geom_line(size = 0.2) +
+    scale_x_datetime(date_labels = "%H:%M:%S") +
+    scale_y_continuous(n.breaks = 3) +
+    facet_grid(rows = vars(axis)) +
+    labs(y = "dBCG") +
+    t +
+    theme(axis.text.y = element_blank())
+
+  # C: Shannon entropy of jerk (NOT jerk norm)
+  pC <- ggplot(bcg_sample, aes(dt, jerk_se)) +
+    geom_line(size = 0.2) +
+    scale_x_datetime(date_labels = "%H:%M:%S") +
+    labs(y = "seBCG") +
+    t +
+    theme(axis.text.y = element_blank())
+
+  # D: Smoothed
+  pD <- ggplot(bcg_sample, aes(dt, jerk_smooth)) +
+    geom_line(size = 0.2) +
+    geom_point(data = filter(bcg_sample, bcg_beat), color = "blue") +
+    scale_x_datetime(date_labels = "%H:%M:%S") +
+    labs(y = "Y") +
+    t +
+    theme(axis.text.y = element_blank())
+
+  cowplot::plot_grid(pA, pB, pC, pD,
+                     align = "v", 
+                     axis = "lr",
+                     ncol = 1, 
+                     labels = "AUTO")
+}
+
+bw_bcg_plot(23)
+```
+
+```{r psd-plot, fig.cap="Signal-to-noise ratio was higher for the triaxial BCG (lower panel, 2.00) than the surge-only BCG (upper panel, 0.17).
+Each panel shows the power spectral density for the BCG.
+Based on previously observed blue whale heart rates, 4-8 bpm was considered signal (gray shading).
+The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal frequencies to the area under the rest of the curve, up to 60 bpm."}
+labels <- bw_psd %>% 
+  filter(between(freq_bpm, 7.99, 8.01)) %>% 
+  group_by(metric) %>% 
+  slice(1) %>% 
+  ungroup() %>% 
+  left_join(psd_s2n, by = "metric") %>% 
+  mutate(s2n_text = sprintf("Signal:noise = %0.2f", s2n))
+ggplot(bw_psd, aes(x = freq_bpm, y = spec)) +
+  geom_area(data = filter(bw_psd, between(freq_bpm, 4, 8)), fill = "grey80") +
+  geom_path() +
+  geom_text(aes(label = s2n_text), labels, hjust = -0.1, vjust = -0.05) +
+  labs(x = "Frequency (bpm)",
+       y = "Power spectrum density") +
+  facet_grid(rows = vars(metric),
+             scales = "free_y") +
+  theme_classic() +
+  theme(strip.background = element_blank(),
+        strip.text = element_blank())
+```
+
+```{r bradycardia-relax, fig.cap="Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
+Bradycardia is greatest at the start of the dive (~4-5 bpm), increasing towards the end (~8-9 bpm)."}
 ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
@@
   geom_line(data = bpm_pred, color = "blue", size = 1.5) +
-  scale_x_continuous(labels = scales::percent) +
+  scale_x_continuous() +
   labs(x = "Dive time (normalized)",
@@
 ```
-
-The magnitude of bradycardia was strongest at the start of dives and relaxed towards the end.
-On average, heartrate increased from `round(predict(bpm_ts, data.frame(dive_norm = 0)), 1)` bpm at the start of dives to `round(predict(bpm_ts, data.frame(dive_norm = 1)), 1)` bpm by the end (Theil-Sen regression, $p < 10^{-10}$).
-
-```{r jerk-peaks, fig.cap="Major peak selection procedure for identifying heartbeats.
-A Even after smoothing, the norm of the jerk vector has peaks that do not correspond to heartbeats (e.g.
-2, 4).
-B In height-prominence space, there are two clear clusters corresponding to valid heartbeats (upper right) and artifacts (lower left).
-C "}
-# Visualize peak thresholding algorithm
-jerk <- bw_400hz$njerk_smooth
-jerk[is.na(bw_400hz$region_id)] <- 0
-window_s <- 1.8
-fs_hz <- 400
-
-# Find peaks
-peaks <- pracma::findpeaks(jerk, minpeakdistance = window * fs_hz)
-jerk_peaks <- slice(bw_400hz, peaks[, 2]) %>% 
-  mutate(
-    height = peaks[, 1],
-    prominence = peak_prominences(jerk, peaks[, 2]),
-    distance = sqrt((max(height) - height)^2 + (max(prominence) - prominence)^2)
-  ) %>% 
-  arrange(dt)
-
-# Using region 14 as an example
-jerk_example <- filter(bw_400hz, region_id == 14)
-peak_example <- jerk_peaks %>% 
-  filter(between(dt, first(jerk_example$dt), last(jerk_example$dt))) %>% 
-  mutate(i = row_number())
-
-# Labeled peaks in jerk
-pA <- ggplot(jerk_example, aes(dt, njerk_smooth)) +
-  geom_line() +
-  geom_point(data = peak_example) +
-  ggrepel::geom_text_repel(aes(label = i), peak_example) +
-  expand_limits(y = 0) +
-  labs(y = bquote(Jerk~(m~s^{-3}))) +
-  theme_classic()
-# Peak height, prominence
-pB <- ggplot(peak_example, aes(height, prominence)) +
-  geom_point() +
-  ggrepel::geom_text_repel(aes(label = i), max.overlaps = Inf) +
-  coord_fixed() +
-  theme_classic()
-
-# All heights, prominences (with threshold)
-
-height_prom <- jerk_peaks %>% select(height, prominence)
-peak_pca <- prcomp(height_prom, scale = FALSE)
-pc1 <- peak_pca$x[, "PC1"]
-pc1_density <- density(pc1)
-pc1_peaks <- pracma::findpeaks(pc1_density$y)
-pc1_valley <- pracma::findpeaks(-pc1_density$y[pc1_peaks[1, 2]:pc1_peaks[2, 2]])
-pc1_thr <- pc1_density$x[pc1_valley[1, 2] + pc1_peaks[1, 2] - 1]
-pc1_axis <- c(peak_pca$sdev[1], 0) %*% t(peak_pca$rotation) %>% 
-  scale(center = -colMeans(height_prom), scale = FALSE)
-pc2_axis <- c(0, peak_pca$sdev[2]) %*% t(peak_pca$rotation) %>% 
-  scale(center = -colMeans(height_prom), scale = FALSE)
-pc1_density_rotate <- cbind(pc1_density$x, pc1_density$y) %*% 
-  t(peak_pca$rotation) %>% 
-  scale(center = -colMeans(height_prom), scale = FALSE)
-pc1_thr_rot <- c(pc1_thr, 0) %*% t(peak_pca$rotation) %>% 
-  scale(center = -colMeans(height_prom), scale = FALSE)
-slope <- (pc2_axis - peak_pca$center)[2] / (pc2_axis - peak_pca$center)[1]
-intercept <- pc1_thr_rot[2] - slope * pc1_thr_rot[1]
-pc1_angle <- atan2(peak_pca$rotation[2, 1], peak_pca$rotation[1, 1]) * 180 / pi
-pc2_angle <- pc1_angle - 90
-
-pC <- ggplot(jerk_peaks, aes(height, prominence)) +
-  geom_point() +
-  annotate("segment", 
-           x = peak_pca$center[1], y = peak_pca$center[2], 
-           xend = pc1_axis[1], yend = pc1_axis[2],
-           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
-  annotate("text", label = "PC1",
-           x = pc1_axis[1], y = pc1_axis[2],
-           hjust = -0.1, angle = pc1_angle,
-           color = "red") + 
-  annotate("segment", 
-           x = peak_pca$center[1], y = peak_pca$center[2], 
-           xend = pc2_axis[1], yend = pc2_axis[2],
-           color = "red", arrow = arrow(length = unit(0.30, "cm"))) +
-  annotate("text", label = "PC2",
-           x = pc2_axis[1], y = pc2_axis[2],
-           hjust = 1.1, angle = pc2_angle,
-           color = "red") + 
-  geom_abline(slope = slope, intercept = intercept, 
-              color = "red", linetype = 2) +
-  coord_fixed() +
-  theme_classic()
-
-# density along pc1
-pD <- as_tibble(peak_pca$x) %>% 
-  ggplot(aes(PC1)) +
-  geom_histogram(aes(y = ..density..), bins = 50) +
-  geom_density(color = "blue") +
-  geom_vline(xintercept = pc1_thr, color = "red", linetype = 2) +
-  theme_classic()
-
-cowplot::plot_grid(pA, pB, pC, pD,
-                   nrow = 2,
-                   labels = "AUTO")
-```
-
-Figure \@ref(fig:oo-bcg-ecg) shows how we can have a caption and cross-reference for a plot.
-
-# Discussion
-
-# Conclusion
-
-# Acknowledgements
-
-<!-- The following line inserts a page break  -->
-
-\newpage
 

### 2021-10-22 R/find_beats.R

@@ -1,120 +1,50 @@
-#' Find beats in BCG with local surge range cue
+#' Find beats in BCG
 #'
-#' @param surge Filtered surge signal
-#' @param window_s Width of search window in seconds (0.25 s by default)
-#' @param fs_hz Sampling frequency of surge in Hz (400 Hz by default)
+#' @param cue `[numeric]` signal used for heart beat cue.
+#' @param fs_hz `[numeric(1)]` Sampling frequency of surge in Hz
+#' @param minperiod `[numeric(1)]` Minimum period between beats (in seconds)
 #'
-#' @return A logical vector the same length as surge indicating a heart beat
+#' @return `[logical]` same length as cue indicating heart beats
 #' @export
-find_beats_lsr <- function(surge, window_s = 0.25, fs_hz = 400) {
-  window <- floor(window_s * fs)
+find_beats <- function(cue, fs_hz, minperiod) {
+  peaks <- pracma::findpeaks(cue, minpeakdistance = fs_hz * minperiod)
+  proms <- peak_prominences(cue, peaks[, 2])
 
-  # Local range of surge
-  surge_range <- local_range(surge, window)
-
-  # Find local peaks in smoothed local range
-  local_range_max <- RcppRoll::roll_max(surge_range, window, fill = NA)
-  range_peaks <- surge_range == local_range_max
-
-  # Keep only most prominent peaks by looking for a break in the distribution
-  range_peak_idx <- which(range_peaks)
-  prominence <- peak_prominences(surge_range, range_peak_idx)
-  prominence_asc <- sort(prominence)
-  prominence_steps <- diff(prominence_asc)
-  prominence_thr <- prominence_asc[which.max(prominence_steps)]
-  range_peaks[range_peak_idx[prominence < prominence_thr]] <- FALSE
-
-  # Align range peaks with surge peaks
-  beats <- logical(length(range_peaks))
-  i <- 1
-  for (peak in which(range_peaks)) {
-    i1 <- max(peak - window / 2, 1)
-    i2 <- min(peak + window / 2, length(surge))
-    b <- which.max(surge[i1:i2]) + i1 - 1
-    beats[b] <- TRUE
-    i <- i + 1
[…]

### 2021-10-22 R/psd.R

@@ -1,0 +1,28 @@
+#' Estimate power spectral density
+#'
+#' @param x `[numeric]` signal
+#' @param fs_hz `[numeric(1)]` sampling rate (Hz)
+#'
+#' @return `[spec]` see `psd::pspectrum`
+#' @export
+pspectrum <- function(x, fs_hz) {
+  suppressMessages(psd::pspectrum(ts(x, frequency = fs_hz)))
+}
+
+#' Calculate signal-to-noise ratio from power spectral density
+#'
+#' @param psd `[spec]` power spectral density
+#' @param signal_bpm `[numeric(2)]` signal band (bpm)
+#'
+#' @return
+#' @export
+s2n <- function(psd, signal_bpm) {
+  signal_hz <- signal_bpm / 60
+  is_signal <- between(psd$freq, signal_hz[1], signal_hz[2])
+  is_noise <- !is_signal
+
+  psignal <- pracma::trapz(psd$freq[is_signal], psd$spec[is_signal])
+  pnoise <- pracma::trapz(psd$freq[is_noise], psd$spec[is_noise])
+
+  psignal / pnoise
+}

### 2021-10-22 R/util.R

@@ -9,10 +9,46 @@
 #'
 #' @return Norm of the jerk vector of A
 #' @export
-njerk <- function(A, fs, p = 3, n = 11) {
+jerk <- function(A, fs, p = 3, n = 11) {
   n <- n + 1 - n %% 2
-  jerk <- apply(A, 2, function(axis) signal::sgolayfilt(axis, p, n, m = 1))
-  apply(jerk, 1, function(row) sqrt(sum(row^2)))
+  apply(A, 2, function(axis) signal::sgolayfilt(axis, p, n, m = 1))
+}
+
+#' Row-wise norm
+#'
+#' @param m `[numeric matrix]`
+#'
+#' @return `[numeric vector]`
+rowNorms <- function(m) {
+  sqrt(rowSums(m^2))
+}
+
+#' Triangular moving average
+#'
+#' @param x Original signal
+#' @param k Window width
+#'
+#' @return Smoothed x
+#' @export
+tma <- function(x, k) {
+  x %>%
+    RcppRoll::roll_mean(k, fill = NA) %>%
+    RcppRoll::roll_mean(k, fill = NA)
+}
+
+#' Shannon entropy
+#'
+#' Defined as -sum(|x| * log(|x|))
+#'
+#' @param x `[numeric] vector or matrix`
+#'
+#' @return Shannon entropy of x
+#' @export
+shannon_entropy <- function(x) {
+  if (is.vector(x)) {
+    x <- matrix(x, ncol = 1)
+  }
+  apply(x, 1, function(row) -sum(abs(row) * log(abs(row))))
 }
 
 #' Calculate beats-per-minute

### 2021-10-22 analysis/interact/global.R

@@ -1,4 +1,7 @@
+library(cetaceanbcg)
+library(shiny)
 library(shinybusy)
+library(tidyverse)
 
 filter_thr0 <- c(1.5, 7.5)
 filter_order0 <- 4
@@ -6,7 +9,9 @@ window0 <- 1.8
 yaw_offset0 <- 0
 
 n_pts <- 1e3
-bw180905_53_10hz_sparse <- bw180905_53_10hz[floor(seq(1, nrow(bw180905_53_10hz), length.out = n_pts)), ]
+bw180905_53_10hz_sparse <- bw180905_53_10hz[floor(seq(1,
+                                                      nrow(bw180905_53_10hz),
+                                                      length.out = n_pts)), ]
 
 update_results <- function(hyperparams) {
   show_modal_progress_circle(.10, "Filtering acceleration")

### 2021-10-22 analysis/interact/server.R

@@ -1,7 +1,3 @@
-library(cetaceanbcg)
-library(shiny)
-library(tidyverse)
-
 server <- function(input, output, session) {
   # Hyperparameters
   hyperparams <- reactive(list(

### 2021-10-22 analysis/scratchpad/divesegmentation.R

@@ -36,7 +36,7 @@ segment_phases <- function(depth, dive_id) {
 }
 
 normalize_dive <- function(dive_phase, dive_id) {
-  by_dive <-split(dive_phase, dive_id) %>%
+  by_dive <- split(dive_phase, dive_id) %>%
     lapply(function(x) {
       result <- rep(NA, length(x))
       b <- which(x == "bottom")

### 2021-10-24 analysis/paper/paper.Rmd
Rough draft
Former-commit-id: 3b8b997db2f0ae536a7b0ed5a850970cc33275bd

@@
 ---
-title: "Accelerometer-derived ballistocardiogram confirms blue whale (*Balaenoptera musculus*) heart rate during diving bradycardia"
+title: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
 author:
-  - Max Czapanskiy:
+  - Max F. Czapanskiy:
       email: maxczap@stanford.edu
-      institute: [SU]
+      institute: [Hopkins]
       correspondence: true
-  - More Authors:
-      email: more@authors.edu
-      institute: [UofA]
+  - Paul J. Ponganis:
+      institute: [Scripps]
+      correspondence: false
+  - James A. Fahlbusch:
+      institute: [Hopkins]
+      correspondence: false
+  - Jeremy A. Goldbogen:
+      institute: [Hopkins]
       correspondence: false
 institute:
-  - SU: Stanford University
-  - UofA: University of Another Place
+  - Hopkins: Hopkins Marine Station, Department of Biology, Stanford University, USA
+  - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      reference_docx: "../templates/template.docx"
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+bibliography: [zotero.bib,references.bib]
+csl: "../templates/the-journal-of-experimental-biology.csl"
 abstract: |
-  Comparative physiologists have long been interested in heart rate, especially in breath-holding divers like marine mammals.
-To date, an electrocardiogram has only been recorded for a single large cetacean individual, but hundreds of accelerometer tags have been deployed on diverse large cetacean species.
-Here we demonstrate a ballistocardiogram method that detects resting apneic heart rate using only high-resolution accelerometer data.
-We validated the method on a killer whale (*Orcinus orca*) simultaneously tagged with accelerometer and ECG.
-We applied the method to accelerometer data from a blue whale (*Balaenoptera musculus*) and confirmed previously observed cardiovascular patterns, including the degree of diving bradycardia in this species (4-8 bpm) and progressive bradycardia relaxation during dives.
-Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and greatly expand our understanding of comparative cardiovascular physiology.
+  Physio-logging methods, which use animal-borne devices to record physiological variables, are entering a new era driven by advances in sensor development.
+However, existing datasets collected with traditional bio-loggers, such as accelerometers, still contain untapped eco-physiological information.
+Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram.
+We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and progressive relaxation of bradycardia during dives.
+Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
 keywords: |
@@
   echo = FALSE,
+  cache = TRUE,
   comment = "#>",
   fig.path = "../figures/",
-  fig.height = 8,
-  fig.width = 8,
+  fig.height = 6,
+  fig.width = 6,
   dpi = 300
@@
 
-Here is a citation [@Marwick2017]
+Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021].
+For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021].
+However, legacy tags with inertial measurement unit (IMU) sensors - accelerometers, magnetometers, and gyroscopes - can also measure important physiological variables under certain conditions.
+In one case, accelerometers revealed the elevated respiratory rate of an emperor penguin (*Aptenodytes forsteri*) following a 27.6 minute dive, the longest recorded for this species [@satoStrokeRatesDiving2011].
+While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes.
+This is particularly important for species that cannot be restrained or studied in captivity.
+For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag on one blue whale (*Balaenoptera musculus*) individual [@goldbogenExtremeBradycardiaTachycardia2019].
+Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001] .
+These existing datasets (and future IMU tag deployments) could hold valuable physiological information, but we lack proper computational methods for mining them.
+
+The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors.
+Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels.
+The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and phonocardiography.
+However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019].
+While the BCG is a three-dimensional phenomenon, it is strongest in the longitudinal (anterior-posterior) axis [@inanBallistocardiographySeismocardiographyReview2015].
+Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010].
+The BCG J wave is analagous to the ECG R wave in that it is the most robust feature in the waveform and typically used for detecting heart beats.
+
+Here we present a method for generating a BCG from bio-logger accelerometry.
+We validated our method with a simultaneously recorded electrocardiogram (ECG) on a captive killer whale (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
+The relative orientation of the tag on the body is uncertain in cetacean bio-logging [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on longitudinal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
+Specifically, we tested three hypotheses to validate our method.
+First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG.
+Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG.
+Third, BCG-derived heartrates would increase from the start to end of dives, consistent with the marine mammal cardiophysiological dive response of progressive bradycardia relaxation [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 
@@
 
-* A captive killer whale (*Orcinus orca*) was double-tagged with a CATS unit and ECG recorder.
+* A captive killer whale was double-tagged with a CATS unit and ECG recorder.
 * The CATS unit was placed behind the left pectoral flipper and the ECG recorder on the ventral center line. 
 * Trainers instructed the whale to hold a submerged position.
-* A blue whale (*Balaenoptera musculus*) was tagged with a CATS unit.
+* A blue whale was tagged with a CATS unit (originally published in [@goldbogenWhyWhalesAre2019]).
 * The tag slid behind the right pectoral flipper, where it remained overnight.
@@
 We manually labeled motionless periods.
-Specifically, we looked for regions with low amplitude y-axis gyroscope signal, since fluke-strokes are readily apparent in this signal (Gough et al.
-2019).
-* CATS tags programmed to collect 400 Hz accelerometer, 50 Hz magnetometer, 50 Hz gyroscope, and 10 Hz pressure. Data processed using CATS Toolbox (Cade et al. 2021 Animal Biotelemetry)
+Specifically, we looked for regions with low amplitude y-axis gyroscope signal, since fluke-strokes are readily apparent in this signal [@goughScalingSwimmingPerformance2019]
+* CATS tags programmed to collect 400 Hz accelerometer, 50 Hz magnetometer, 50 Hz gyroscope, and 10 Hz pressure. Data processed using CATS Toolbox [@cadeToolsIntegratingInertial2021].
 * Relevant permits (IACUC, NMFS, ???)
@@
 
-The BCG waveform is three dimensional, but strongest in the longitudinal axis (Inan et al.
-2014 IEEE JBHI).
-We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of Lee et al.
-(2016 Sensors).
+The BCG waveform is three dimensional, but strongest in the longitudinal axis [@inanBallistocardiographySeismocardiographyReview2015].
+We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].
 
@@
 
-1. Remove noise and de-trend acceleration with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) (cite `signal`).
-2. Enhance peaks by differentiating acceleration using a 4th order Savitzky-Golay filter (cite `signal`).
+1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) using package `signal` [@R-signal].
+2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter using package `signal`. Differentiation exaggerates impulses like the J wave.
 3.
-Make all peaks positive and further enhance peaks by calculating the Shannon entropy.
-$H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$ where $k$ is the acceleration axis.
+Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis).
+Additionally, the Shannon entropy is strictly positive, which facilitates peak detection.
 In the 1d case, $k$ is surge only.
 4. Remove noise by applying a triangular moving average smoother.
-5. Extract peaks and heuristically remove minor peaks (see supplemental).
+5. Extract peaks and heuristically remove noisy peaks (see supplemental).
 
-This procedure can be applied to either 1d (i.e., surge-only) or 3d acceleration. In the case of 3d acceleration, the band-pass and Savitzky-Golay filters are applied to each axis independently.
+This procedure can be applied to either 1d (i.e., surge-only) or 3d acceleration. In the case of 3d acceleration, the band-pass and Savitzky-Golay filters were applied to each axis independently.
 
@@
 
-We calculated the percent error in paired BCG- and ECG-derived instantaneous heart rates (1d BCG only).
+We fit ordinary least squares regression to BCG-derived instantaneous heart rates with respect to ECG-derived and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1.
+We calculated the mean and standard deviation of absolute error as an equivalence measure (1d BCG only).
 
@@
 We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios.
-For both BCGs, we calculated the power spectral density (cite package `psd`).
-Blue whale apneic heart rate (in the absence of high effort activities such as feeding) is 4-8 bpm (Goldbogen et al.
-2019 PNAS), so we quantified "signal" as the integration of the power spectral density curve from 4-8 bpm and "noise" as the integrated remainder, up to 60 bpm.
+For both BCGs, we calculated the power spectral density using package `psd` [@Barbour2014].
+Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.
 
-We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals (Goldbogen et al.
-2019 PNAS, McDonald and Ponganis, 2014).
+We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration.
 Dive times were normalized from 0 (start of dive) to 1 (end of dive).
-We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (cite package `RobustLinearReg`; Sen, 1968; and Theil, 1950) due to heteroscedascity and tested whether the slope was greater than 0.
+We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) with package `RobustLinearReg` [@R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.
 
@@
 
-The data and code used in this analysis were packaged as a research compendium using `rrtools` (Marwick et al., 2018; cite `rrtools`).
+The data and code used in this analysis were packaged as a research compendium using package `rrtools` [@marwickPackagingDataAnalytical2018; @rrtools2019].
+The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.
 
@@
 sd_bpm_diff <- sd(ecg_bpm$bpm_diff)
+
+bpm_lm <- lm(bcg_bpm ~ ecg_bpm, ecg_bpm)
+bpm_stderr <- sqrt(diag(vcov(bpm_lm)))
 ```
@@
 We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold.
-BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean \u00B1 standard deviation).
+BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation).
+The ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
 
@@
 
-The triaxial BCG (Fig.
-\@ref(fig:bw-bcg)) produced clearer signal than the surge-only BCG.
+The 3d BCG (Fig.
+\@ref(fig:bw-bcg)) produced a more robust signal than the surge-only 1d BCG.
 The signal-to-noise ratio was 2.00 for the triaxial BCG, compared to 0.17 for the surge-only BCG (Fig.
@@
 
-```{r brady-relax}
+```{r brady-relax-model}
 bw10hz <- bw180905_53_10hz %>%
@@
 3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
-Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig \@ref(bradycardia-relax)).
+Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig \@ref(fig:brady-relax)).
+
+## Reproducibility
+
+The research compendium containing data, code, and an executable version of this manuscript was archived on Zenodo (TODO).
+We developed the research compendium as an R package to facilitate investigation and adoption by other researchers.
+Publishing data and code in standardized formats (such as an R package) is a critical step towards transparency and computational reproducibility [@alstonBeginnerGuideConducting2021; @powersOpenScienceReproducibility2019; @stoddenEmpiricalAnalysisJournal2018].
+
+## Conclusions
+
+Here we presented a ballistocardiogram method for detecting resting apneic heartrate in cetaceans using accelerometers.
+We validated the method in a controlled setting with simultaneous ECG and in a field setting by confirming expected physiological patterns.
+As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors.
+Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can still derive new insights from traditional sensors.
 
@@
 
-* Everyone who helped collect and process the blue whale data
-* The Sea World trainers
+* Everyone who helped collect and process the blue whale data.
+* The Sea World trainers.
+* Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
 
@@
 Based on previously observed blue whale heart rates, 4-8 bpm was considered signal (gray shading).
-The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal frequencies to the area under the rest of the curve, up to 60 bpm."}
+The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal band to the area under the rest of the curve, up to 60 bpm."}
 labels <- bw_psd %>% 
@@
 
-```{r bradycardia-relax, fig.cap="Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
-Bradycardia is greatest at the start of the dive (~4-5 bpm), increasing towards the end (~8-9 bpm)."}
+```{r brady-relax, fig.cap="Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
+Bradycardia is greatest at the start of the dive (~4-5 bpm), relaxing towards the end (~8-9 bpm).
+Points indicate instantaneous heart rates and the line is a Theil-Sen regression."}
 ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +

### 2021-10-24 R/bib.R

@@ -1,0 +1,44 @@
+#' Create crossref.bib
+#'
+#' Uses package rcrossref to look up DOIs in dois.txt and create crossref.bib.
+write_crossref <- function() {
+  for (pkg in c("rcrossref", "bib2df")) {
+    if (!requireNamespace(pkg, quietly = TRUE)) {
+      stop(sprintf("write_crossref() requires package '%s'", pkg))
+    }
+  }
+  # Fetch from crossref
+  dois <- readLines("analysis/paper/dois.txt")
+  refs <- rcrossref::cr_cn(dois = dois, format = "bibtex")
+  bib_path <- "analysis/paper/crossref.bib"
+  if (file.exists(bib_path)) {
+    message("Overwriting crossref.bib")
+    file.remove(bib_path)
+  }
+  file.create(bib_path)
+  for (ref in refs) {
+    write(paste0(ref, "\n"), file = bib_path, append = TRUE)
+  }
+
+  # Fix entry names
+  get_first_author <- function(author_list) {
+    author_list[1] %>% stringr::str_match(".* ([^ ]+)$") %>% `[`(1, 2)
+  }
+  append_suffix <- function(keys) {
+    if (length(keys) > 1) {
+      paste0(keys, letters[seq(length(keys))])
+    } else {
+      keys
+    }
+  }
+  bibdf <- bib2df::bib2df(bib_path) %>%
+    dplyr::mutate(BIBTEXKEY = paste(purrr::map_chr(AUTHOR, get_first_author),
+                                    YEAR,
+                                    sep = "-")) %>%
+    dplyr::group_by(BIBTEXKEY) %>%
+    dplyr::mutate(BIBTEXKEY = append_suffix(BIBTEXKEY)) %>%
+    dplyr::ungroup()
+  bib2df::df2bib(bibdf, bib_path)
+}
+
+

### 2021-10-24 R/dives.R

@@ -10,7 +10,7 @@
 #'   to individual dives. Inter-dive periods marked as NA.
 #' @export
 split_dives <- function(dt, depth, surface = 0, min_depth = 0, min_dur = 0) {
-  dive_start <- depth >= surface & lag(depth) < surface
+  dive_start <- depth >= surface & dplyr::lag(depth) < surface
   dive_start[1] <- FALSE
   dive_id <- cumsum(dive_start)
   dive_id[depth < surface | dive_id == 0] <- NA
@@ -28,7 +28,7 @@ split_dives <- function(dt, depth, surface = 0, min_depth = 0, min_dur = 0) {
 
 #' Normalize dive time
 #'
-#' Normalize the time in each dive to [0, 1] where 0 is start of dive and 1 is
+#' Normalize the time in each dive to `[0, 1]` where 0 is start of dive and 1 is
 #' the end.
 #'
 #' @param dive_id A factor of dive id's, as from \code{split_dives()}.

### 2021-10-24 R/find_beats.R

@@ -14,7 +14,7 @@ find_beats <- function(cue, fs_hz, minperiod) {
   # Skip if less than 50 peaks
   if (nrow(peaks) >= 50) {
     dist_max <- sqrt((max(peaks[, 1]) - peaks[, 1])^2 + (max(proms) - proms)^2)
-    dist_dens <- density(dist_max)
+    dist_dens <- stats::density(dist_max)
     dist_thr <- dist_dens$x[pracma::findpeaks(-dist_dens$y)[, 2]]
     is_beat <- peaks[, 2][dist_max <= dist_thr]
   } else {
@@ -57,7 +57,7 @@ filter_acc <- function(acc, fs, upper) {
 choose_peaks <- function(x) {
   x_sorted <- sort(x)
   steps <- diff(x_sorted)
-  thr <- if (max(steps) > mean(steps) + 2 * sd(steps)) {
+  thr <- if (max(steps) > mean(steps) + 2 * stats::sd(steps)) {
     x_sorted[which.max(steps)]
   } else {
     x_sorted[1]

### 2021-10-24 R/psd.R

@@ -6,7 +6,7 @@
 #' @return `[spec]` see `psd::pspectrum`
 #' @export
 pspectrum <- function(x, fs_hz) {
-  suppressMessages(psd::pspectrum(ts(x, frequency = fs_hz)))
+  suppressMessages(psd::pspectrum(stats::ts(x, frequency = fs_hz)))
 }
 
 #' Calculate signal-to-noise ratio from power spectral density
@@ -14,11 +14,11 @@ pspectrum <- function(x, fs_hz) {
 #' @param psd `[spec]` power spectral density
 #' @param signal_bpm `[numeric(2)]` signal band (bpm)
 #'
-#' @return
+#' @return `[numeric(1)]`
 #' @export
 s2n <- function(psd, signal_bpm) {
   signal_hz <- signal_bpm / 60
-  is_signal <- between(psd$freq, signal_hz[1], signal_hz[2])
+  is_signal <- dplyr::between(psd$freq, signal_hz[1], signal_hz[2])
   is_noise <- !is_signal
 
   psignal <- pracma::trapz(psd$freq[is_signal], psd$spec[is_signal])

### 2021-10-24 R/zzz.R

@@ -1,2 +1,0 @@
-#' @import tidyverse
-NULL

### 2021-10-24 README.Rmd
Update README

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
+Authors <- "Czapanskiy, M. F., Ponganis, P. J., Fahlbusch, J. A., and Goldbogen, J. A."
 ```
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///main?urlpath=rstudio)
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
-
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors` (in prep). _`r Title`_.
 
@@
 
-Please cite this compendium as:
+Following publication, please cite this compendium as:
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> `r Authors` (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for "`r Title`"_. Accessed `r format(Sys.Date(), "%d %b %Y")`.
 
@@
 **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
-
-### Contributions
-
-We welcome contributions from everyone.
-Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
-Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md).
-By participating in this project you agree to abide by its terms.

### 2021-10-24 R/bib.R
Remove unused dependencies

@@ -1,44 +1,0 @@
-#' Create crossref.bib
-#'
-#' Uses package rcrossref to look up DOIs in dois.txt and create crossref.bib.
-write_crossref <- function() {
-  for (pkg in c("rcrossref", "bib2df")) {
-    if (!requireNamespace(pkg, quietly = TRUE)) {
-      stop(sprintf("write_crossref() requires package '%s'", pkg))
-    }
-  }
-  # Fetch from crossref
-  dois <- readLines("analysis/paper/dois.txt")
-  refs <- rcrossref::cr_cn(dois = dois, format = "bibtex")
-  bib_path <- "analysis/paper/crossref.bib"
-  if (file.exists(bib_path)) {
-    message("Overwriting crossref.bib")
-    file.remove(bib_path)
-  }
-  file.create(bib_path)
-  for (ref in refs) {
-    write(paste0(ref, "\n"), file = bib_path, append = TRUE)
-  }
-
-  # Fix entry names
-  get_first_author <- function(author_list) {
-    author_list[1] %>% stringr::str_match(".* ([^ ]+)$") %>% `[`(1, 2)
-  }
-  append_suffix <- function(keys) {
-    if (length(keys) > 1) {
-      paste0(keys, letters[seq(length(keys))])
-    } else {
-      keys
-    }
-  }
-  bibdf <- bib2df::bib2df(bib_path) %>%
-    dplyr::mutate(BIBTEXKEY = paste(purrr::map_chr(AUTHOR, get_first_author),
-                                    YEAR,
-                                    sep = "-")) %>%
-    dplyr::group_by(BIBTEXKEY) %>%
-    dplyr::mutate(BIBTEXKEY = append_suffix(BIBTEXKEY)) %>%
-    dplyr::ungroup()
-  bib2df::df2bib(bibdf, bib_path)
-}
-
-

### 2021-10-27 README.Rmd
Add DOCKERFILE
Incorporate co-author feedback

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///main?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/FlukeAndFeather/cetaceanbcg/main?urlpath=rstudio)
 
@@
     illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
+  - [:file\_folder: supplementary-materials](/analysis/supplementary-materials):
     Supplementary materials including notes and other documents
@@
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+[main.zip](/archive/main.zip). After unzipping:
 - open the `.Rproj` file in RStudio

### 2021-10-27 analysis/paper/paper.Rmd
Incorporate co-author feedback

@@
       correspondence: false
+  - T. L. Schmitt:
+      institute: [SeaWorld]
+      correspondence: false
   - Jeremy A. Goldbogen:
@@
   - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
+  - SeaWorld: Veterinary Services, SeaWorld of California, USA
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021].
-However, legacy tags with inertial measurement unit (IMU) sensors - accelerometers, magnetometers, and gyroscopes - can also measure important physiological variables under certain conditions.
-In one case, accelerometers revealed the elevated respiratory rate of an emperor penguin (*Aptenodytes forsteri*) following a 27.6 minute dive, the longest recorded for this species [@satoStrokeRatesDiving2011].
+However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables.
+Through careful inspection and analysis of high-resolution acceleration, scientists have measured elevated respiration rates following record-breaking dives [@satoStrokeRatesDiving2011], near-continuous feeding by small cetaceans [@wisniewskaUltraHighForagingRates2016], social interactions between large cetaceans  [@goldbogenUsingAccelerometersDetermine2014], and important biomechanical variables including movement speed [@cadeDeterminingForwardSpeed2018].
 While physio-logging tags with cutting-edge biomedical technologies push the boundaries of physiological field research, simpler IMU tags have fewer logistical constraints and provide access to more species and larger sample sizes.
-This is particularly important for species that cannot be restrained or studied in captivity.
-For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag on one blue whale (*Balaenoptera musculus*) individual [@goldbogenExtremeBradycardiaTachycardia2019].
-Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001] .
-These existing datasets (and future IMU tag deployments) could hold valuable physiological information, but we lack proper computational methods for mining them.
+This is particularly important for species that cannot be restrained or studied in managed care.
+For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (_Balaenoptera musculus_) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999].
+Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001].
+These existing datasets (and future IMU tag deployments) could hold additional valuable physiological information, awaiting proper computational methods for mining them.
 
-The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors.
+The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care.
 Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels.
-The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and phonocardiography.
+The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and echocardiography.
 However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019].
-While the BCG is a three-dimensional phenomenon, it is strongest in the longitudinal (anterior-posterior) axis [@inanBallistocardiographySeismocardiographyReview2015].
-Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010].
-The BCG J wave is analagous to the ECG R wave in that it is the most robust feature in the waveform and typically used for detecting heart beats.
+While the BCG is a three-dimensional phenomenon, it is strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015].
+Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010].
+The BCG J wave is the most robust feature in the waveform and typically used for detecting heart beats.
 
 Here we present a method for generating a BCG from bio-logger accelerometry.
-We validated our method with a simultaneously recorded electrocardiogram (ECG) on a captive killer whale (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
-The relative orientation of the tag on the body is uncertain in cetacean bio-logging [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on longitudinal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
+We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (_Orcinus orca_) and applied it to detect heartrate in a blue whale.
+The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on anterior-posterior acceleration (surge), we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
 Specifically, we tested three hypotheses to validate our method.
@@
 Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG.
-Third, BCG-derived heartrates would increase from the start to end of dives, consistent with the marine mammal cardiophysiological dive response of progressive bradycardia relaxation [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
+Third, BCG-derived heartrates would increase during the bottom and latter phases of dives, consistent with the progressive increase in heartrate routinely observed before and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 
@@
 
-* A captive killer whale was double-tagged with a CATS unit and ECG recorder.
-* The CATS unit was placed behind the left pectoral flipper and the ECG recorder on the ventral center line. 
-* Trainers instructed the whale to hold a submerged position.
-* A blue whale was tagged with a CATS unit (originally published in [@goldbogenWhyWhalesAre2019]).
-* The tag slid behind the right pectoral flipper, where it remained overnight.
-* While tagged, the blue whale engaged in apparent resting behavior.
-We manually labeled motionless periods.
-Specifically, we looked for regions with low amplitude y-axis gyroscope signal, since fluke-strokes are readily apparent in this signal [@goughScalingSwimmingPerformance2019]
-* CATS tags programmed to collect 400 Hz accelerometer, 50 Hz magnetometer, 50 Hz gyroscope, and 10 Hz pressure. Data processed using CATS Toolbox [@cadeToolsIntegratingInertial2021].
-* Relevant permits (IACUC, NMFS, ???)
+**Killer whale**
+
+A killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit.
+The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps.
+All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021].
+The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019].
+Briefly, we recorded ECG at 100 Hz and identified individual heart beats from visually verified R-waves using a customized peak detection program (K.
+Ponganis; Origin 2017, OriginLab Co., Northampton, MA).
+ECG and IMU were recorded during a submerged breath hold.
+
+**Blue whale**
+
+A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]).
+Tag configuration and data processing followed the same procedure as the killer whale, with one addition.
+The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.
+Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to manually labeled motionless periods.
+Motionless periods were identified during or near the bottom phase of dives between fluke strokes.
+Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].
 
@@
 
-The BCG waveform is three dimensional, but strongest in the longitudinal axis [@inanBallistocardiographySeismocardiographyReview2015].
+The BCG waveform is three dimensional, but strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015].
 We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].
-
 For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.
@@
 
-1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) using package `signal` [@R-signal].
-2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter using package `signal`. Differentiation exaggerates impulses like the J wave.
+1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
+2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
 3.
@@
 Additionally, the Shannon entropy is strictly positive, which facilitates peak detection.
-In the 1d case, $k$ is surge only.
+In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
 4. Remove noise by applying a triangular moving average smoother.
@@
 We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios.
-For both BCGs, we calculated the power spectral density using package `psd` [@Barbour2014].
-Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.
+For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014].
+Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified _signal_ as the integration of the power spectral density curve from 4-8 bpm and _noise_ as the integrated remainder, up to 60 bpm.
 
@@
 Dive times were normalized from 0 (start of dive) to 1 (end of dive).
-We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) with package `RobustLinearReg` [@R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.
+We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) [R package `RobustLinearReg`; @R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.
 
@@
 
-The data and code used in this analysis were packaged as a research compendium using package `rrtools` [@marwickPackagingDataAnalytical2018; @rrtools2019].
+The data and code used in this analysis were packaged as a research compendium [R package `rrtools`; @marwickPackagingDataAnalytical2018; @rrtools2019].
 The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.
@@
 BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation).
-The ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
+Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
 
@@
 bw_elg <- bw180905_53_elg %>% 
-  mutate(depth_min = map2_dbl(start, stop, function(t1, t2) {
-    bw180905_53_10hz %>% 
-      filter(between(dt, t1, t2)) %>% 
-      pull(depth) %>% 
-      min()
+  mutate(
+    depth_min = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% min()
     }),
     depth_max = map2_dbl(start, stop, function(t1, t2) {
-    bw180905_53_10hz %>% 
-      filter(between(dt, t1, t2)) %>% 
-      pull(depth) %>% 
-      max()
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% max()
     })
@@
 The 3d BCG (Fig.
-\@ref(fig:bw-bcg)) produced a more robust signal than the surge-only 1d BCG.
-The signal-to-noise ratio was 2.00 for the triaxial BCG, compared to 0.17 for the surge-only BCG (Fig.
-\@ref(fig:psd-plot)).
+\@ref(fig:bw-bcg)) produced a more robust signal than the 1d BCG, which used only anterior-posterior acceleration.
+The signal-to-noise ratio was 2.00 for the 3d BCG, compared to 0.17 for the 1d BCG (Fig.
+\@ref(fig:bw-validation-plots)**A**).
 
@@
                           surface = 2, min_depth = 10, min_dur = 5 * 60),
-    dive_norm = normalize_dives(dive_id)
+    dive_norm = normalize_dives(dive_id),
+    rid_left = approx(bw_elg$start, 
+                      bw_elg$region_id, 
+                      dt, 
+                      "constant")$y,
+    rid_right = approx(bw_elg$stop, 
+                       bw_elg$region_id, 
+                       dt, 
+                       "constant", 
+                       yleft = 0)$y + 1
   )
@@
 3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
-Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig \@ref(fig:brady-relax)).
+Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig.
+\@ref(fig:bw-validation-plots)**B**).
 
@@
 As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors.
+Current IMU tag designs limit BCG analysis to motionless periods, but future dimple- or limpet-style tags could reduce acceleration noise, boost the signal-to-noise ratio, and make the method more widely applicable.
 Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can still derive new insights from traditional sensors.
@@
 * Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
+* Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
+* M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.
 
@@
 
-```{r oo-bcg-ecg, fig.cap="The ECG (A) and BCG (E) produced nearly identical heart beat predictions (red and blue points, respectively).
-B-D display the intermediate steps in the BCG signal processing procedure.
-B: surge after filtering, C: after differencing, D: Shannon entropy.
-Y-axis labeling follows Lee et al.
-(2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
+```{r oo-bcg-ecg, fig.cap="The ECG (**A**) and BCG (**E**), recorded by the ECG and CATS IMU tags, respectively, produced nearly identical heart beat predictions for the killer whale.
+**B-D** display the intermediate steps in the BCG signal processing procedure.
+**B**: Anterior-posterior axis acceleration (surge) after band-pass filtering.
+Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box.
+**C**: Peaks enhanced after forward differencing.
+**D**: A strictly positive signal after calculating Shannon entropy.
+Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
 t <- theme_classic() +
@@
         strip.background = element_blank())
+t_yvoid <- t +
+  theme(axis.text.y = element_blank(),
+        axis.ticks.y = element_blank())
 
@@
 # B: Filtered surge
+ijk_range <- as.POSIXct("2021-08-16 10:09:31.4", tz = "US/Pacific") + c(0, 0.4)
+ijk_complex <- oo_400hz %>% 
+  filter(between(dt, ijk_range[1], ijk_range[2])) %>% 
+  mutate(t = as.numeric(dt - min(dt), unit = "secs"))
+ijk_labels <- slice(ijk_complex, c(41, 53, 67, 85, 110)) %>% 
+  mutate(wave = c("H", "I", "J", "K", "L"))
+pB_inset <- ggplot(ijk_complex, aes(dt, surge_filt)) + 
+  geom_line() +
+  geom_text(aes(label = wave),
+            ijk_labels,
+            vjust = c(-0.2, 1.2, 0.75, -0.6, -0.2),
+            hjust = c(0.5, 0.5, -0.6, 0.5, 0.5),
+            fontface = "bold",
+            size = 3) +
+  theme_void()
+
+inset_range <- as.POSIXct("2021-08-16 10:09:29.9", tz = "US/Pacific") + c(0, 4)
 pB <- ggplot(oo_400hz, aes(dt, surge_filt)) +
   geom_line(size = 0.2) +
+  annotation_custom(grob = ggplotGrob(pB_inset),
+                    xmin = inset_range[1], xmax = inset_range[2],
+                    ymin = 0.25, ymax = 0.5) +
+  annotate("rect",
+           xmin = ijk_range[1], 
+           xmax = ijk_range[2],
+           ymin = min(ijk_complex$surge_filt) - 0.02, 
+           ymax = max(ijk_complex$surge_filt) + 0.02,
+           fill = NA,
+           color = "grey50",
+           lty = 3) +
   scale_x_datetime(date_labels = "%H:%M:%S") +
+  expand_limits(y = 0.5) +
   labs(y = bquote(f[HR])) +
-  t +
-  theme(axis.text.y = element_blank())
+  t_yvoid
 
@@
   labs(y = "dBCG") +
-  t +
-  theme(axis.text.y = element_blank())
+  t_yvoid
 
@@
   labs(y = "seBCG") +
-  t +
-  theme(axis.text.y = element_blank())
+  t_yvoid
 
@@
   labs(y = "Y") +
-  t +
-  theme(axis.text.y = element_blank())
+  t_yvoid
 
@@
                    ncol = 1, 
+                   rel_heights = c(1, 1.5, 0.75, 0.75, 1),
                    labels = "AUTO")
@@
 
-```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG.
-A: filtered triaxial acceleration, B: after differencing, C: Shannon entropy, D: after smoothing.
-Identified heart beats in blue.
-Y-axis labeling follows Lee et al.
-(2016 Sensors) and y-axis values were excluded because the filtering process introduces magnitude distortion; only the relative shape of the signal is relevant to the analysis."}
+```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG during a motionless period in a blue whale dive.
+**A**: Band-pass filtered triaxial acceleration, where surge is along the anterior-posterior axis, sway is along the lateral axis, and heave is along the dorso-ventral axis.
+**B**: Differencing the filtered acceleration enhances peaks.
+**C**: Calculating the Shannon entropy combines information from all three axes and makes the signal strictly positive.
+**D**: Smoothing the Shannon entropy facilitates robust peak detection.
+Detected heart beats in blue.
+Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values were excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
 bw_bcg_plot <- function(rid) {
   bcg_sample <- filter(bw_400hz, region_id == rid)
-
-  t <- theme_classic() +
-  theme(axis.title.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_blank())
 
@@
     labs(y = bquote(f[HR])) +
-    t +
-    theme(axis.text.y = element_blank())
+    t_yvoid
 
@@
     labs(y = "dBCG") +
-    t +
-    theme(axis.text.y = element_blank())
+    t_yvoid
 
@@
     labs(y = "seBCG") +
-    t +
-    theme(axis.text.y = element_blank())
+    t_yvoid
 
@@
     labs(y = "Y") +
-    t +
-    theme(axis.text.y = element_blank())
+    t_yvoid
 
@@
 
-```{r psd-plot, fig.cap="Signal-to-noise ratio was higher for the triaxial BCG (lower panel, 2.00) than the surge-only BCG (upper panel, 0.17).
+```{r bw-validation-plots, fig.cap="**A** Signal-to-noise ratio was higher for the 3d BCG (lower panel) than the 1d BCG (anterior-posterior acceleration only; upper panel).
 Each panel shows the power spectral density for the BCG.
 Based on previously observed blue whale heart rates, 4-8 bpm was considered signal (gray shading).
-The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal band to the area under the rest of the curve, up to 60 bpm."}
+The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal band to the area under the rest of the curve, up to 60 bpm.
+**B** Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
+Bradycardia is greatest at the start of the dive (~4-5 bpm), relaxing towards the end (~8-9 bpm).
+Points indicate instantaneous heart rates and the line is a Theil-Sen regression."}
 labels <- bw_psd %>% 
@@
   mutate(s2n_text = sprintf("Signal:noise = %0.2f", s2n))
-ggplot(bw_psd, aes(x = freq_bpm, y = spec)) +
+pA <- ggplot(bw_psd, aes(x = freq_bpm, y = spec)) +
   geom_area(data = filter(bw_psd, between(freq_bpm, 4, 8)), fill = "grey80") +
@@
         strip.text = element_blank())
-```
 
-```{r brady-relax, fig.cap="Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
-Bradycardia is greatest at the start of the dive (~4-5 bpm), relaxing towards the end (~8-9 bpm).
-Points indicate instantaneous heart rates and the line is a Theil-Sen regression."}
-ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
+pB <- ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
   geom_point() +
@@
   theme_classic()
+
+cowplot::plot_grid(pA, pB, nrow = 1, labels = "AUTO")
 ```

### 2021-10-27 analysis/supplementary-materials/supplemental.Rmd
Incorporate co-author feedback

---
title: "Supplementary Materials"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,
  fig.height = 6,
  fig.width = 6,
  dpi = 300
)
library(cetaceanbcg)
library(tidyverse)
```

```{r motionless}
bw_elg <- bw180905_53_elg %>% 
  mutate(
    depth_min = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% min()
    }),
    depth_max = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% max()
    }),
    ygyr_min = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% min()
    }),
    ygyr_max = map2_dbl(start, stop, function(t1, t2) {
      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% max()
    })
  )
bw10hz <- bw180905_53_10hz %>%
  mutate(
    rid_left = approx(bw_elg$start, 
                      bw_elg$region_id, 
                      dt, 
                      "constant")$y,
    rid_right = approx(bw_elg$stop, 
                       bw_elg$region_id, 
                       dt, 
                       "constant", 
                       yleft = 0)$y + 1,
    motion_class = case_when(
      depth < 2 ~ "surface",
      rid_left == rid_right ~ "motionless",
      TRUE ~ "active"),
    dive_id = split_dives(dt, depth,
                          surface = 2, min_depth = 10, min_dur = 5 * 60)
  )
```

```{r gyr_entropy, fig.cap="Motionless periods were manually identified by looking for periods of low amplitude rotational veloctiy ($\\omega$) around the lateral axis, indicative of cetaceans' dorso-ventral fluking motion."}
divesample <- filter(bw10hz, dive_id == 1)
diveelg <- bw_elg %>% 
  filter(start >= first(divesample$dt), stop <= last(divesample$dt))

depth_profile <- ggplot(divesample, aes(dt, depth)) +
  geom_line() +
  geom_rect(aes(xmin = start, xmax = stop, ymin = depth_min, ymax = depth_max),
            diveelg,
            inherit.aes = FALSE,
            fill = "orchid1",
            color = NA,
            alpha = 0.5) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  scale_y_reverse("Depth (m)") +
  theme_classic() +
  theme(axis.title.x = element_blank())
ygyr_profile <- ggplot(divesample, aes(dt, gw[, 2])) +
  geom_line() +
  geom_rect(aes(xmin = start, xmax = stop, ymin = ygyr_min, ymax = ygyr_max),
            diveelg,
            inherit.aes = FALSE,
            fill = "orchid1",
            color = NA,
            alpha = 0.5) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  scale_y_continuous(bquote(omega~(rad~s^{-1}))) +
  theme_classic() +
  theme(axis.title.x = element_blank())
cowplot::plot_grid(depth_profile, ygyr_profile, ncol = 1, align = "v")
```

### 2021-10-27 data-raw/bw180905_53_10hz.R

@@ -9,12 +9,13 @@ bw180905_53_mat <- R.matlab::readMat("analysis/data/raw_data/bw180905-53.mat")
 bw180905_53_10hz <- tibble(
   dn = as.vector(bw180905_53_mat$DN10Hz),
   depth = as.vector(bw180905_53_mat$depth10Hz),
-  aw = as.matrix(bw180905_53_mat$Aw10Hz)
+  aw = as.matrix(bw180905_53_mat$Aw10Hz),
+  gw = as.matrix(bw180905_53_mat$Gw10Hz)
 ) %>%
   mutate(dt = as.POSIXct((dn - 719529) * 86400,
                          origin = "1970-01-01",
                          tz = "UTC") %>%
            lubridate::force_tz(tz)) %>%
-  select(dt, depth, aw)
+  select(dt, depth, aw, gw)
 
 usethis::use_data(bw180905_53_10hz, overwrite = TRUE)

### 2021-10-29 analysis/supplementary-materials/supplemental.Rmd
Create supplemental

@@
 ---
-title: "Supplementary Materials"
+title: "Supplemental material for:"
+subtitle: "An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: pdf_document
+output:
+  bookdown::pdf_document2:
+    fig_caption: yes
+    toc: false
+header-includes:
+  - \renewcommand{\figurename}{Figure S}
+  - \makeatletter
+  - \def\fnum@figure{\figurename\thefigure}
+  - \makeatother
 ---
@@
 knitr::opts_chunk$set(
-  echo=FALSE,
-  fig.height = 6,
-  fig.width = 6,
+  echo = FALSE,
+  cache = TRUE,
+  message = FALSE,
+  fig.height = 5,
+  fig.width = 5,
   dpi = 300
@@
 
-```{r motionless}
+```{r bw-data}
 bw_elg <- bw180905_53_elg %>% 
@@
     }),
+    depth_med = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% 
+        first() %>% 
+        median()
+    }),
     ygyr_min = map2_dbl(start, stop, function(t1, t2) {
@@
       bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% max()
-    })
+    }),
+    ygyr_med = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% 
+        first() %>% 
+        median()
+    }),
   )
-bw10hz <- bw180905_53_10hz %>%
+
+bw_10hz <- bw180905_53_10hz %>% 
+  mutate(rid_left = approx(bw_elg$start, 
+                           bw_elg$region_id, 
+                           dt, 
+                           "constant")$y,
+         rid_right = approx(bw_elg$stop, 
+                            bw_elg$region_id, 
+                            dt, 
+                            "constant", 
+                            yleft = 0)$y + 1,
+         region_id = ifelse(rid_left == rid_right, rid_left, NA),
+         dive_id = split_dives(dt, depth,
+                          surface = 2, min_depth = 10, min_dur = 5 * 60))
+
+bw_400hz <- bw180905_53_400hz %>%
   mutate(
+    across(surge:heave, 
+           filter_acc, fs = 400, upper = 10.0, 
+           .names = "{.col}_filt"),
+    jerk = jerk(cbind(surge_filt, sway_filt, heave_filt), 
+                fs = 400, p = 4, n = 2 * 400 + 1),
+    jerk_se = shannon_entropy(jerk),
+    jerk_smooth = tma(jerk_se, 2 * 400),
+    # Annotate regions
     rid_left = approx(bw_elg$start, 
@@
                        yleft = 0)$y + 1,
-    motion_class = case_when(
-      depth < 2 ~ "surface",
-      rid_left == rid_right ~ "motionless",
-      TRUE ~ "active"),
-    dive_id = split_dives(dt, depth,
-                          surface = 2, min_depth = 10, min_dur = 5 * 60)
+    region_id = ifelse(rid_left == rid_right, rid_left, NA),
+    # Zero-out signal in non-valid regions (i.e. remove movement artifacts)
+    jerk_smooth = ifelse(is.na(region_id), 0, jerk_smooth)
   )
@@
 
-```{r gyr_entropy, fig.cap="Motionless periods were manually identified by looking for periods of low amplitude rotational veloctiy ($\\omega$) around the lateral axis, indicative of cetaceans' dorso-ventral fluking motion."}
-divesample <- filter(bw10hz, dive_id == 1)
+```{r peak-heur, fig.cap="**A:** Minor peaks (hollow points) in the ballistocardiogram (BCG, line) were not considered heart beats.
+Only major peaks (solid points) were retained for analysis.
+The BCG for one motionless period shown here.
+**B:** We used peak height and prominence (i.e.
+height relative to the contour surrounding a higher peak) to heuristically differentiate major and minor peaks.
+For each peak, we calculated the Euclidean distance (in height-prominence space) to the highest peak overall.
+The peaks in **A** shown here in height-prominence space.
+**C:** The distance to the highest peak exhibited a bimodal distribution.
+We chose a distance threshold (dashed red line) corresponding to the valley in the density curve.
+**D:** All peaks found in the BCG across all motionless periods in height-prominence space.
+Solid and hollow points as in **A**.
+The dashed red curve corresponds to the distance threshold in **C**."}
+fs_hz <- 400
+minperiod <- 2
+cue <- bw_400hz$jerk_smooth
+peaks <- pracma::findpeaks(cue, minpeakdistance = fs_hz * minperiod)
+proms <- peak_prominences(cue, peaks[, 2])
+
+dist_max <- sqrt((max(peaks[, 1]) - peaks[, 1])^2 + (max(proms) - proms)^2)
+dist_dens <- stats::density(dist_max)
+dist_thr <- dist_dens$x[pracma::findpeaks(-dist_dens$y)[, 2]]
+is_beat <- peaks[, 2][dist_max <= dist_thr]
+beats <- logical(length(cue))
+beats[is_beat] <- TRUE
+
+peak_tbl <- tibble(dt = bw_400hz$dt[peaks[, 2]],
+                   height = peaks[, 1],
+                   prominence = proms,
+                   is_beat = dist_max <= dist_thr)
+
+regionsample <- filter(bw_400hz, region_id == 6)
+peaksample <- filter(peak_tbl, between(dt, 
+                                       first(regionsample$dt), 
+                                       last(regionsample$dt))) %>% 
+  arrange(dt) %>% 
+  mutate(peak_lbl = seq(n()))
+
+max_height <- max(peaks[, 1])
+thr_curve <- tibble(
+  t = seq(0, 2 * pi, length.out = 1000),
+  height = max_height + cos(t) * dist_thr,
+  prominence = max_height + sin(t) * dist_thr,
+) %>% 
+  filter(height <= max_height, prominence <= max_height)
+
+pA <- ggplot(regionsample, aes(dt, jerk_smooth)) +
+  geom_line() +
+  geom_point(aes(y = height, shape = is_beat), data = peaksample) +
+  geom_text(aes(y = height, label = peak_lbl), 
+            data = peaksample, 
+            vjust = -0.3) +
+  scale_shape_manual(values = c(21, 16)) +
+  labs(y = "Y") +
+  theme_classic(base_size = 10) +
+  theme(axis.title.x = element_blank(),
+        legend.position = "FALSE")
+
+pB <- ggplot(peaksample, aes(height, prominence, shape = is_beat)) +
+  geom_point() +
+  ggrepel::geom_text_repel(aes(label = peak_lbl), force = 3) +
+  scale_shape_manual(values = c(21, 16)) +
+  scale_x_continuous(n.breaks = 4) +
+  labs(x = "Peak height",
+       y = "Peak prominence") +
+  coord_fixed() +
+  theme_classic(base_size = 10) +
+  theme(legend.position = "FALSE")
+
+pC <- ggplot(peak_tbl, aes(dist_max)) +
+  geom_density() +
+  geom_vline(xintercept = dist_thr, linetype = 2, color = "red") +
+  scale_x_continuous(n.breaks = 4) +
+  labs(x = "Distance from highest peak") +
+  theme_classic(base_size = 10)
+
+pD <- ggplot(peak_tbl, aes(height, prominence)) +
+  geom_point(aes(shape = is_beat)) +
+  geom_line(data = thr_curve, linetype = 2, color = "red") +
+  scale_shape_manual(values = c(21, 16)) +
+  labs(x = "Peak height",
+       y = "Peak prominence") +
+  coord_fixed() +
+  theme_classic(base_size = 10) +
+  theme(legend.position = "FALSE")
+
+cowplot::plot_grid(pA, pB, pC, pD, ncol = 2, labels = "AUTO")
+```
+
+```{r ygyr, fig.cap="Motionless periods were manually identified by looking for periods of low amplitude rotational veloctiy ($\\omega$) around the lateral axis, indicative of cetaceans' dorso-ventral fluking motion.
+The depth and $\\omega$ profiles for the first dive shown here, with motionless periods in pink."}
+divesample <- filter(bw_10hz, dive_id == 1)
+norm_y <- function(ymin, ymax, sign) {
+  (ymin + ymax) / 2 + sign * max(ymax - ymin) / 2
+}
 diveelg <- bw_elg %>% 
-  filter(start >= first(divesample$dt), stop <= last(divesample$dt))
+  filter(start >= first(divesample$dt), stop <= last(divesample$dt)) %>% 
+  mutate(depth_min2 = norm_y(depth_min, depth_max, -1),
+         depth_max2 = norm_y(depth_min, depth_max, 1),
+         ygyr_min2 = norm_y(ygyr_min, ygyr_max, -1),
+         ygyr_max2 = norm_y(ygyr_min, ygyr_max, 1))
 
@@
   geom_line() +
-  geom_rect(aes(xmin = start, xmax = stop, ymin = depth_min, ymax = depth_max),
+  geom_rect(aes(xmin = start, 
+                xmax = stop, 
+                ymin = depth_min2, 
+                ymax = depth_max2),
             diveelg,
@@
   scale_y_reverse("Depth (m)") +
-  theme_classic() +
+  theme_classic(base_size = 10) +
   theme(axis.title.x = element_blank())
@@
   geom_line() +
-  geom_rect(aes(xmin = start, xmax = stop, ymin = ygyr_min, ymax = ygyr_max),
+  geom_rect(aes(xmin = start, xmax = stop, ymin = ygyr_min2, ymax = ygyr_max2),
             diveelg,
@@
   scale_y_continuous(bquote(omega~(rad~s^{-1}))) +
-  theme_classic() +
+  theme_classic(base_size = 10) +
   theme(axis.title.x = element_blank())
-cowplot::plot_grid(depth_profile, ygyr_profile, ncol = 1, align = "v")
+cowplot::plot_grid(depth_profile, ygyr_profile, 
+                   ncol = 1, 
+                   align = "v", 
+                   labels = "AUTO")
 ```
+
+```{r bw-profile, fig.cap="Two hours of blue whale tag data were used for ballistocardiographic analysis.
+During these hours, the animal made repeated, shallow dives.
+Because the BCG can be obscured by dynamic body movements, only relatively motionless periods were retained for analysis (pink).
+These periods were identified by visual examination of the rotational velocity around the lateral axis."}
+ggplot(bw_10hz, aes(dt, depth)) +
+  geom_line(size = 0.5) +
+  geom_line(aes(group = region_id), 
+            data = drop_na(bw_10hz, region_id), 
+            size = 0.5, 
+            color = "orchid1") +
+  scale_x_datetime(date_labels = "%H:%M:%S") +
+  scale_y_reverse("Depth (m)") +
+  theme_classic(base_size = 10) +
+  theme(axis.title.x = element_blank())
+```

### 2021-10-29 analysis/interact/global.R
Remove extraneous files

@@ -1,211 +1,0 @@
-library(cetaceanbcg)
-library(shiny)
-library(shinybusy)
-library(tidyverse)
-
-filter_thr0 <- c(1.5, 7.5)
-filter_order0 <- 4
-window0 <- 1.8
-yaw_offset0 <- 0
-
-n_pts <- 1e3
-bw180905_53_10hz_sparse <- bw180905_53_10hz[floor(seq(1,
-                                                      nrow(bw180905_53_10hz),
-                                                      length.out = n_pts)), ]
-
-update_results <- function(hyperparams) {
-  show_modal_progress_circle(.10, "Filtering acceleration")
-
-  window <- hyperparams$window
-  fs <- 400
-  fny <- fs / 2
-  acc_filter <- signal::butter(hyperparams$filter_order,
-                               hyperparams$filter_thr / fny,
-                               type = "pass")
-  apply_filter <- function(x) signal::filtfilt(acc_filter, x)
-
-  rid_fun <- function(x, xout) {
-    approx(x, bw180905_53_elg$region_id, xout, "constant", yleft = 0)$y
-  }
-  bw_400hz <- bw180905_53_400hz %>%
-    mutate(
-      across(c(surge, sway, heave), apply_filter, .names = "{.col}_filt"),
-      rid_left = rid_fun(bw180905_53_elg$start, dt),
-      rid_right = rid_fun(bw180905_53_elg$stop, dt) + 1,
-      region_id = ifelse(rid_left == rid_right, rid_left, NA),
-      jerk = njerk(cbind(surge_filt, sway_filt, heave_filt), fs),
-      jerk_smooth = tma(jerk, window * fs)
-    )  %>%
-    select(-c(rid_left, rid_right))
-
-  update_modal_progress(.40, "Finding heartbeats")
-
-  bw_400hz_beats <- bw_400hz %>%
-    filter(!is.na(region_id)) %>%
-    group_by(region_id) %>%
-    mutate(surge_range = local_range(surge_filt, window * fs),
-           bcg_beat = find_beats_jerk(jerk_smooth, surge_filt, window, fs),
-           bcg_bpm = bpm(bcg_beat, dt)) %>%
-    ungroup() %>%
-    select(dt, surge_range, bcg_beat, bcg_bpm)
-
[…]

### 2021-10-29 analysis/interact/server.R

@@ -1,51 +1,0 @@
-server <- function(input, output, session) {
-  # Hyperparameters
-  hyperparams <- reactive(list(
-    filter_thr = c(input$filt_thr_lower, input$filt_thr_upper),
-    filter_order = as.integer(input$filt_order),
-    window = input$window,
-    yaw_offset = input$yaw_offset * pi / 180
-  ))
-
-  # Data
-  bw_data <- reactiveVal(update_results(list(
-    filter_thr = filter_thr0,
-    filter_order = filter_order0,
-    window = window0,
-    yaw_offset = yaw_offset0
-  )))
-
-  # I/O
-  depth_click <- reactiveValues(x = NULL, y = NULL)
-  dive_click <- reactiveValues(x = NULL, y = NULL)
-  observeEvent(input$update,
-               bw_data(update_results(hyperparams())))
-  observeEvent(input$depth_click, {
-    depth_click$x <- convert_clickx(input$depth_click$x)
-    depth_click$y <- input$depth_click$y
-    dive_click$x <- NULL
-    dive_click$y <- NULL
-  })
-  observeEvent(input$dive_click, {
-    dive_click$x <- convert_clickx(input$dive_click$x)
-    dive_click$y <- input$dive_click$y
-  })
-
-  # Plots
-  output$depth_plot <- renderPlot(
-    ggplot(bw180905_53_10hz_sparse, aes(dt, depth)) +
-      geom_line(size = 0.2) +
-      annotate_click(depth_click$x, depth_click$y) +
-      scale_x_datetime(date_labels = "%H:%M:%S") +
-      scale_y_reverse() +
-      theme_minimal()
-  )
-  # Why do I have to do bw_data()[2]$bw_10hz? No clue why bw_data()$bw_10hz
-  # doesn't work
-  output$dive_plot <- renderPlot(
-    bw_dive_plot(bw_data()[2]$bw_10hz, depth_click, dive_click)
-  )
-  output$bcg_plot <- renderPlot(
-    bw_bcg_plot(bw_data()[1]$bw_400hz, dive_click)
-  )
-}

### 2021-10-29 analysis/interact/ui.R

@@ -1,44 +1,0 @@
-ui <- fillPage(
-
-  fillRow(
-
-    fillCol(
-      h1("Tune BCG Hyperparameters"),
-      numericInput("filt_thr_lower",
-                   "Filter lower threshold",
-                   value = filter_thr0[1],
-                   min = 0, max = 50),
-      numericInput("filt_thr_upper",
-                   "Filter upper threshold",
-                   value = filter_thr0[2],
-                   min = 0, max = 50),
-      numericInput("filt_order",
-                   "Filter order",
-                   value = filter_order0,
-                   min = 1, max = 40,
-                   step = 1),
-      numericInput("window_size",
-                   "Window size (seconds)",
-                   value = window0,
-                   min = 0.25, max = 10),
-      numericInput("yaw_offset",
-                   "Yaw offset (degrees)",
-                   value = yaw_offset0,
-                   min = -180, max = 180),
-      actionButton("update",
-                   "Update")
-    ),
-
-    fillCol(
-      plotOutput("depth_plot", click = "depth_click", height = "100%"),
-      plotOutput("dive_plot", click = "dive_click", height = "100%"),
-      plotOutput("bcg_plot", height = "100%"),
-      flex = c(1, 1, 2)
-    ),
-
-    flex = c(1, 3)
-  ),
-
-  padding = 10
-
-)

### 2021-10-29 analysis/scratchpad/bcg_s2n.R

@@ -1,110 +1,0 @@
-library(cetaceanbcg)
-library(tidyverse)
-
-bcg_psd_results <- readRDS("analysis/data/derived_data/bcg_psd_results.rds")
-n <- as.integer(readLines("analysis/data/derived_data/psd_par_n.txt"))
-
-# "Signal" band (typical blue whale diving heartrates)
-bw_bpm <- c(4, 8) / 60
-
-# Calculate signal-to-noise ratio by integrating the PSD curve
-calc_s2n <- function(freq, spec) {
-  is_signal <- between(freq, bw_bpm[1], bw_bpm[2])
-  is_noise <- !is_signal
-
-  psignal = pracma::trapz(freq[is_signal], spec[is_signal])
-  pnoise = pracma::trapz(freq[is_noise], spec[is_noise])
-
-  psignal / pnoise
-}
-
-# Ranked signal-to-noise result by bandpass parameters
-bcg_s2n <- bcg_psd_results %>%
-  drop_na(norm_fun) %>%
-  group_by(low, width, norm_fun) %>%
-  summarize(s2n = calc_s2n(freq, spec), .groups = "drop") %>%
-  arrange(-s2n)
-
-# Parameter space, covered by sampling. Use interpolation to fill in gaps.
-low <- 2^seq(-2, 2, length.out = n)
-width <- 2^seq(1, 5, length.out = n)
-interp_s2n <- akima::interp(
-  x = log2(bcg_s2n$low[bcg_s2n$norm_fun == "jerk"]),
-  y = log2(bcg_s2n$width[bcg_s2n$norm_fun == "jerk"]),
-  z = bcg_s2n$s2n[bcg_s2n$norm_fun == "jerk"],
-  xo = log2(low),
-  yo = log2(width),
-  linear = FALSE,
-  extrap = FALSE
-)
-bcg_s2n_interp <- expand_grid(low, width) %>%
-  mutate(i = match(log2(low), interp_s2n$x),
-         j = match(log2(width), interp_s2n$y),
-         s2n = interp_s2n$z[cbind(i, j)]) %>%
-  select(-i, -j)
-
-interp_s2n <- akima::interp(
-  x = bcg_s2n$low[bcg_s2n$norm_fun == "jerk"],
-  y = bcg_s2n$width[bcg_s2n$norm_fun == "jerk"],
-  z = bcg_s2n$s2n[bcg_s2n$norm_fun == "jerk"],
-  xo = low,
[…]

### 2021-10-29 analysis/scratchpad/create_bw_400hz.R

@@ -1,39 +1,0 @@
-library(cetaceanbcg)
-library(tidyverse)
-
-fs <- 400
-fny <- fs / 2
-window <- 1.8 # s
-bw_elg <- bw180905_53_elg
-
-bandpass_filter <- function(x, pass) {
-  signal::butter(4, pass / fny, type = "pass") %>%
-    signal::filtfilt(x)
-}
-
-# Apply filters as with killer whale data, but narrower band
-bw_400hz <- bw180905_53_400hz %>%
-  mutate(
-    # Pass-band filter acceleration
-    across(surge:heave,
-           bandpass_filter, pass = c(1.0, 10.0),
-           .names = "{.col}_filt"),
-    # Calculate jerk norm and apply smoother
-    njerk = njerk(cbind(surge_filt, sway_filt, heave_filt),
-                  p = 4, n = window * fs),
-    njerk_smooth = tma(njerk, window * fs),
-    # Annotate regions
-    rid_left = approx(bw_elg$start, bw_elg$region_id, dt, "constant")$y,
-    rid_right = approx(bw_elg$stop, bw_elg$region_id, dt, "constant", yleft = 0)$y + 1,
-    region_id = ifelse(rid_left == rid_right, rid_left, NA),
-    # Find heart beats using jerk cue
-    surge_range = local_range(surge_filt, window * fs),
-    bcg_beat = find_beats_jerk(njerk_smooth, !is.na(region_id), window, fs)
-  ) %>%
-  # Calculate heart rate within each region
-  group_by(region_id) %>%
-  mutate(bcg_bpm = bpm(bcg_beat, dt)) %>%
-  ungroup() %>%
-  select(-c(rid_left, rid_right))
-
-saveRDS(bw_400hz, "analysis/data/derived_data/bw_400hz.rds")

### 2021-10-29 analysis/scratchpad/divesegmentation.R

@@ -1,81 +1,0 @@
-split_dives <- function(dt, depth, surface = 0, min_depth = 0, min_dur = 0) {
-  dive_start <- depth >= surface & lag(depth) < surface
-  dive_start[1] <- FALSE
-  dive_id <- cumsum(dive_start)
-  dive_id[depth < surface | dive_id == 0] <- NA
-  dive_id <- factor(dive_id)
-  dive_depth <- tapply(depth, dive_id, max)
-  dive_dur <- tapply(dt,
-                     dive_id,
-                     function(x) as.numeric(max(x) - min(x), unit = "secs"))
-  not_dives <- which(dive_depth < min_depth | dive_dur < min_dur)
-  dive_id[dive_id %in% not_dives] <- NA
-  dive_id <- droplevels(dive_id)
-  levels(dive_id) <- seq_along(levels(dive_id))
-  dive_id
-}
-
-segment_phases <- function(depth, dive_id) {
-  by_dive <- split(depth, dive_id) %>%
-    lapply(function(x) {
-      i <- seq_along(x)
-      depth_lm <- lm(x ~ i)
-      depth_seg <- segmented::segmented(depth_lm, seg.Z = ~i, npsi = 2)
-      psi <- floor(depth_seg$psi[, "Est."])
-      result <- character(length(x))
-      result[1:psi[1]] <- "descent"
-      result[psi[1]:psi[2]] <- "bottom"
-      result[psi[2]:length(x)] <- "ascent"
-      result
-    })
-  phases_all <- rep(NA, length(dive_id))
-  for (dive in names(by_dive)) {
-    phases_all[!is.na(dive_id) & dive_id == dive] <- by_dive[[dive]]
-  }
-  factor(phases_all, levels = c("descent", "bottom", "ascent"))
-}
-
-normalize_dive <- function(dive_phase, dive_id) {
-  by_dive <- split(dive_phase, dive_id) %>%
-    lapply(function(x) {
-      result <- rep(NA, length(x))
-      b <- which(x == "bottom")
-      result[x == "bottom"] <- (b - min(b)) / (max(b) - min(b))
-      result
-    })
-  all_bottom <- rep(NA, length(dive_id))
-  for (dive in names(by_dive)) {
-    all_bottom[!is.na(dive_id) & dive_id == dive] <- by_dive[[dive]]
-  }
[…]

### 2021-10-29 analysis/scratchpad/psd.R

@@ -1,123 +1,0 @@
-
-
-# Signal band within these bounds
-bw_bpm <- c(4, 8) / 60
-
-bcg_psd <- function(low, width, bw400hz,
-                    fs = 400, window = 1.8, bpm = c(4, 8)) {
-  pass <- c(low, low + width)
-  fny <- fs / 2
-  bandpass_filter <- function(x) {
-    signal::butter(4, pass / fny, type = "pass") %>%
-      signal::filtfilt(x)
-  }
-  acc <- as.matrix(bw400hz[, c("surge", "heave", "sway")])
-  acc_filt <- apply(acc, 2, bandpass_filter)
-
-  if (sum(is.finite(acc_filt)) == 0) {
-    return(tibble())
-  }
-
-  surge_filt <- acc[, 1]
-  lsr <- local_range(surge_filt, window * fs)
-  j = njerk(acc_filt, p = 4, n = window * fs)
-  njerk_smooth = tma(j, window * fs)
-
-  # PSD of surge, LSR, and jerk
-  psds <- list(surge_filt, lsr, njerk_smooth) %>%
-    map(~ ts(.x[!is.na(bw400hz$region_id)], frequency = 400)) %>%
-    map(~ suppressMessages(psd::pspectrum(.x)))
-
-  freq <- psds[[1]]$freq
-
-  tibble(freq,
-         surge = psds[[1]]$spec,
-         lsr = psds[[2]]$spec,
-         jerk = psds[[3]]$spec) %>%
-    pivot_longer(surge:jerk,
-                 names_to = "norm_fun",
-                 values_to = "spec")
-}
-
-# Latin hypercube sampling
-low <- 2^seq(-2, 2, length.out = 256)
-width <- 2^seq(1, 5, length.out = 256)
-bcg_lhs <- tibble(
-  low = sample(low, length(low), replace = FALSE),
-  width = sample(width, length(width), replace = FALSE)
-)
-pb <- progress::progress_bar$new(total = nrow(bcg_lhs))
-bcg_psd_results <- bcg_lhs %>%
-  group_by_all() %>%
-  group_modify(function(x, y) {
-    pb$tick()
-    bcg_psd(y$low, y$width, bw_400hz)
-  }) %>%
-  ungroup()
-
-calc_s2n <- function(freq, spec) {
-  is_signal <- between(freq, bw_bpm[1], bw_bpm[2])
-  is_noise <- !is_signal
-
[…]

### 2021-10-29 analysis/scratchpad/psd_par.R

@@ -1,90 +1,0 @@
-library(cetaceanbcg)
-library(doParallel)
-library(tidyverse)
-
-n <- as.integer(commandArgs(trailingOnly = TRUE))
-
-# use the environment variable SLURM_NTASKS_PER_NODE to set
-# the number of cores to use
-if (Sys.getenv("SLURM_NTASKS_PER_NODE") == "") {
-  ncores <- 1
-} else {
-  ncores <- Sys.getenv("SLURM_NTASKS_PER_NODE")
-}
-registerDoParallel(cores = ncores)
-
-bw_400hz <- readRDS("analysis/data/derived_data/bw_400hz.rds")
-
-# Signal band within these bounds
-bw_bpm <- c(4, 8) / 60
-
-bcg_psd <- function(low, width, bw400hz,
-                    fs = 400, window = 1.8, bpm = c(4, 8), max_bpm = 64) {
-  tryCatch({
-    pass <- c(low, low + width)
-    fny <- fs / 2
-    bandpass_filter <- function(x) {
-      signal::butter(4, pass / fny, type = "pass") %>%
-        signal::filtfilt(x)
-    }
-    acc <- as.matrix(bw400hz[, c("surge", "heave", "sway")])
-    acc_filt <- apply(acc, 2, bandpass_filter)
-
-    if (sum(is.finite(acc_filt)) == 0) {
-      return(tibble())
-    }
-
-    surge_filt <- acc[, 1]
-    lsr <- local_range(surge_filt, window * fs)
-    j = njerk(acc_filt, p = 4, n = window * fs)
-    njerk_smooth = tma(j, window * fs)
-
-    # PSD of surge, LSR, and jerk
-    psds <- list(surge_filt, lsr, njerk_smooth) %>%
-      map(~ ts(.x[!is.na(bw400hz$region_id)], frequency = 400)) %>%
-      map(~ suppressMessages(psd::pspectrum(.x)))
-
-    freq <- psds[[1]]$freq
-
-    tibble(freq,
-           surge = psds[[1]]$spec,
-           lsr = psds[[2]]$spec,
-           jerk = psds[[3]]$spec) %>%
-      pivot_longer(surge:jerk,
-                   names_to = "norm_fun",
-                   values_to = "spec") %>%
-      filter(freq * 60 <= max_bpm)
-  }, error = function(e) {
-    cat(sprintf("low = %0.3f; width = %0.3f; error = \"%s\"",
[…]

### 2021-10-29 analysis/paper/paper.Rmd
Update paper with supplemental references

@@
 The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.
-Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to manually labeled motionless periods.
-Motionless periods were identified during or near the bottom phase of dives between fluke strokes.
-Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].
 
@@
 4. Remove noise by applying a triangular moving average smoother.
-5. Extract peaks and heuristically remove noisy peaks (see supplemental).
+5. Extract peaks and heuristically remove noisy peaks (Fig. S1).
 
@@
 ## BCG application to blue whale
+
+Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to motionless periods (Fig.
+S2).
+These periods occured during or near the bottom phase of dives between fluke strokes.
+Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].
 
@@
 
-We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig.
-\@ref(fig:bw-profile)).
+We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. S3).
 
@@
 
-```{r bw-profile, fig.cap="Blue whale dive profile. Motionless periods indicated by pink boxes."}
-ggplot(bw180905_53_10hz, aes(dt, depth)) +
-  geom_line(size = 0.5) +
-  geom_rect(aes(xmin = start, xmax = stop, ymin = depth_min, ymax = depth_max),
-            bw_elg,
-            inherit.aes = FALSE,
-            fill = "orchid1",
-            color = NA,
-            alpha = 0.5) +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  scale_y_reverse("Depth (m)") +
-  theme_classic() +
-  theme(axis.title.x = element_blank())
-```
-
 ```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG during a motionless period in a blue whale dive.

### 2021-11-08 analysis/paper/paper.Rmd
Draft v2

@@
 Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram.
-We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and progressive relaxation of bradycardia during dives.
+We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and increase in heart rate prior to and during ascent.
 Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
@@
 This is particularly important for species that cannot be restrained or studied in managed care.
-For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (_Balaenoptera musculus_) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999].
+For example, of the sixteen species of baleen whales (Mysticeti), heart rate has only been recorded with an electrocardiogram tag in the wild for one blue whale (*Balaenoptera musculus*) [@goldbogenExtremeBradycardiaTachycardia2019; but see @ponganisHeartRateElectrocardiogram1999].
 Conversely, IMU tags have been deployed on hundreds of individuals of nearly every species in the clade for the last twenty years [@nowacekBuoyantBalaenidsUps2001].
@@
 However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019].
-While the BCG is a three-dimensional phenomenon, it is strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015].
+While the BCG is a three-dimensional phenomenon, it is strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015].
 Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010].
@@
 Here we present a method for generating a BCG from bio-logger accelerometry.
-We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (_Orcinus orca_) and applied it to detect heartrate in a blue whale.
-The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on anterior-posterior acceleration (surge), we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
+We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
+The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
 Specifically, we tested three hypotheses to validate our method.
@@
 Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG.
-Third, BCG-derived heartrates would increase during the bottom and latter phases of dives, consistent with the progressive increase in heartrate routinely observed before and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
+Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 
@@
 
-A killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit.
+A 3807 kg (**needs confirmation**) killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit.
 The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps.
 All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021].
+This rotation aligned the tag's x-, y-, and z- axes with the cranio-caudal, lateral, and dorso-ventral axes of the whale, respectively.
 The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019].
-Briefly, we recorded ECG at 100 Hz and identified individual heart beats from visually verified R-waves using a customized peak detection program (K.
+Briefly, we attached the tag on the mid-lateral left chest posterior to the pectoral fin and recorded ECG at 100 Hz.
+Individual heart beats were identified from visually verified R-waves using a customized peak detection program (K.
 Ponganis; Origin 2017, OriginLab Co., Northampton, MA).
-ECG and IMU were recorded during a submerged breath hold.
+ECG and IMU were recorded during a spontaneous breath hold while the whale rested at the surface.
 
@@
 
-The BCG waveform is three dimensional, but strongest in the anterior-posterior axis [@inanBallistocardiographySeismocardiographyReview2015].
-We tested 1d and 3d metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].
+The BCG waveform is three dimensional, but strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015].
+We tested both 1-dimensional (cranio-caudal only) and 3-dimensional metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].
 For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.
@@
 Additionally, the Shannon entropy is strictly positive, which facilitates peak detection.
-In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
+In the 1-dimensional case, $k$ is surge (cranio-caudal acceleration) only.
 4.  Remove noise by applying a triangular moving average smoother.
@@
 
-This procedure can be applied to either 1d (i.e., surge-only) or 3d acceleration. In the case of 3d acceleration, the band-pass and Savitzky-Golay filters were applied to each axis independently.
+This procedure may be applied to either 1-dimensional (i.e., cranio-caudal only) or 3-dimensional acceleration.
+In the case of 3-dimensional acceleration, the band-pass and Savitzky-Golay filters were applied to each axis independently.
 
-## ECG validation on killer whale
+## BCG validation with killer whale ECG
 
 We fit ordinary least squares regression to BCG-derived instantaneous heart rates with respect to ECG-derived and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1.
-We calculated the mean and standard deviation of absolute error as an equivalence measure (1d BCG only).
+We calculated the mean and standard deviation of absolute error as an equivalence measure (1-dimensional BCG only).
 
@@
 These periods occured during or near the bottom phase of dives between fluke strokes.
-Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].
+Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [*sensu* @goughScalingSwimmingPerformance2019].
 
-We tested whether the 3d BCG was more robust than 1d BCG in field data by comparing the signal-to-noise ratios.
+We tested whether the 3-dimensional BCG was more robust than 1-dimensional BCG in field data by comparing the signal-to-noise ratios.
 For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014].
-Previously recorded blue whale apneic heart rate was 4-8 bpm [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified _signal_ as the integration of the power spectral density curve from 4-8 bpm and _noise_ as the integrated remainder, up to 60 bpm.
+Previously recorded blue whale apneic heart rate was 4-8 beats per minute (bpm) [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.
 
-We also tested whether BCG-derived instantaneous heart rates exhibited bradycardia relaxation over the course of dives, consistent with diving physiology patterns in marine mammals [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
+We also tested whether BCG-derived instantaneous heart rates were consistent with the range and pattern of heart rates previously observed in the blue whale and other marine mammals; namely a gradual increase in heart rate later in the dive, especially during the final ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 We assigned dive start and end times when the whale swam deeper than 2 m, retaining dives that exceeded 10 m depth and 5 minutes duration.
@@
 
-## ECG validation on killer whale
+## BCG validation with killer whale ECG
 
@@
 We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold.
-BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean &#177; standard deviation).
-Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean &#177; standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
+BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean ± standard deviation).
+Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean ± standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
 
@@
 
-We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. S3).
+We generated 1-dimensional and 3-dimensional BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig.
+S3).
 
@@
 
-The 3d BCG (Fig.
-\@ref(fig:bw-bcg)) produced a more robust signal than the 1d BCG, which used only anterior-posterior acceleration.
-The signal-to-noise ratio was 2.00 for the 3d BCG, compared to 0.17 for the 1d BCG (Fig.
+The 3-dimensional BCG (Fig.
+\@ref(fig:bw-bcg)) produced a more robust signal than the 1-dimensional BCG, which used only cranio-caudal acceleration.
+The signal-to-noise ratio was 2.00 for the 3-dimensional BCG, compared to 0.17 for the 1-dimensional BCG (Fig.
 \@ref(fig:bw-validation-plots)**A**).
@@
 
-3d BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
+3-dimensional BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
 Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig.
@@
 
-* Everyone who helped collect and process the blue whale data.
-* The Sea World trainers.
-* Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
-* Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
-* M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.
+-   Everyone who helped collect and process the blue whale data.
+-   The Sea World trainers.
+-   Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
+-   Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
+-   M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.
 
@@
 
-```{r oo-bcg-ecg, fig.cap="The ECG (**A**) and BCG (**E**), recorded by the ECG and CATS IMU tags, respectively, produced nearly identical heart beat predictions for the killer whale.
+```{r oo-bcg-ecg, fig.cap="The ECG (**A**, recorded by ECG tag) and 1-dimensional BCG (**E**, processed from the cranio-caudal acceleration recorded by the IMU tag) produced nearly identical heart beat predictions for the killer whale.
 **B-D** display the intermediate steps in the BCG signal processing procedure.
-**B**: Anterior-posterior axis acceleration (surge) after band-pass filtering.
+**B**: Cranio-caudal axis acceleration after band-pass filtering.
 Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box.
-**C**: Peaks enhanced after forward differencing.
+**C**: Peaks enhanced after forward differencing acceleration (i.e., jerk).
 **D**: A strictly positive signal after calculating Shannon entropy.
-Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
+Y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
 t <- theme_classic() +
@@
   expand_limits(y = 0.5) +
-  labs(y = bquote(f[HR])) +
+  labs(y = "Cranio-caudal acceleration (filtered)") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "dBCG") +
+  labs(y = "Cranio-caudal jerk") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "seBCG") +
+  labs(y = "Shannon entropy") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "Y") +
+  labs(y = "BCG") +
   t_yvoid
@@
 
-```{r bw-bcg, fig.cap="Example of signal processing for 3d BCG during a motionless period in a blue whale dive.
-**A**: Band-pass filtered triaxial acceleration, where surge is along the anterior-posterior axis, sway is along the lateral axis, and heave is along the dorso-ventral axis.
+```{r bw-bcg, fig.cap="Example of signal processing for 3-dimensional BCG during a motionless period in a blue whale dive.
+**A**: Band-pass filtered triaxial acceleration, where surge is along the cranio-caudal axis, sway is along the lateral axis, and heave is along the dorso-ventral axis.
 **B**: Differencing the filtered acceleration enhances peaks.
@@
   pA <- bcg_sample %>% 
-    rename(Surge = surge_filt, Sway = sway_filt, Heave = heave_filt) %>% 
-    pivot_longer(Surge:Heave, names_to = "axis", values_to = "acc") %>% 
-    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
+    rename(`Cranio-caudal` = surge_filt, 
+           Lateral = sway_filt, 
+           `Dorso-ventral` = heave_filt) %>% 
+    pivot_longer(`Cranio-caudal`:`Dorso-ventral`, 
+                 names_to = "axis", 
+                 values_to = "acc") %>% 
+    mutate(axis = factor(axis, levels = c("Cranio-caudal", 
+                                          "Lateral", 
+                                          "Dorso-ventral"))) %>% 
     ggplot(aes(dt, acc)) +
@@
     facet_grid(rows = vars(axis)) +
-    labs(y = bquote(f[HR])) +
+    labs(y = "Acceleration (filtered)") +
     t_yvoid
@@
   pB <- bcg_sample %>% 
-    mutate(jerk = set_names(as_tibble(jerk), c("Surge", "Sway", "Heave"))) %>% 
+    mutate(jerk = set_names(as_tibble(jerk), c("Cranio-caudal", 
+                                               "Lateral", 
+                                               "Dorso-ventral"))) %>% 
     unpack(jerk) %>% 
-    pivot_longer(Surge:Heave, names_to = "axis", values_to = "jerk") %>% 
-    mutate(axis = factor(axis, levels = c("Surge", "Sway", "Heave"))) %>% 
+    pivot_longer(`Cranio-caudal`:`Dorso-ventral`, 
+                 names_to = "axis", 
+                 values_to = "jerk") %>% 
+    mutate(axis = factor(axis, levels = c("Cranio-caudal", 
+                                          "Lateral", 
+                                          "Dorso-ventral"))) %>% 
     ggplot(aes(dt, jerk)) +
@@
     facet_grid(rows = vars(axis)) +
-    labs(y = "dBCG") +
+    labs(y = "Jerk") +
     t_yvoid
@@
     scale_x_datetime(date_labels = "%H:%M:%S") +
-    labs(y = "seBCG") +
+    labs(y = "Shannon entropy") +
     t_yvoid
@@
     scale_x_datetime(date_labels = "%H:%M:%S") +
-    labs(y = "Y") +
+    labs(y = "BCG") +
     t_yvoid
@@
 
-```{r bw-validation-plots, fig.cap="**A** Signal-to-noise ratio was higher for the 3d BCG (lower panel) than the 1d BCG (anterior-posterior acceleration only; upper panel).
+```{r bw-validation-plots, fig.cap="**A** Signal-to-noise ratio was higher for the 3-dimensional BCG (lower panel) than the 1-dimensional BCG (cranio-caudal acceleration only; upper panel).
 Each panel shows the power spectral density for the BCG.
@@
 The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal band to the area under the rest of the curve, up to 60 bpm.
-**B** Heart rates observed in the 3d BCG followed characteristic diving physiology patterns.
-Bradycardia is greatest at the start of the dive (~4-5 bpm), relaxing towards the end (~8-9 bpm).
-Points indicate instantaneous heart rates and the line is a Theil-Sen regression."}
+**B** Heart rates observed in the 3-dimensional BCG followed characteristic diving physiology patterns.
+Heart rate is lowest at the start of the dive (~4-5 bpm), increasing towards ascent (~8-9 bpm).
+Points indicate instantaneous heart rates and the line is a Theil-Sen regression.
+Outliers likely represent premature beats which are common in heart rate profiles during dives of cetaceans, seals, and penguins [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014; @andrews1997; @wright2014]."}
 labels <- bw_psd %>% 
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 

### 2021-11-08 analysis/paper/paper.Rmd
Draft v3

@@
 A 3807 kg (**needs confirmation**) killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit.
+We attached the CATS tag on the mid-lateral left chest posterior to the pectoral fin (Movie S1).
 The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps.
@@
 The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019].
-Briefly, we attached the tag on the mid-lateral left chest posterior to the pectoral fin and recorded ECG at 100 Hz.
+Briefly, the tag was attached approximately midline on the ventral chest just caudal (posterior) to the axilla and we recorded the ECG at 100 Hz.
 Individual heart beats were identified from visually verified R-waves using a customized peak detection program (K.
@@
 A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]).
-Tag configuration and data processing followed the same procedure as the killer whale, with one addition.
-The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.
+The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
+Tag configuration and data processing followed the same procedure as the killer whale.
+The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**).
+We downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.
 
@@
 \@ref(fig:oo-bcg-ecg)).
-We collected 14 s of simultaneous ECG and BCG data during a motionless, submerged breath hold.
+We collected 14 s of simultaneous ECG and BCG data during a motionless breath hold at the surface.
 BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean ± standard deviation).
@@
 -   Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
--   Funding from Office of Naval Research N000141912455, Stanford Terman Fellowship
--   M. F. Czapanskiy was supported by the Stanford University William R. and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.
+
+# Footnotes
+
+## Author contributions
+
+Conceptualization: M.F.C.,J.A.F.,P.J.P.,J.A.G.; Methodology: M.F.C.,J.A.F.,P.J.P.; Software: M.F.C.; Formal analysis: M.F.C.,P.J.P.; Investigation: M.F.C.,J.A.F.,P.J.P.; Resources: P.J.P.,J.A.G.; Writing - original draft: M.F.C; Writing - review & editing: M.F.C.,J.A.F.,P.J.P.,J.A.G.; Supervision: P.J.P.,J.A.G.; Project administration: P.J.P.,J.A.G.; Funding acquisition: J.A.G.
+
+## Funding
+
+This work was supported by grant N000141912455 from the Office of Naval Research.
+M.F.C.
+was supported by the Stanford University William R.
+and Sara Hart Kimball Fellowship and a Stanford Data Science Scholar Fellowship.
+
+## Data availability
+
+All data and code used in this analysis are available on Zenodo (DOI needed).
+
+## Competing interests
+
+The authors declare no competing interests.
 
@@
   theme(axis.text.y = element_blank(),
-        axis.ticks.y = element_blank())
+        axis.ticks.y = element_blank(),
+        strip.text = element_blank())
 
@@
   expand_limits(y = 0.5) +
-  labs(y = "Cranio-caudal acceleration (filtered)") +
+  labs(y = "Acceleration\n(filtered)") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "Cranio-caudal jerk") +
+  labs(y = "Jerk") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "Shannon entropy") +
+  labs(y = "Shannon\nentropy") +
   t_yvoid
@@
 ```{r bw-bcg, fig.cap="Example of signal processing for 3-dimensional BCG during a motionless period in a blue whale dive.
-**A**: Band-pass filtered triaxial acceleration, where surge is along the cranio-caudal axis, sway is along the lateral axis, and heave is along the dorso-ventral axis.
-**B**: Differencing the filtered acceleration enhances peaks.
-**C**: Calculating the Shannon entropy combines information from all three axes and makes the signal strictly positive.
+**A**: Band-pass filtered triaxial acceleration, with cranio-caudal in orange, lateral in blue, and dorso-ventral in green.
+**B**: Peaks enhanced after forward differencing acceleration (i.e., jerk).
+**C**: The Shannon entropy combines information from all three axes and makes the signal strictly positive.
 **D**: Smoothing the Shannon entropy facilitates robust peak detection.
 Detected heart beats in blue.
-Y-axis labeling follows [@leePhysiologicalSignalMonitoring2016]; y-axis values were excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
+Y-axis values excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
 bw_bcg_plot <- function(rid) {
   bcg_sample <- filter(bw_400hz, region_id == rid)
+  palette <- rgb(c(230, 86, 0), c(159, 180, 158), c(0, 233, 115), maxColorValue = 255)
 
-  # A: Filtered surge
-  pA <- bcg_sample %>% 
+  # A: Filtered acceleration
+  acceleration_data <- bcg_sample %>% 
     rename(`Cranio-caudal` = surge_filt, 
@@
                                           "Lateral", 
-                                          "Dorso-ventral"))) %>% 
-    ggplot(aes(dt, acc)) +
+                                          "Dorso-ventral")))
+  pA <- ggplot(acceleration_data, aes(dt, acc, color = axis)) +
     geom_line(size = 0.2) +
@@
     scale_y_continuous(n.breaks = 3) +
+    scale_color_manual(values = palette) +
     facet_grid(rows = vars(axis)) +
-    labs(y = "Acceleration (filtered)") +
-    t_yvoid
+    labs(y = "Acceleration\n(filtered)") +
+    t_yvoid +
+    theme(legend.position = "none",
+          panel.spacing = unit(0, "lines"))
 
   # B: Jerk vectors
-  pB <- bcg_sample %>% 
+  jerk_data <- bcg_sample %>% 
     mutate(jerk = set_names(as_tibble(jerk), c("Cranio-caudal", 
@@
                                           "Lateral", 
-                                          "Dorso-ventral"))) %>% 
-    ggplot(aes(dt, jerk)) +
+                                          "Dorso-ventral")))
+  pB <-  ggplot(jerk_data, aes(dt, jerk, color = axis)) +
     geom_line(size = 0.2) +
@@
     scale_y_continuous(n.breaks = 3) +
+    scale_color_manual(values = palette) +
     facet_grid(rows = vars(axis)) +
     labs(y = "Jerk") +
-    t_yvoid
+    t_yvoid +
+    theme(legend.position = "none",
+          panel.spacing = unit(0, "lines"))
 
@@
     scale_x_datetime(date_labels = "%H:%M:%S") +
-    labs(y = "Shannon entropy") +
+    labs(y = "Shannon\nentropy") +
     t_yvoid
@@
                      ncol = 1, 
+                     rel_heights = c(1.5, 1.5, 1, 1),
                      labels = "AUTO")
@@
 Points indicate instantaneous heart rates and the line is a Theil-Sen regression.
-Outliers likely represent premature beats which are common in heart rate profiles during dives of cetaceans, seals, and penguins [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014; @andrews1997; @wright2014]."}
+Outliers likely represent premature beats which are common in heart rate profiles during dives of cetaceans, pinnipeds, and penguins [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014; @andrews1997; @wright2014]."}
 labels <- bw_psd %>% 

### 2021-11-08 analysis/supplementary-materials/supplemental.Rmd
Draft v3

@@
 
+<!--
+ABD: Use color instead of open/close
+C: Axis tick label clipping
+D: open and close doesn't work because of overplotting
+-->
 ```{r peak-heur, fig.cap="**A:** Minor peaks (hollow points) in the ballistocardiogram (BCG, line) were not considered heart beats.
@@
 
+<!-- James' feedback: either make this wider (not square) or break it into separate hours. -->
 ```{r bw-profile, fig.cap="Two hours of blue whale tag data were used for ballistocardiographic analysis.

### 2021-11-30 analysis/paper/paper.Rmd
Incorporating Todd's feedback

@@
   - Scripps: Scripps Institution of Oceanography, University of California San Diego, USA
-  - SeaWorld: Veterinary Services, SeaWorld of California, USA
+  - SeaWorld: Animal Health Department, SeaWorld of California, USA
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021].
+Recent advances in physio-logging (recording physiological variables using animal-borne devices) have largely been driven by new developments in sensor technology [@hawkesIntroductionThemeIssue2021; @williams2021].
 For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021].
@@
 The BCG originated as a clinical tool in the first half of the 20th century [@starrStudiesEstimationCardiac1939], but was largely superseded by electro- and echocardiography.
-However the medical community has recently returned to ballistocardiography as a potential means of passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography], which has led to substantial progress in signal processing methodology for generating interpretable BCGs [@sadekBallistocardiogramSignalProcessing2019].
+However, potential novel applications like passive monitoring of heart function in at-risk populations [@giovangrandi2011ballistocardiography] has led to a recent resurgence of ballistocardiography research, with advances in hardware [@andreozzi2021] and signal processing methodology [@sadekBallistocardiogramSignalProcessing2019].
 While the BCG is a three-dimensional phenomenon, it is strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015].
@@
 
-A 3807 kg (**needs confirmation**) killer whale in managed care at the SeaWorld, San Diego, CA marine facilities was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld display permit.
+A 3868 kg adult female killer whale in managed care at SeaWorld of California, San Diego, CA was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld USDA APHIS display permit.
 We attached the CATS tag on the mid-lateral left chest posterior to the pectoral fin (Movie S1).
@@
 
-A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]).
+A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 [previously published by @goughScalingSwimmingPerformance2019].
 The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
@@
 
-1.  Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
+1.  Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) (R package `signal`) [@R-signal].
 2.  Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
@@
 We tested whether the 3-dimensional BCG was more robust than 1-dimensional BCG in field data by comparing the signal-to-noise ratios.
-For both BCGs, we calculated the power spectral density [R package `psd`; @Barbour2014].
+For both BCGs, we calculated the power spectral density (R package `psd`) [@Barbour2014].
 Previously recorded blue whale apneic heart rate was 4-8 beats per minute (bpm) [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.
@@
 Dive times were normalized from 0 (start of dive) to 1 (end of dive).
-We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) [R package `RobustLinearReg`; @R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.
+We regressed instantaneous heart rate against normalized dive time using robust Theil-Sen regression (to account for heteroscedascity) (R package `RobustLinearReg`) [@R-RobustLinearReg; @Sen-1968; @theilRankInvariantMethodLinear1992] and tested whether the slope was greater than 0.
 
@@
 
-The data and code used in this analysis were packaged as a research compendium [R package `rrtools`; @marwickPackagingDataAnalytical2018; @rrtools2019].
+The data and code used in this analysis were packaged as a research compendium (R package `rrtools`) [@marwickPackagingDataAnalytical2018; @rrtools2019].
 The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.
@@
 
--   Everyone who helped collect and process the blue whale data.
--   The Sea World trainers.
--   Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the open science R community for developing tools and educational resources that facilitate open science practices.
+The authors are grateful to the SeaWorld of California Killer Whale training staff for their efforts and support.
+We also thank Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the R community for developing tools and educational resources to facilitate open science practices.
 

### 2021-12-03 analysis/paper/paper.Rmd
Fixes #1
Remove section numbering

@@
       fig_caption: yes
+      number_sections: no
       reference_docx: "../templates/template.docx"
@@
 We also thank Anna Krystalli, Ben Marwick, Karthik Ram, Nicholas Tierney, and other members of the R community for developing tools and educational resources to facilitate open science practices.
+This is a SeaWorld Parks and Entertainment Technical Contribution number 2021-12.
 

### 2022-02-02 README.Rmd
Remove Binder link
Address reviewer feedback

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/FlukeAndFeather/cetaceanbcg/main?urlpath=rstudio)
+[![DOI](https://zenodo.org/badge/399999855.svg)](https://zenodo.org/badge/latestdoi/399999855)
 
@@
 
-## How to run in your broswer or download and run locally
+## How to download and run locally
 

### 2022-02-02 analysis/paper/paper.Rmd
Address reviewer feedback

@@
 Here we present a computational method for extracting heartrate from high-resolution accelerometer data using a ballistocardiogram.
-We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the method recovers previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and increase in heart rate prior to and during ascent.
+We validated our method with simultaneous accelerometer-electrocardiogram tag deployments in a controlled setting on a killer whale (*Orcinus orca*) and demonstrate the predictions correspond with previously observed cardiovascular patterns in a blue whale (*Balaenoptera musculus*), including the magnitude of apneic bradycardia and increase in heart rate prior to and during ascent.
 Our ballistocardiogram method may be applied to mine heart rates from previously collected accelerometery and expand our understanding of comparative cardiovascular physiology.
@@
 For example, new physio-logging tags can detect regional changes in blood flow by incorporating functional near-infrared spectroscopy sensors [@mcknightShiningNewLight2021].
-However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables.
+However, traditional inertial measurement unit (IMU) tags equipped with accelerometers and other inertial sensors can also measure important physiological and related variables, such as wing beat frequency [@patterson2019] and feeding rate [@divirgilio2018].
 Through careful inspection and analysis of high-resolution acceleration, scientists have measured elevated respiration rates following record-breaking dives [@satoStrokeRatesDiving2011], near-continuous feeding by small cetaceans [@wisniewskaUltraHighForagingRates2016], social interactions between large cetaceans [@goldbogenUsingAccelerometersDetermine2014], and important biomechanical variables including movement speed [@cadeDeterminingForwardSpeed2018].
@@
 
-The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care.
+The ballistocardiogram (BCG) has potential applications to using accelerometers as heartrate monitors in both the wild and in managed care [@giovangrandi2011ballistocardiography; @sadekBallistocardiogramSignalProcessing2019; @inanBallistocardiographySeismocardiographyReview2015].
 Ballistocardiography is a noninvasive method for measuring cardiac function based on the ballistic forces involved in the heart ejecting blood into the major vessels.
@@
 While the BCG is a three-dimensional phenomenon, it is strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015].
-Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these are the IJK complex, which progressively occurs during systole [@pinheiroTheoryDevelopmentsUnobtrusive2010].
-The BCG J wave is the most robust feature in the waveform and typically used for detecting heart beats.
+Along this axis, the waveform is composed of multiple peaks and valleys; most prominent of these is the so-called IJK complex [@pinheiroTheoryDevelopmentsUnobtrusive2010].
+The precise physiological mechanism underlying the BCG waveform has not been fully resolved [@kim2016], but it has been established that the IJK complex occurs during systole and, in humans, occurs at approximately the same time as the T-wave in an electrocardiogram (ECG) [@inanBallistocardiographySeismocardiographyReview2015].
+The BCG J wave is the most robust feature in the waveform and is typically used for detecting heart beats [@inanBallistocardiographySeismocardiographyReview2015].
 
 Here we present a method for generating a BCG from bio-logger accelerometry.
-We validated our method with a simultaneously recorded electrocardiogram (ECG) on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
-The relative orientation of the tag on the body is uncertain in cetacean bio-logging in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
+We validated our method with a simultaneously recorded ECG on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
+The relative orientation of the tag on the body is often uncertain when bio-loggers are deployed in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
 Specifically, we tested three hypotheses to validate our method.
-First, a one-dimensional BCG would, in a controlled setting, produce statistically equivalent instantaneous heartrates as an ECG.
+First, a one-dimensional BCG would, in a controlled setting, produce instantaneous heartrates that are statistically equivalent to ECG instantaneous heartrates.
 Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG.
@@
 
-A 3868 kg adult female killer whale in managed care at SeaWorld of California, San Diego, CA was double-tagged with Customized Animal Tracking Solutions IMU (CATS, www.cats.is) and electrocardiogram (ECG) tags on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld USDA APHIS display permit.
+A 3868 kg adult female killer whale in managed care at SeaWorld of California, San Diego, CA was double-tagged with an archival Customized Animal Tracking Solutions IMU (CATS, www.cats.is) tag and a custom-built, archival ECG tag on August 16, 2021 as part of clinical animal cardiac evaluations under the SeaWorld USDA APHIS display permit.
+The ECG tag hardware and data processing procedures were previously described by @bickettHeartRatesHeart2019.
+Both tags were deployed by hand and attached with suction cups.
 We attached the CATS tag on the mid-lateral left chest posterior to the pectoral fin (Movie S1).
-The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps.
+The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps (*TODO: get tag 97 accelerometer accuracy/precision from Will G)*.
 All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021].
 This rotation aligned the tag's x-, y-, and z- axes with the cranio-caudal, lateral, and dorso-ventral axes of the whale, respectively.
-The ECG tag hardware and data processing followed the methods in [@bickettHeartRatesHeart2019].
-Briefly, the tag was attached approximately midline on the ventral chest just caudal (posterior) to the axilla and we recorded the ECG at 100 Hz.
-Individual heart beats were identified from visually verified R-waves using a customized peak detection program (K.
+We attached the ECG tag approximately midline on the ventral chest just caudal (posterior) to the axilla and we recorded the ECG at 100 Hz.
+Individual heart beats in the ECG record were identified from visually verified R-waves using a customized peak detection program (K.
 Ponganis; Origin 2017, OriginLab Co., Northampton, MA).
@@
 
-A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 [previously published by @goughScalingSwimmingPerformance2019].
-The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
-Tag configuration and data processing followed the same procedure as the killer whale.
+A 24.5 m blue whale was tagged with an archival, suction-cup CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 [previously published by @goughScalingSwimmingPerformance2019].
+We deployed the tag using a 4 m fiberglass pole from a 6.3 m rigid-hulled inflatable boat [as described by @goldbogen2006] The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
+Tag configuration and data processing followed the same procedure as the killer whale, including sampling rates for inertial sensors and video.
 The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**).
-We downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.
+We downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools [@cadeToolsIntegratingInertial2021].
 
@@
 The BCG waveform is three dimensional, but strongest in the cranio-caudal axis [@inanBallistocardiographySeismocardiographyReview2015].
-We tested both 1-dimensional (cranio-caudal only) and 3-dimensional metrics for identifying heartbeats in acceleration data based on the methods of [@leePhysiologicalSignalMonitoring2016].
-For windowed operations, we used 0.5 s for killer whale data and 2.0 s for blue whale data.
+We tested both 1-dimensional (cranio-caudal only) and 3-dimensional metrics for identifying heartbeats in acceleration data based on the methods of @leePhysiologicalSignalMonitoring2016.
+For windowed operations (such as moving averages and signal filters), we used 0.5 s windows for killer whale data and 2.0 s windows for blue whale data, corresponding to 200 and 800 data points, respectively.
+The different window sizes were determined through trial and error to remove noise while retaining signal shape.
+Generally, longer windows will be necessary for larger animals due to their slower heartrates [@Stahl1967b].
 
@@
 
-1.  Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) (R package `signal`) [@R-signal].
-2.  Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
+1.
+First we removed noise and de-trended the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25 Hz], blue whale: [1-10 Hz]) (R package `signal`) [@R-signal].
+The lower cut-off frequency de-trended the data.
+1 Hz should be appropriate for most marine mammal species.
+The upper cut-off frequency removed noise.
+A suitable upper cut-off frequency will depend on body size; larger species' bodies produce lower magnitude accelerations [@martínlópez2021], so more conservative upper cut-off frequencies may be applied to remove more noise without sacrificing signal shape clarity.
+2.
+Then we enhanced the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`).
+Differentiation (i.e., $a_{t+1}-a_t$, where $a_t$ is the observed acceleration at time step $t$) exaggerates peaks, like the J wave, but it is sensitive to noisy signals.
+Therefore, the signal should be de-noised prior to differentiation.
+A moving average smoother can remove noise, but it would also reduce the amplitude of peaks.
+Hence, differentiating Savitzky-Golay filters are preferred in peak-detection algorithms because they remove noise while retaining the general shape of peaks [@samann2019].
+We described the resulting signal as "differenced acceleration", rather than jerk, because we did not take the derivative of acceleration with respect to time.
+The purpose of this signal was to exaggerate a phenomenon in the signal (i.e., the J wave), not to describe a physical quantity (i.e., jerk).
 3.
-Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis).
-Additionally, the Shannon entropy is strictly positive, which facilitates peak detection.
-In the 1-dimensional case, $k$ is surge (cranio-caudal acceleration) only.
-4.  Remove noise by applying a triangular moving average smoother.
-5.  Extract peaks and heuristically remove noisy peaks (Fig. S1).
+We further enhanced the peaks in the differenced acceleration signal by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis).
+Additionally, Shannon entropy is strictly positive, which facilitated peak detection.
+In the 1D BCG, $k$ was surge (cranio-caudal acceleration).
+In the 3D BCG, $k$ included surge, sway (lateral acceleration), and heave (dorso-ventral acceleration).
+4.
+After enhancing the peaks through differentiation and entropy calculation, we removed residual noise by applying a triangular moving average (TMA) smoother.
+TMAs are equivalent to applying a simple moving average in two passes, which applies greater weight to the middle part of the window and retains peaks and valleys more clearly.
+After steps 2 and 3, the signal was clear enough that TMAs provided satisfactory results, obviating the need for a more complex algorithm like a Savitzky-Golay filter at this stage.
+We described the resulting signal as the BCG.
+5.
+The BCG contained major peaks (corresponding to heartbeats) and minor peaks (residual noise) (Fig S1**A**).
+We extracted all peaks from the BCG and applied a clustering algorithm to retain major peaks and reject minor peaks.
+First, we extracted all peaks in the BCG signal using `findpeaks()` (R package `pracma`) [@R-pracma] with a minimum peak distance equivalent to the window size (0.5 s for the killer whale, 2.0 s for the blue whale).
+For each peak, we calculated its absolute height and its prominence (i.e., height relative to the lowest valley between a peak and its higher neighbors).
+Then, we calculated each peak's Euclidean distance in height-prominence space from the highest peak (Fig S1**B**) and estimated the density distribution of these distances (Fig S1**C**).
+The density distribution was bimodal, with a low-distance peak corresponding to major peaks and a high-distance peak corresponding to minor peaks.
+We used the distance corresponding to the valley between the two peaks as a threshold for rejecting minor peaks (Fig S1**D**).
 
@@
 
-We fit ordinary least squares regression to BCG-derived instantaneous heart rates with respect to ECG-derived and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1.
+We fit ordinary least squares regression to BCG-derived instantaneous heartrates with respect to ECG-derived heartrate and tested 1) if the intercept was significantly different than 0 and 2) if the slope was significantly different than 1.
 We calculated the mean and standard deviation of absolute error as an equivalence measure (1-dimensional BCG only).
@@
 Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to motionless periods (Fig.
-S2).
-These periods occured during or near the bottom phase of dives between fluke strokes.
+\@ref(fig:bw-bcg)**B-C**).
+These periods occurred during or near the bottom phase of dives between fluke strokes.
 Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [*sensu* @goughScalingSwimmingPerformance2019].
+We used gyroscopes for stroke detection because they were 1) separate sensors from the accelerometers and 2) strokes are clearly visible in gyroscope signals and are robust to tag placement.
 
@@
 Previously recorded blue whale apneic heart rate was 4-8 beats per minute (bpm) [@goldbogenExtremeBradycardiaTachycardia2019], so we quantified *signal* as the integration of the power spectral density curve from 4-8 bpm and *noise* as the integrated remainder, up to 60 bpm.
+The sample size recorded by @goldbogenExtremeBradycardiaTachycardia2019 was one individual, so we could not account for potential inter-individual variation.
+Nonetheless, 4-8 bpm was the best available estimate for typical blue whale apneic heart rate.
 
@@
 
-The data and code used in this analysis were packaged as a research compendium (R package `rrtools`) [@marwickPackagingDataAnalytical2018; @rrtools2019].
-The research compendium was written as an R package so other researchers can read, run, and modify the methods described here.
+The data and code used in this analysis were packaged as a research compendium, containing the data, code, and an executable version of this manuscript.
+We used the R package `rrtools` [@marwickPackagingDataAnalytical2018; @rrtools2019] to initialize the compendium, which was written as an R package.
+This approach promotes reproducibility and facilitates adoption by other researchers [@alstonBeginnerGuideConducting2021; @powersOpenScienceReproducibility2019; @stoddenEmpiricalAnalysisJournal2018].
+The steps described in section **Signal processing** were implemented as functions in the R package, and the executable manuscript demonstrates how to use those functions to perform the analyses presented in this study.
 
@@
 We collected 14 s of simultaneous ECG and BCG data during a motionless breath hold at the surface.
+Logistical constraints prevented us from gathering a longer sample, as these data were collected secondary to other projects.
 BCG-derived instantaneous heart rates were within `r sprintf("%0.1f%% \u00B1 %0.1f%%", mean_bpm_diff * 100, sd_bpm_diff * 100)` of the ECG-derived rates (mean ± standard deviation).
 Ordinary least squares regression of BCG heartrates on ECG heartrates yielded a slope of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[2], bpm_stderr[2])` and intercept of `r sprintf("%0.2f \u00B1 %0.2f", coef(bpm_lm)[1], bpm_stderr[1])` (mean ± standard error), which were not significantly different from the hypothesized 1 and 0, respectively.
+(Fig.
+\@ref(fig:validation-plots)**C**).
 
@@
 We generated 1-dimensional and 3-dimensional BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig.
-S3).
+\@ref(fig:bw-bcg)**A-C**).
 
@@
 The 3-dimensional BCG (Fig.
-\@ref(fig:bw-bcg)) produced a more robust signal than the 1-dimensional BCG, which used only cranio-caudal acceleration.
+\@ref(fig:bw-bcg)) produced a more robust signal (i.e., higher signal-to-noise ratio) than the 1-dimensional BCG, which used only cranio-caudal acceleration.
 The signal-to-noise ratio was 2.00 for the 3-dimensional BCG, compared to 0.17 for the 1-dimensional BCG (Fig.
-\@ref(fig:bw-validation-plots)**A**).
+\@ref(fig:validation-plots)**A**).
+Although the power spectral density curve for the 1-dimensional BCG had a peak in the 4-8 bpm frequency range, most of the signal's power was concentrated in lower frequencies.
+Conversely, the 3-dimensional BCG's power was concentrated precisely in the 4-8 bpm frequency range, with only a smaller peak in the lower frequencies.
 
@@
 Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig.
-\@ref(fig:bw-validation-plots)**B**).
+\@ref(fig:validation-plots)**B**).
 
-## Reproducibility
+## Limitations and considerations for future applications
 
-The research compendium containing data, code, and an executable version of this manuscript was archived on Zenodo (TODO).
-We developed the research compendium as an R package to facilitate investigation and adoption by other researchers.
-Publishing data and code in standardized formats (such as an R package) is a critical step towards transparency and computational reproducibility [@alstonBeginnerGuideConducting2021; @powersOpenScienceReproducibility2019; @stoddenEmpiricalAnalysisJournal2018].
+While the BCG method presented here holds the potential to mine existing and future marine mammal bio-logging datasets for information about cardiovascular function, it has several limitations compared to ECG methods.
+Most importantly, BCGs are highly sensitive to movement artifacts [@inanBallistocardiographySeismocardiographyReview2015], so only motionless periods are valid for analysis.
+This limits the behavioral and physiological contexts in which heartrate may be measured.
+For example, the BCG is unlikely to quantify neither the magnitude of surface tachycardia [@goldbogenExtremeBradycardiaTachycardia2019] nor exercise modulation of bradycardia [@noren2012], due to movement artifacts during those activities.
+Secondarily, we did not test whether the BCG is robust to tag placement location.
+The blue whale data presented in this study was collected when a dorsally-deployed tag slipped to the lateral chest cavity behind a flipper, where it is reasonable to expect greater accelerations caused by heart beats than from a tag farther from the animal's center of mass.
+It is possible that the ballistic forces generated by heart beats are strong enough to produce an interpretable BCG for a variety of potential tag deployment locations, but this likely varies with animal body size, as well as accelerometer sampling rate and sensitivity.
+
+When auditing existing bio-logging data and planning future tag deployments for BCG analysis, careful consideration should be paid to sampling rate.
+As a rule of thumb in signal processing, the sampling rate should be at least twice the frequency of the phenomenon of interest.
+In the case of the BCG, the relevant frequency is that of the BCG waveform, *not* the heartrate.
+In humans, the power of the IJK-complex (the part of the BCG waveform used for heart beat detection) is concentrated between 4-7 Hz [@moukadem2018].
+It is unlikely that marine mammal BCG waveforms have a higher frequency than humans, owing to their generally larger body sizes.
+Therefore, it is possible that BCGs may be generated for accelerometer sampling rates as low as 10-15 Hz.
+Conservatively, the authors recommend a sampling rate of no less than 50 Hz (i.e., twice the upper cut-off frequency of the widest bandpass filter used in this study).
+
+Future bio-logging BCG methodology research should address the limitations imposed by tag-placement and movement artifacts.
+We used accelerometers in this study because of their prevalence in bio-logging research, but it is possible that other widely used bio-logger sensors, such as gyroscopes and/or magnetometers, could produce a clearer signal in a greater variety of contexts.
+Alternative bio-logger housing designs, such as limpet-style tags [@andrews2008] or "marine skin" [@nassar2018], could reduce noise, boost the signal-to-noise ratio, and make the method more widely applicable.
 
@@
 As accelerometer tags have been deployed on many cetacean species for multiple decades, this method may be applied to mine existing datasets and better understand how heartrate scales with body size and other biological factors.
-Current IMU tag designs limit BCG analysis to motionless periods, but future dimple- or limpet-style tags could reduce acceleration noise, boost the signal-to-noise ratio, and make the method more widely applicable.
-Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can still derive new insights from traditional sensors.
+It may also provide additional data for conservation physiology applications.
+For example, BCGs extracted from gliding phases before and after controlled sonar exposure experiments could quantify the physiological response to anthropogenic disturbance [@southall2019].
+Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can derive new insights from traditional sensors.
 
@@
 
-Conceptualization: M.F.C.,J.A.F.,P.J.P.,J.A.G.; Methodology: M.F.C.,J.A.F.,P.J.P.; Software: M.F.C.; Formal analysis: M.F.C.,P.J.P.; Investigation: M.F.C.,J.A.F.,P.J.P.; Resources: P.J.P.,J.A.G.; Writing - original draft: M.F.C; Writing - review & editing: M.F.C.,J.A.F.,P.J.P.,J.A.G.; Supervision: P.J.P.,J.A.G.; Project administration: P.J.P.,J.A.G.; Funding acquisition: J.A.G.
+Conceptualization: M.F.C.,J.A.F.,P.J.P.,J.A.G.; Methodology: M.F.C.,J.A.F.,P.J.P.; Software: M.F.C.; Formal analysis: M.F.C.,P.J.P.; Investigation: M.F.C.,J.A.F.,P.J.P.; Resources: P.J.P.,J.A.G.,T.L.S.; Writing - original draft: M.F.C; Writing - review & editing: M.F.C.,J.A.F.,P.J.P.,J.A.G.,T.L.S.; Supervision: P.J.P.,J.A.G.; Project administration: P.J.P.,J.A.G.; Funding acquisition: J.A.G.
 
@@
 Inset shows the IJK complex with surrounding H and L waves for the region bounded by the dashed box.
-**C**: Peaks enhanced after forward differencing acceleration (i.e., jerk).
+**C**: Peaks enhanced after forward differencing acceleration (see section **Procedure**, step 2).
 **D**: A strictly positive signal after calculating Shannon entropy.
-Y-axis values excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
-t <- theme_classic() +
+Y-axis scale units excluded because filtering introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
+t <- theme_classic(base_size = 8) +
   theme(axis.title.x = element_blank(),
@@
         strip.background = element_blank())
-t_yvoid <- t +
+t2 <- t + theme(plot.margin = margin(l = 0.5, unit = "cm"))
+t_yvoid <- t2 +
   theme(axis.text.y = element_blank(),
@@
   expand_limits(x = max(oo_400hz$dt) + 1) +
-  t
+  t2
 
@@
   expand_limits(y = 0.5) +
-  labs(y = "Acceleration\n(filtered)") +
+  labs(y = "Filtered\nacceleration") +
   t_yvoid
@@
   scale_x_datetime(date_labels = "%H:%M:%S") +
-  labs(y = "Jerk") +
+  labs(y = "Differenced\nacceleration") +
   t_yvoid
@@
 
-```{r bw-bcg, fig.cap="Example of signal processing for 3-dimensional BCG during a motionless period in a blue whale dive.
-**A**: Band-pass filtered triaxial acceleration, with cranio-caudal in orange, lateral in blue, and dorso-ventral in green.
-**B**: Peaks enhanced after forward differencing acceleration (i.e., jerk).
-**C**: The Shannon entropy combines information from all three axes and makes the signal strictly positive.
-**D**: Smoothing the Shannon entropy facilitates robust peak detection.
+```{r bw-bcg, fig.cap="A 2-hour sample of resting blue whale data.
+**A**: Depth profile consisted of ten dives to 20-30 m.
+Red dashed box indicates the dive expanded in **B, C**.
+**B, C**: Depth and rotational velocity around the lateral axis for a single dive.
+Rotational velocity was used to identify motionless periods (pink).
+Red dashed boxes indicate the motionless period in **D**-**G**.
+**D**: Band-pass filtered triaxial acceleration, with cranio-caudal in orange, lateral in blue, and dorso-ventral in green.
+**E**: Peaks enhanced after forward differencing acceleration (see section **Procedure**, step 2).
+**F**: The Shannon entropy combines information from all three axes and makes the signal strictly positive.
+**G**: Smoothing the Shannon entropy facilitates robust peak detection.
 Detected heart beats in blue.
-Y-axis values excluded because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
+Y-axis scale units excluded in **D**-**G** because the filtering process introduces magnitude distortion and only the relative shape of the signal is relevant to the analysis."}
+# Plot the BCG for a motionless period in blue whale data
 bw_bcg_plot <- function(rid) {
@@
   palette <- rgb(c(230, 86, 0), c(159, 180, 158), c(0, 233, 115), maxColorValue = 255)
+
+  t_lmargin <- t_yvoid +
+    theme(plot.margin = margin(l = 0.5, unit = "cm"))
 
@@
     facet_grid(rows = vars(axis)) +
-    labs(y = "Acceleration\n(filtered)") +
-    t_yvoid +
+    labs(y = "Filtered\nacceleration") +
+    t_lmargin +
     theme(legend.position = "none",
@@
     facet_grid(rows = vars(axis)) +
-    labs(y = "Jerk") +
-    t_yvoid +
+    labs(y = "Differenced\nacceleration") +
+    t_lmargin +
     theme(legend.position = "none",
@@
     labs(y = "Shannon\nentropy") +
-    t_yvoid
+    t_lmargin
 
@@
     labs(y = "BCG") +
-    t_yvoid
+    t_lmargin
 
@@
                      rel_heights = c(1.5, 1.5, 1, 1),
-                     labels = "AUTO")
+                     labels = c("D", "E", "F", "G"),
+                     label_size = 10)
 }
 
-bw_bcg_plot(23)
+# Plot the depth profile and motionless periods for the blue whale
+bw_10hz <- bw180905_53_10hz %>% 
+  mutate(rid_left = approx(bw_elg$start, 
+                           bw_elg$region_id, 
+                           dt, 
+                           "constant")$y,
+         rid_right = approx(bw_elg$stop, 
+                            bw_elg$region_id, 
+                            dt, 
+                            "constant", 
+                            yleft = 0)$y + 1,
+         region_id = ifelse(rid_left == rid_right, rid_left, NA),
+         dive_id = split_dives(dt, depth,
+                          surface = 2, min_depth = 10, min_dur = 5 * 60))
+divesample <- filter(bw_10hz, dive_id == 5)
+
+p_depth <- ggplot(bw_10hz, aes(dt, depth)) +
+  geom_line(size = 0.2) +
+  annotate("rect", 
+           xmin = min(divesample$dt) - 60,
+           xmax = max(divesample$dt) + 60, 
+           ymin = -1,
+           ymax = max(divesample$depth) + 1,
+           fill = NA,
+           color = "red",
+           linetype = "dotted") +
+  scale_x_datetime(date_labels = "%H:%M") +
+  scale_y_reverse("Depth (m)") +
+  theme_classic(base_size = 10) +
+  theme(axis.title.x = element_blank())
+
+# Plot the depth and y-axis gyroscope for a motionless period
+norm_y <- function(ymin, ymax, sign) {
+  (ymin + ymax) / 2 + sign * max(ymax - ymin) / 2
+}
+bw_elg <- bw180905_53_elg %>% 
+  mutate(
+    depth_min = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% min()
+    }),
+    depth_max = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% max()
+    }),
+    depth_med = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), "depth"] %>% 
+        first() %>% 
+        median()
+    }),
+    ygyr_min = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% min()
+    }),
+    ygyr_max = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% max()
+    }),
+    ygyr_med = map2_dbl(start, stop, function(t1, t2) {
+      bw180905_53_10hz[between(bw180905_53_10hz$dt, t1, t2), ]$gw[,2] %>% 
+        first() %>% 
+        median()
+    }),
+  )
+diveelg <- bw_elg %>% 
+  filter(start >= first(divesample$dt), stop <= last(divesample$dt)) %>% 
+  mutate(depth_min2 = norm_y(depth_min, depth_max, -1),
+         depth_max2 = norm_y(depth_min, depth_max, 1),
+         ygyr_min2 = norm_y(ygyr_min, ygyr_max, -1),
+         ygyr_max2 = norm_y(ygyr_min, ygyr_max, 1))
+
+depth_profile <- ggplot(divesample, aes(dt, depth)) +
+  geom_line(size = 0.5) +
+  geom_rect(aes(xmin = start, 
+                xmax = stop, 
+                ymin = depth_min2, 
+                ymax = depth_max2),
+            diveelg,
+            inherit.aes = FALSE,
+            fill = "orchid1",
+            color = NA,
+            alpha = 0.5) +
+  geom_rect(aes(xmin = start, 
+                xmax = stop, 
+                ymin = depth_min2, 
+                ymax = depth_max2),
+            filter(diveelg, region_id == 23),
+            inherit.aes = FALSE,
+            fill = NA,
+            color = "red",
+            linetype = "dotted",
+            alpha = 0.5) +
+  scale_x_datetime(date_labels = "%H:%M") +
+  scale_y_reverse("Depth (m)") +
+  theme_classic(base_size = 10) +
+  theme(axis.title.x = element_blank(),
+        plot.margin = margin(l = 0.5, unit = "cm"))
+ygyr_profile <- ggplot(divesample, aes(dt, gw[, 2])) +
+  geom_line(size = 0.2) +
+  geom_rect(aes(xmin = start, 
+                xmax = stop, 
+                ymin = ygyr_min2, 
+                ymax = ygyr_max2),
+            diveelg,
+            inherit.aes = FALSE,
+            fill = "orchid1",
+            color = NA,
+            alpha = 0.5) +
+  geom_rect(aes(xmin = start, 
+                xmax = stop, 
+                ymin = ygyr_min2, 
+                ymax = ygyr_max2),
+            filter(diveelg, region_id == 23),
+            inherit.aes = FALSE,
+            fill = NA,
+            color = "red",
+            linetype = "dotted",
+            alpha = 0.5) +
+  scale_x_datetime(date_labels = "%H:%M") +
+  scale_y_continuous(bquote(omega~(rad~s^{-1}))) +
+  theme_classic(base_size = 10) +
+  theme(axis.title.x = element_blank(),
+        plot.margin = margin(l = 0.5, unit = "cm"))
+p_motionless <- cowplot::plot_grid(depth_profile, ygyr_profile, 
+                                   ncol = 1, 
+                                   align = "v", 
+                                   labels = c("B", "C"),
+                                   label_size = 10)
+
+cowplot::plot_grid(
+  cowplot::plot_grid(p_depth, p_motionless, 
+                     nrow = 1,
+                     labels = c("A", ""),
+                     label_size = 10),
+  bw_bcg_plot(23),
+  ncol = 1,
+  rel_heights = c(2, 3)
+)
 ```
 
-```{r bw-validation-plots, fig.cap="**A** Signal-to-noise ratio was higher for the 3-dimensional BCG (lower panel) than the 1-dimensional BCG (cranio-caudal acceleration only; upper panel).
+```{r validation-plots, fig.cap="**A** The blue whale signal-to-noise ratio was higher for the 3-dimensional BCG (lower panel) than the 1-dimensional BCG (cranio-caudal acceleration only; upper panel).
 Each panel shows the power spectral density for the BCG.
@@
 The signal-to-noise ratio was calculated as the ratio of the area under the curve in the signal band to the area under the rest of the curve, up to 60 bpm.
-**B** Heart rates observed in the 3-dimensional BCG followed characteristic diving physiology patterns.
+**B** Heart rates observed in the 3-dimensional blue whale BCG followed characteristic diving physiology patterns.
 Heart rate is lowest at the start of the dive (~4-5 bpm), increasing towards ascent (~8-9 bpm).
 Points indicate instantaneous heart rates and the line is a Theil-Sen regression.
-Outliers likely represent premature beats which are common in heart rate profiles during dives of cetaceans, pinnipeds, and penguins [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014; @andrews1997; @wright2014]."}
+Outliers likely represent premature beats which are common in heart rate profiles during dives of cetaceans, pinnipeds, and penguins [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014; @andrews1997; @wright2014].
+**C**: BCG- and ECG-derived instantaneous heartrates were equivalent in the killer whale.
+The slope and intercept of the ordinary least squares regression of BCG- on ECG-derived instantaneous heartrates (solid blue line) were not significantly different from 1 and 0, respectively (dashed black line)."}
 labels <- bw_psd %>% 
@@
   mutate(s2n_text = sprintf("Signal:noise = %0.2f", s2n))
+bcg_labels <- bw_psd %>% 
+  filter(between(freq_bpm, 7.99, 8.01)) %>% 
+  group_by(metric) %>% 
+  slice(1) %>% 
+  ungroup() %>% 
+  mutate(bcg_lbl = c("1D BCG", "3D BCG"),
+         spec = c(1.75e-7, 3.2e-9))
 pA <- ggplot(bw_psd, aes(x = freq_bpm, y = spec)) +
@@
   geom_text(aes(label = s2n_text), labels, hjust = -0.1, vjust = -0.05) +
+  geom_text(aes(label = bcg_lbl), bcg_labels, hjust = 0, vjust = 1) +
   labs(x = "Frequency (bpm)",
-       y = "Power spectrum density") +
+       y = "Power spectral density") +
   facet_grid(rows = vars(metric),
@@
 pB <- ggplot(bpm_data, aes(dive_norm, bcg_bpm)) +
-  geom_point() +
-  geom_line(data = bpm_pred, color = "blue", size = 1.5) +
+  geom_point(size = 0.5) +
+  geom_line(data = bpm_pred, color = "blue", size = 1) +
   scale_x_continuous() +
@@
 
-cowplot::plot_grid(pA, pB, nrow = 1, labels = "AUTO")
+pC <- ggplot(ecg_bpm, aes(ecg_bpm, bcg_bpm)) +
+  geom_abline(slope = 1, intercept = 0, size = 1, linetype = "dashed") +
+  geom_smooth(method = "lm", se = FALSE, size = 1) +
+  geom_point(size = 0.5) +
+  scale_x_continuous("ECG (bpm)", breaks = seq(55, 70, by = 5)) +
+  scale_y_continuous("BCG (bpm)", breaks = seq(55, 70, by = 5)) +
+  coord_fixed() +
+  theme_classic() +
+  theme(axis.title.x = element_text())
+
+cowplot::plot_grid(
+  pA, 
+  cowplot::plot_grid(pB, pC, ncol = 1, labels = c("B", "C")),
+  nrow = 1, 
+  labels = c("A", "")
+)
 ```

### 2022-02-02 analysis/supplementary-materials/supplemental.Rmd
Address reviewer feedback

--- 2021-10-29 analysis/supplementary-materials/supplemental.Rmd
@@
 
-```{r peak-heur, fig.cap="**A:** Minor peaks (hollow points) in the ballistocardiogram (BCG, line) were not considered heart beats.
-Only major peaks (solid points) were retained for analysis.
+<!--
+ABD: Use color instead of open/close
+C: Axis tick label clipping
+D: open and close doesn't work because of overplotting
+-->
+```{r peak-heur, fig.cap="Heuristic method for discriminating between major and minor peaks.
+**A:** Minor peaks (red points) in the ballistocardiogram (BCG, line) were not considered heart beats.
+Only major peaks (blue points) were retained for analysis.
 The BCG for one motionless period shown here.
@@
 **C:** The distance to the highest peak exhibited a bimodal distribution.
-We chose a distance threshold (dashed red line) corresponding to the valley in the density curve.
+We chose a distance threshold (dashed purple line) corresponding to the valley in the density curve.
 **D:** All peaks found in the BCG across all motionless periods in height-prominence space.
 Solid and hollow points as in **A**.
-The dashed red curve corresponds to the distance threshold in **C**."}
+The dashed purple curve corresponds to the distance threshold in **C**."}
 fs_hz <- 400
@@
   geom_line() +
-  geom_point(aes(y = height, shape = is_beat), data = peaksample) +
+  geom_point(aes(y = height, color = is_beat), data = peaksample) +
   geom_text(aes(y = height, label = peak_lbl), 
@@
             vjust = -0.3) +
-  scale_shape_manual(values = c(21, 16)) +
-  labs(y = "Y") +
+  scale_color_manual(values = c("red", "blue")) +
+  labs(x = "", y = "Y") +
   theme_classic(base_size = 10) +
@@
 
-pB <- ggplot(peaksample, aes(height, prominence, shape = is_beat)) +
+pB <- ggplot(peaksample, aes(height, prominence, color = is_beat)) +
   geom_point() +
   ggrepel::geom_text_repel(aes(label = peak_lbl), force = 3) +
-  scale_shape_manual(values = c(21, 16)) +
+  scale_color_manual(values = c("red", "blue")) +
   scale_x_continuous(n.breaks = 4) +
@@
   geom_density() +
-  geom_vline(xintercept = dist_thr, linetype = 2, color = "red") +
+  geom_vline(xintercept = dist_thr, linetype = 2, color = "purple") +
   scale_x_continuous(n.breaks = 4) +
@@
 pD <- ggplot(peak_tbl, aes(height, prominence)) +
-  geom_point(aes(shape = is_beat)) +
-  geom_line(data = thr_curve, linetype = 2, color = "red") +
-  scale_shape_manual(values = c(21, 16)) +
+  geom_point(aes(color = is_beat), shape = 21) +
+  geom_line(data = thr_curve, linetype = 2, color = "purple") +
+  scale_color_manual(values = c("red", "blue")) +
   labs(x = "Peak height",
@@
 ```
-
-```{r ygyr, fig.cap="Motionless periods were manually identified by looking for periods of low amplitude rotational veloctiy ($\\omega$) around the lateral axis, indicative of cetaceans' dorso-ventral fluking motion.
-The depth and $\\omega$ profiles for the first dive shown here, with motionless periods in pink."}
-divesample <- filter(bw_10hz, dive_id == 1)
-norm_y <- function(ymin, ymax, sign) {
-  (ymin + ymax) / 2 + sign * max(ymax - ymin) / 2
-}
-diveelg <- bw_elg %>% 
-  filter(start >= first(divesample$dt), stop <= last(divesample$dt)) %>% 
-  mutate(depth_min2 = norm_y(depth_min, depth_max, -1),
-         depth_max2 = norm_y(depth_min, depth_max, 1),
-         ygyr_min2 = norm_y(ygyr_min, ygyr_max, -1),
-         ygyr_max2 = norm_y(ygyr_min, ygyr_max, 1))
-
-depth_profile <- ggplot(divesample, aes(dt, depth)) +
-  geom_line() +
-  geom_rect(aes(xmin = start, 
-                xmax = stop, 
-                ymin = depth_min2, 
-                ymax = depth_max2),
-            diveelg,
-            inherit.aes = FALSE,
-            fill = "orchid1",
-            color = NA,
-            alpha = 0.5) +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  scale_y_reverse("Depth (m)") +
-  theme_classic(base_size = 10) +
-  theme(axis.title.x = element_blank())
-ygyr_profile <- ggplot(divesample, aes(dt, gw[, 2])) +
-  geom_line() +
-  geom_rect(aes(xmin = start, xmax = stop, ymin = ygyr_min2, ymax = ygyr_max2),
-            diveelg,
-            inherit.aes = FALSE,
-            fill = "orchid1",
-            color = NA,
-            alpha = 0.5) +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  scale_y_continuous(bquote(omega~(rad~s^{-1}))) +
-  theme_classic(base_size = 10) +
-  theme(axis.title.x = element_blank())
-cowplot::plot_grid(depth_profile, ygyr_profile, 
-                   ncol = 1, 
-                   align = "v", 
-                   labels = "AUTO")
-```
-
-```{r bw-profile, fig.cap="Two hours of blue whale tag data were used for ballistocardiographic analysis.
-During these hours, the animal made repeated, shallow dives.
-Because the BCG can be obscured by dynamic body movements, only relatively motionless periods were retained for analysis (pink).
-These periods were identified by visual examination of the rotational velocity around the lateral axis."}
-ggplot(bw_10hz, aes(dt, depth)) +
-  geom_line(size = 0.5) +
-  geom_line(aes(group = region_id), 
-            data = drop_na(bw_10hz, region_id), 
-            size = 0.5, 
-            color = "orchid1") +
-  scale_x_datetime(date_labels = "%H:%M:%S") +
-  scale_y_reverse("Depth (m)") +
-  theme_classic(base_size = 10) +
-  theme(axis.title.x = element_blank())
-```

### 2022-02-09 analysis/supplementary-materials/supplemental.Rmd
Fix Fig S1. Requires manual editing now, boo.

@@
 
-regionsample <- filter(bw_400hz, region_id == 6)
+regionsample <- filter(bw_400hz, region_id == 6) %>% 
+  mutate(secs = as.numeric(dt - min(dt), unit = "secs"))
 peaksample <- filter(peak_tbl, between(dt, 
@@
   arrange(dt) %>% 
-  mutate(peak_lbl = seq(n()))
+  mutate(peak_lbl = seq(n()),
+         secs = as.numeric(dt - min(regionsample$dt)))
 
@@
 
-pA <- ggplot(regionsample, aes(dt, jerk_smooth)) +
+pA <- ggplot(regionsample, aes(secs, jerk_smooth)) +
   geom_line() +
@@
   scale_color_manual(values = c("red", "blue")) +
-  labs(x = "", y = "Y") +
+  labs(x = "Seconds", y = "Y") +
   theme_classic(base_size = 10) +
-  theme(axis.title.x = element_blank(),
-        legend.position = "FALSE")
+  theme(legend.position = "FALSE")
 
@@
 
-cowplot::plot_grid(pA, pB, pC, pD, ncol = 2, labels = "AUTO")
+# Figure S1 requires manual adjustments. Save to .svg, edit in Inkscape, then
+# export back to PNG and insert here.
+S1 <- cowplot::plot_grid(pA, pB, pC, pD, ncol = 2, labels = "AUTO")
+ggsave(here::here("analysis/figures/S1.svg"), S1)
+cowplot::ggdraw() +
+  cowplot::draw_image(here::here("analysis/figures/S1_edited.svg"))
 ```

### 2022-02-09 analysis/paper/paper.Rmd
Almost done with edits for reviewers!

@@
 
-Here we present a method for generating a BCG from bio-logger accelerometry.
+Here we present a method for generating a BCG from bio-logger cranio-caudal acceleration.
 We validated our method with a simultaneously recorded ECG on an adult killer whale in managed care (*Orcinus orca*) and applied it to detect heartrate in a blue whale.
-The relative orientation of the tag on the body is often uncertain when bio-loggers are deployed in the wild [@johnsonDigitalAcousticRecording2003], so in addition to a one-dimensional BCG based solely on cranio-caudal acceleration, we also generated a three-dimensional BCG, which we expected would be more robust in a field setting.
+The relative orientation of a tag on a cetacean's body is often uncertain when bio-loggers are deployed in the wild [@johnsonDigitalAcousticRecording2003], so isolating acceleration along the cranio-caudal axis is subject to error.
+Therefore, we also compared a tri-axial BCG to the cranio-caudal BCG.
 Specifically, we tested three hypotheses to validate our method.
-First, a one-dimensional BCG would, in a controlled setting, produce instantaneous heartrates that are statistically equivalent to ECG instantaneous heartrates.
-Second, a three-dimensional BCG would, in a field setting, produce a more robust signal than a one-dimensional BCG.
+First, a cranio-caudal (1D) BCG would, in a controlled setting, produce instantaneous heartrates that are statistically equivalent to ECG instantaneous heartrates.
+Second, a tri-axial (3D) BCG would, in a field setting, produce a more robust signal than a 1D BCG.
 Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
@@
 For example, BCGs extracted from gliding phases before and after controlled sonar exposure experiments could quantify the physiological response to anthropogenic disturbance [@southall2019].
-Even as the field of physio-logging progresses with new hardware innovations, this method demonstrates that computational advances can derive new insights from traditional sensors.
+Even as the field of physio-logging progresses with new hardware innovations [@williams2021; @fahlman2021; @hawkesIntroductionThemeIssue2021], this method demonstrates that computational advances can derive new insights from traditional sensors.
 

### 2022-02-15 analysis/paper/paper.Rmd
Fix references (add pkg versions) and finish revisions

@@
 bibliography: [zotero.bib,references.bib]
-csl: "../templates/the-journal-of-experimental-biology.csl"
+csl: "../templates/the-company-of-biologists.csl"
 abstract: |
@@
 We attached the CATS tag on the mid-lateral left chest posterior to the pectoral fin (Movie S1).
-The CATS tag recorded acceleration at 400 Hz, magnetometer and gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps (*TODO: get tag 97 accelerometer accuracy/precision from Will G)*.
+The CATS tag recorded tri-axial acceleration at 400 Hz, tri-axial magnetometer and tri-axial gyroscope at 50 Hz, pressure at 10 Hz, and video at 30 fps.
+The IMU in the CATS tag was a MPU-9250 (InvenSense, San Jose, CA; www.invensense.com).
+The accelerometer had dynamic range of ±4 g, sensitivity of 8,192 LSB g^-1^, and accuracy of 6.1 × 10^-6^ g.
 All sensors were rotated from the tag's frame of reference to that of the whale using MATLAB (MathWorks, Inc., v2020b) tools for processing CATS data [@cadeToolsIntegratingInertial2021].
@@
 We deployed the tag using a 4 m fiberglass pole from a 6.3 m rigid-hulled inflatable boat [as described by @goldbogen2006] The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
-Tag configuration and data processing followed the same procedure as the killer whale, including sampling rates for inertial sensors and video.
+Tag configuration and data processing followed the same procedure as the killer whale, including accelerometer specification and sampling rates for inertial sensors and video.
 The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**).
@@
 This is a SeaWorld Parks and Entertainment Technical Contribution number 2021-12.
+We thank Lucy Hawkes and two anonymous reviewers for comments on the manuscript.
 
@@
   geom_smooth(method = "lm", se = FALSE, size = 1) +
-  geom_point(size = 0.5) +
+  geom_point(shape = 21) +
   scale_x_continuous("ECG (bpm)", breaks = seq(55, 70, by = 5)) +

### 2022-04-27 analysis/paper/paper.Rmd
Update Zenodo citation
Second round of revisions

@@
 Second, a tri-axial (3D) BCG would, in a field setting, produce a more robust signal than a 1D BCG.
-Third, BCG-derived heartrates would increase during the latter phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
+Third, BCG-derived heartrates would increase during the later phases of dives, consistent with the progressive increase in heartrate routinely observed prior to and during ascent [@goldbogenExtremeBradycardiaTachycardia2019; @mcdonaldDeepdivingSeaLions2014].
 
@@
 A 24.5 m blue whale was tagged with an archival, suction-cup CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 [previously published by @goughScalingSwimmingPerformance2019].
-We deployed the tag using a 4 m fiberglass pole from a 6.3 m rigid-hulled inflatable boat [as described by @goldbogen2006] The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
+We deployed the tag using a 4 m fiberglass pole from a 6.3 m rigid-hulled inflatable boat and recovered it via radio VHF tracking [as described by @goldbogen2006].
+The tag slid behind the left pectoral flipper, similar to the placement of the CATS tag on the killer whale.
 Tag configuration and data processing followed the same procedure as the killer whale, including accelerometer specification and sampling rates for inertial sensors and video.
@@
 Differentiation (i.e., $a_{t+1}-a_t$, where $a_t$ is the observed acceleration at time step $t$) exaggerates peaks, like the J wave, but it is sensitive to noisy signals.
-Therefore, the signal should be de-noised prior to differentiation.
-A moving average smoother can remove noise, but it would also reduce the amplitude of peaks.
+Therefore, additional noise reduction is necessary prior to differentiation.
+A moving average smoother could remove noise, but it would also reduce the amplitude of peaks.
 Hence, differentiating Savitzky-Golay filters are preferred in peak-detection algorithms because they remove noise while retaining the general shape of peaks [@samann2019].
@@
 
-We generated 1-dimensional and 3-dimensional BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig.
+We generated 1-dimensional and 3-dimensional BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (`r sprintf("%0.1f%%", motionless_minutes / 120 * 100)` of the 2-hour record) (Fig.
 \@ref(fig:bw-bcg)**A-C**).
@@
 The 3-dimensional BCG (Fig.
-\@ref(fig:bw-bcg)) produced a more robust signal (i.e., higher signal-to-noise ratio) than the 1-dimensional BCG, which used only cranio-caudal acceleration.
-The signal-to-noise ratio was 2.00 for the 3-dimensional BCG, compared to 0.17 for the 1-dimensional BCG (Fig.
+\@ref(fig:bw-bcg)) produced a more robust signal (i.e., higher signal-to-noise ratio) than the 1-dimensional BCG, which used only cranio-caudal acceleration (Fig.
 \@ref(fig:validation-plots)**A**).
+The signal-to-noise ratio was 2.00 for the 3-dimensional BCG, compared to 0.17 for the 1-dimensional BCG.
 Although the power spectral density curve for the 1-dimensional BCG had a peak in the 4-8 bpm frequency range, most of the signal's power was concentrated in lower frequencies.
@@
 
-3-dimensional BCG-derived heart rates exhibited a relaxation of bradycardia over the course of dives.
+The 3-dimensional BCG exhibited increasing heart rates over the course of dives.
 Average heart rate increased from `r format(bpm_pred$bcg_bpm[1], digits = 2)` bpm at the start of dives to `r format(bpm_pred$bcg_bpm[nrow(bpm_pred)], digits = 2)` bpm at the end of dives (Theil-Sen regression, $p < 10^{-10}$) (Fig.
@@
 This limits the behavioral and physiological contexts in which heartrate may be measured.
-For example, the BCG is unlikely to quantify neither the magnitude of surface tachycardia [@goldbogenExtremeBradycardiaTachycardia2019] nor exercise modulation of bradycardia [@noren2012], due to movement artifacts during those activities.
+For example, the BCG is probably an inappropriate method for quantifying the magnitude of surface tachycardia [@goldbogenExtremeBradycardiaTachycardia2019] and exercise modulation of bradycardia [@noren2012], due to movement artifacts during those activities.
 Secondarily, we did not test whether the BCG is robust to tag placement location.
@@
 
-All data and code used in this analysis are available on Zenodo (DOI needed).
+All data and code used in this analysis are available on Zenodo [@cetaceanbcgzenodo].
 

### 2022-04-28 analysis/paper/paper.Rmd
Knit figures to PDF

@@
   fig.width = 6,
+  dev = "pdf",
   dpi = 300
@@
 While the BCG method presented here holds the potential to mine existing and future marine mammal bio-logging datasets for information about cardiovascular function, it has several limitations compared to ECG methods.
-Most importantly, BCGs are highly sensitive to movement artifacts [@inanBallistocardiographySeismocardiographyReview2015], so only motionless periods are valid for analysis.
+Most importantly, BCGs are highly sensitive to movement artifacts [@inanBallistocardiographySeismocardiographyRemovview2015], so only motionless periods are valid for analysis.
 This limits the behavioral and physiological contexts in which heartrate may be measured.
